---
title: "Design-Weighted LCA: Data and Model Selection"
subtitle: "Institutional trust in Mexico: LAPOP AmericasBarometer 2023"
author: "Kevin"
date: today
format:
  pdf:
    toc: false
    number-sections: false
    fig-width: 7
    fig-height: 4.5
    fig-align: center
    df-print: kable
    geometry: margin=1in
    fig-crop: false   # MiKTeX's pdfcrop is fragile; cropping is cosmetic only
execute:
  warning: false
  message: false
---

# Purpose

This script is the heavy half of the pipeline: it presents the raw data and the dictionary, then searches over K with the design-weighted pseudo-likelihood, and, once the analyst hard-codes `cfg$K_force`, produces the model-fit diagnostics (item discrimination, bivariate residuals) and saves every object the segments script needs to `{cfg$out_dir}`. Nothing here is interpretive: interpretation, labeling, prediction, and the handoff live in `survey_lca_segments.qmd`, which renders in seconds because this script did the fitting.

```{r setup}
#| include: false
library(matrixStats); library(clue); library(survey); library(furrr)
library(poLCA); library(arrow); library(tidyverse)   # tidyverse LAST (masking)
source("survey_lca_source.R")
source("survey_data_config.R")
init_parallel(cfg)
K_ready <- !is.null(cfg$K_force)
```

# The data

The stacked bars show the items as read, with missingness explicit: the grey mass is exactly what the complete-case rule removes. The dictionary pairs each analysis name with its source code, the question text, and the response labels as text, exactly what the segment labeler will later read; it is split across pages so it never overflows.

```{r data-missingness}
#| fig-height: 5
#| fig-cap: "All respondents, all items, missingness shown explicitly (grey). The complete-case analysis sample removes rows contributing to the grey mass."
print(plot_item_stack(survey_dat_original, items,
                      "Raw items with missingness", show_missing = TRUE))
cat(nrow(survey_dat), "of", nrow(survey_dat_original),
    "cases are complete on items and demographics and enter the analysis.\n")
```

```{r data-dictionary}
#| results: asis
# Two tables so the dictionary fits PDF pages regardless of item count; the
# split is computed from length(items), nothing hard-coded. format = "pipe"
# lets Pandoc build a wrapping table (LaTeX tabular does not wrap long
# question text), and cat() guarantees emission under results: asis.
halves <- split(seq_along(items), ceiling(2 * seq_along(items) / length(items)))
walk(seq_along(halves), function(b) {
  cat(knitr::kable(dictionary[halves[[b]], ], format = "pipe",
                   col.names = c(
                     "Analysis name", "Source variable", "Survey question",
                                 "Response labels"),
                   caption = paste0("Item dictionary (", b, " of ", 
                                    length(halves),
                                    "): the renamed analysis variable, its source variable ", "in the .sav, the question wording, and the response ", "labels the model and the segment labeler read.")),
      sep = "\n")
  cat("\n\n")
  if (b < length(halves)) cat("\\newpage\n\n")
})
```

# Preparing the items

`prepare_items()` (source file) verifies the configured columns, coerces design variables to base types, recodes each item to consecutive integers over its substantive categories, and fixes the sum-to-n scale for the information criteria (rationale documented at its definition).

```{r prepare}
prep_fit     <- prepare_items(cfg)
dat_prepared <- prep_fit$dat_prepared
cats         <- prep_fit$cats
w_vec        <- prep_fit$w_vec
scale_ic     <- prep_fit$scale_ic
knitr::kable(prep_fit$summary, align = "lrr",
             caption = "Per-item category counts and missing footprint.")
```

# Searching over K

Every candidate K is fit with `cfg$n_starts` seeded random starts; the table reports the weighted BIC and AIC on the sum-to-n scale, entropy, and the per-K inclusion probabilities. **Nothing is auto-selected**: read this evidence, set `cfg$K_force` in the config script, and re-render for that model's diagnostics.

```{r enumerate}
# df_k() and entropy_R2() are defined in survey_lca_source.R; w_vec and the
# sum-to-n scale_ic came from prepare_items() (rationale documented there).
enum <- furrr::future_map_dfr(cfg$K_range, function(K) {
  fit  <- fit_lca(dat_prepared, w_vec, cats, cfg$items, K, cfg$n_starts)
  ll_n <- fit$ll * scale_ic                       # log-likelihood on the sum-to-n scale
  tibble(K = K,
         logLik = ll_n,
         df = df_k(K, cats),
         AIC = -2 * ll_n + 2 * df_k(K, cats),
         BIC = -2 * ll_n + df_k(K, cats) * log(nrow(dat_prepared)),
         entropy = entropy_R2(fit$post, K),
         converged = fit$converged)
}, .options = furrr::furrr_options(seed = TRUE,
     packages = c("matrixStats", "clue", "purrr", "dplyr", "tidyr")))

# K is the ANALYST's decision (cfg$K_force), made by iterating over the
# enumeration and the per-K diagnostics; nothing is auto-selected.
K_star <- as.integer(cfg$K_force %||% NA)

enum |>
  mutate(across(c(logLik, AIC, BIC), ~ round(.x, 1)),
         entropy = round(entropy, 3)) |>
  knitr::kable(align = "r",
               caption = "Weighted fit statistics by number of segments. Nothing is selected here: read BIC, AIC, and entropy together and set cfg$K_force.")
```

```{r fig-enum}
#| fig-cap: "Enumeration diagnostics across K. Look for the BIC elbow and read entropy alongside it; neither decides alone."
enum |>
  dplyr::select(K, BIC, entropy) |>
  pivot_longer(-K, names_to = "criterion", values_to = "value") |>
  mutate(criterion = recode(criterion, BIC = "Weighted BIC", entropy = "Relative entropy")) |>
  ggplot(aes(K, value)) +
  geom_line(linewidth = 0.6, color = viridisLite::viridis(1, begin = 0.4)) +
  geom_point(size = 2.4, color = viridisLite::viridis(1, begin = 0.4)) +
  facet_wrap(~ criterion, scales = "free_y") +
  scale_x_continuous(breaks = cfg$K_range) +
  labs(x = "Number of latent classes (K)", y = NULL) +
  theme_lca()
```

```{r k-gate}
#| results: asis
if (!K_ready) cat("**`cfg$K_force` is `NULL`.** Review the enumeration evidence above, set a candidate K in `survey_data_config.R`, and re-render for that model's discrimination and bivariate-residual diagnostics and the saved objects.\n")
```

# The chosen model and its diagnostics

```{r fit-final}
#| eval: !expr K_ready
K_star <- as.integer(cfg$K_force)
# Fit at K_star on the complete cases, weighted and unweighted. The weighted fit sets
# the label order; the unweighted fit is aligned to it so "Class k" is consistent.
fit_w <- fit_lca(dat_prepared, w_vec, cats, cfg$items, K_star, cfg$n_starts)
fit_u <- align_to(fit_lca(dat_prepared, rep(1, nrow(dat_prepared)), cats, cfg$items,
                          K_star, cfg$n_starts), fit_w)

# Posteriors recomputed from the FINAL parameters, then written back so the same replicate
# design serves both the measurement-model CIs and the profiling regression.
inp_full  <- make_inputs(dat_prepared, cfg$items, cats)
post_w    <- posterior_of(fit_w$pi, fit_w$rho, inp_full$Y)
colnames(post_w) <- paste0("post_class", seq_len(K_star))
modal_int <- max.col(post_w, ties.method = "first")
dat_prepared <- dat_prepared |>
  bind_cols(as_tibble(post_w)) |>
  mutate(modal_class   = factor(modal_int, levels = seq_len(K_star)),
         max_posterior = post_w[cbind(seq_len(n()), modal_int)])

# Report the fitted model: convergence, fit on the sum-to-n scale, classification
# sharpness, and the weighted prevalences that the segments script will re-present
# with design-based intervals.
cat(sprintf(paste0("Fitted K = %d on %d complete cases (%d random starts).\n",
                   "Converged: %s | weighted log-likelihood (sum-to-n scale): %.1f\n",
                   "Entropy R2: %.3f | mean maximum posterior: %.3f\n"),
            K_star, nrow(dat_prepared), cfg$n_starts,
            ifelse(isTRUE(fit_w$converged), "yes", "NO - inspect before trusting"),
            fit_w$ll * scale_ic, entropy_R2(post_w, K_star),
            mean(dat_prepared$max_posterior)))

knitr::kable(
  tibble(Segment = paste("Segment", seq_len(K_star)),
         `Weighted prevalence` = round(fit_w$pi, 3),
         `Unweighted prevalence` = round(fit_u$pi, 3),
         `Modal-assignment share` = round(as.numeric(prop.table(
           table(factor(modal_int, levels = seq_len(K_star))))), 3)),
  align = "lrrr",
  caption = "Segment prevalences at the chosen K. The weighted column is the population estimate (design-based intervals appear in the segments script); the modal-assignment share is the sample share of respondents whose highest posterior is that segment, and differs from the model prevalence whenever classification is uncertain.")
```

## Which items carry the segmentation

```{r item-discrimination}
#| eval: !expr K_ready
# Discrimination of each item under the fitted weighted model: how differently the classes
# answer it, measured as the mean over class pairs of the total-variation (TV) distance
# between their class-conditional response distributions. 0 = all classes answer alike (a
# drop candidate); near 1 = strong separation. Loop-free: map over items, then over pairs.
# Computed here; shown in the figure below and in the screen table further down.
class_pairs <- utils::combn(K_star, 2, simplify = FALSE)   # list of c(a, b) class-index pairs

item_discrim <- tibble(
  item = cfg$items,
  discrimination = map_dbl(fit_w$rho, function(rho_j) {
    # rho_j is (categories x classes); rho_j[, k] is class k's response distribution.
    # TV distance for one pair of classes, then averaged over all pairs.
    mean(map_dbl(class_pairs, ~ 0.5 * sum(abs(rho_j[, .x[1]] - rho_j[, .x[2]]))))
  })
) |>
  arrange(desc(discrimination))

knitr::kable(item_discrim |> mutate(discrimination = round(discrimination, 3)),
             col.names = c("Item", "Discrimination"), align = "lr",
             caption = "Item discrimination: the mean over segment pairs of the total-variation distance between their response distributions. 0 means every segment answers the item alike (a drop candidate); values near 1 mean the item separates segments sharply.")
```

```{r fig-discrimination}
#| eval: !expr K_ready
#| fig-cap: "Item discrimination: how much each item separates the segments. Low bars identify candidates for removal in a sensitivity refit."
item_discrim |>
  mutate(item = factor(item, levels = rev(item))) |>
  ggplot(aes(discrimination, item)) +
  geom_col(fill = viridisLite::viridis(1, begin = 0.45), width = 0.7) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "grey55") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "Discrimination (mean pairwise total-variation distance)", y = NULL) +
  theme_lca()
```

```{r item-screen}
#| eval: !expr K_ready
# Automated screen: does each item earn its class-DEPENDENT parameters by BIC? Compare, on
# the SAME respondents, the fitted model against a restricted model in which item j is made
# class-INDEPENDENT (its response distribution shared across classes, so it tells us nothing
# about the latent classes). Under local independence that restricted model factorises into
# an LCA on the OTHER items times item j's marginal, so its weighted pseudo-log-likelihood
# is ll(LCA without item j) + ll(item j's weighted marginal), and its BIC is directly
# comparable to the full model's. delta_bic = BIC(restricted) - BIC(full): NEGATIVE means
# BIC prefers dropping item j's class-dependence, so the item is flagged.
#
# This is the slowest chunk: it refits the LCA once per item. Lower cfg$n_starts here if you
# need speed, at some risk of a local optimum. It is reproducible because fit_lca() seeds
# each start from cfg$seed.
n_eff   <- nrow(dat_prepared)
# Put every log-likelihood on the same sum-to-n scale used for the criteria in
# Section 8 (scale_ic = n / sum(w_vec)), so the screen's BIC comparison is
# calibrated the same way as the K-selection BIC.
ll_full <- fit_w$ll * scale_ic
df_full <- df_k(K_star, cats)                    # df_k() was defined in Section 8

# Weighted log-likelihood of one item under a single (class-independent) distribution.
marginal_ll <- function(item) {
  y <- as.integer(dat_prepared[[item]])
  p <- as.numeric(xtabs(w_vec ~ y)); p <- p / sum(p)   # weighted response proportions, 1..C
  sum(w_vec * log(p[y]))
}

bic_screen <- tibble(item = cfg$items) |>
  mutate(delta_bic = furrr::future_map_dbl(cfg$items, function(j) {
    keep     <- setdiff(cfg$items, j)
    fit_wo   <- fit_lca(dat_prepared, w_vec, cats[keep], keep, K_star, cfg$n_starts)
    ll_restr <- (fit_wo$ll + marginal_ll(j)) * scale_ic            # sum-to-n scale, as above
    df_restr <- df_k(K_star, cats[keep]) + (cats[[j]] - 1)         # item j: one shared distribution
    (-2 * ll_restr + df_restr * log(n_eff)) - (-2 * ll_full + df_full * log(n_eff))
  }, .options = furrr::furrr_options(seed = TRUE,
       packages = c("matrixStats", "clue", "purrr", "dplyr", "tidyr"))))

# Combine the effect size and the decision into one ranked, flagged table.
item_screen <- item_discrim |>
  left_join(bic_screen, by = "item") |>
  mutate(flag = if_else(delta_bic < 0, "drop candidate", "keep")) |>
  arrange(delta_bic)

item_screen |>
  mutate(discrimination = round(discrimination, 3), delta_bic = round(delta_bic, 1)) |>
  knitr::kable(align = "lrrl",
               col.names = c("Item", "Discrimination", "Delta BIC", "Screen"),
               caption = "Automated indicator screen, most-droppable first. Discrimination is the mean pairwise total-variation distance (effect size). Delta BIC is BIC(item made class-independent) minus BIC(full model); a negative value means BIC prefers dropping the item's class-dependence, so it is flagged. The more negative the gap, the stronger the case (beyond about 6 is strong, beyond 10 very strong).")
```

## Local independence: bivariate residuals

```{r bvr}
#| eval: !expr K_ready
# bvr_pairs() is defined in survey_lca_source.R; fit$rho is positional in cfg$items
# order, so the items are passed in exactly that order.
bvr_tbl <- bvr_pairs(dat_prepared, w_vec, cfg$items, fit_w)

bvr_tbl |>
  mutate(bvr = round(bvr, 2)) |>
  slice_head(n = 10) |>
  knitr::kable(align = "llrr",
               col.names = c("Item A", "Item B", "Pair df", "BVR"),
               caption = "Largest bivariate residuals (BVR), a per-pair local-independence index: the design-weighted observed two-way table against the model-implied table, as Pearson X2 / (n-scaled) per degree of freedom. Descriptive ranking only; no chi-square test applies under the weighted pseudo-likelihood. Pairs standing well above the rest of the table mark associations the classes do not explain; absolute SRS cutoffs do not transfer under unequal weights.")
```

```{r bvr-heatmap}
#| eval: !expr K_ready
#| fig-height: 6
#| fig-cap: "Bivariate residuals as a heatmap: darker cells are item pairs whose association the segments explain least well (a descriptive ranking, not a test). Persistent dark pairs argue for dropping/merging an item or trying K+1."
# Column names come from bvr_pairs() (item_a, item_b, df, bvr); axis order is
# cfg$items, so this adapts to any item set with no hard coding.
bvr_tbl |>
  mutate(item_a = factor(item_a, levels = cfg$items),
         item_b = factor(item_b, levels = cfg$items)) |>
  ggplot(aes(item_a, item_b, fill = bvr)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.1f", bvr)), size = 2.0) +
  scale_fill_viridis_c(name = "BVR", direction = -1) +
  labs(x = NULL, y = NULL) +
  theme_lca() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Design-based confidence intervals (computed here, presented in the segments script)

```{r jackknife-ci}
#| eval: !expr K_ready
options(survey.lonely.psu = "adjust")

# Report PSU structure so the analyst sees the degrees of freedom they actually have.
# Note: dplyr::count() is namespace-qualified because matrixStats (loaded for
# rowLogSumExps) also exports a count(), and it loads after dplyr here.
psu_structure <- dat_prepared |>
  dplyr::distinct(.data[[cfg$strata]], .data[[cfg$psu]]) |>
  dplyr::count(.data[[cfg$strata]], name = "psus_in_stratum")
n_psu_total <- nrow(dplyr::distinct(dat_prepared, .data[[cfg$psu]]))
n_strata    <- n_distinct(dat_prepared[[cfg$strata]])
lonely_any  <- any(psu_structure$psus_in_stratum < 2)

# Declare the design and build stratified jackknife replicate weights.
des <- svydesign(ids = as.formula(paste0("~", cfg$psu)),
                 strata = as.formula(paste0("~", cfg$strata)),
                 weights = as.formula(paste0("~", cfg$weight)),
                 data = dat_prepared, nest = TRUE)
rep_des <- as.svrepdesign(des, type = "JKn")

# Parameter bookkeeping as a TIBBLE, in the exact order flatten_fit() produces.
# This carries item/category/class as real columns, so nothing is recovered from
# string parsing later and the survey-question names are preserved downstream.
param_meta <- bind_rows(
  tibble(kind = "pi", item = NA_character_, category = NA_integer_, class = seq_len(K_star)),
  imap_dfr(cats, function(Cj, item_name)
    expand_grid(class = seq_len(K_star), category = seq_len(Cj)) |>   # category varies fastest
      transmute(kind = "rho", item = item_name, category, class))
)
param_names <- param_meta |>
  mutate(nm = if_else(kind == "pi",
                      sprintf("pi[%d]", class),
                      sprintf("rho[%s,cat%d,class%d]", item, category, class))) |>
  pull(nm)
flatten_fit <- function(fit) set_names(c(fit$pi, unlist(map(fit$rho, as.vector))), param_names)

# The statistic survey::withReplicates evaluates on each replicate: refit the weighted EM
# on the replicate weights, warm-started at the full-sample fit and aligned to it.
theta_fun <- function(w_rep, data) {
  inp <- make_inputs(data, cfg$items, cats)
  fit <- em_run(inp$Y, inp$OH, cats, w_rep, K_star,
                init = fit_w[c("pi", "rho")], maxit = 400L)
  flatten_fit(align_to(fit, fit_w))
}

rep_var <- withReplicates(rep_des, theta_fun)
est_vec <- flatten_fit(fit_w)
se_vec  <- sqrt(diag(vcov(rep_var)))

# Tidy table of every probability with its design-based 95% CI: item/category/class
# are columns, ready to filter, join, and plot.
ci_tbl <- param_meta |>
  mutate(estimate = as.numeric(est_vec), se = as.numeric(se_vec),
         lo = pmax(0, estimate - 1.96 * se),
         hi = pmin(1, estimate + 1.96 * se))

# What the design actually supports, then confirmation that every parameter got
# an interval. Intervals here are Wald on the probability scale, truncated to
# [0, 1]; the domain estimates in the segments script use the logit scale.
cat(sprintf(paste0("Design: %d strata, %d PSUs -> %d jackknife replicates",
                   " (%d residual df).\n%s",
                   "Design-based SEs computed for %d parameters",
                   " (%d prevalences, %d conditional probabilities).\n"),
            n_strata, n_psu_total, ncol(rep_des$repweights), n_psu_total - n_strata,
            ifelse(lonely_any,
                   "Warning: at least one stratum contributes a single PSU; survey.lonely.psu = 'adjust' is in force.\n",
                   ""),
            nrow(ci_tbl), K_star, nrow(ci_tbl) - K_star))
```

# Unweighted benchmark (poLCA) for the validation table

```{r polca-benchmark}
#| eval: !expr K_ready
# poLCA needs a data frame of items coded 1..C with no NA (it uses na.rm).
polca_df <- dat_prepared |> dplyr::select(all_of(cfg$items)) |> as.data.frame()
polca_f  <- as.formula(paste0("cbind(", paste(cfg$items, collapse = ", "), ") ~ 1"))

# --- substantive benchmark: poLCA at K_star, unweighted ---
set.seed(cfg$seed + 7L)
pl <- poLCA(polca_f, data = polca_df, nclass = K_star, nrep = 10,
            maxiter = 5000, na.rm = TRUE, verbose = FALSE)
pl_aligned <- align_to(list(pi = as.numeric(pl$P),
                            rho = map(pl$probs, t),
                            post = pl$posterior), fit_w)

# --- poLCA's own BIC-selected K across the candidate range ---
polca_bic <- furrr::future_map_dfr(cfg$K_range, function(K) {
  set.seed(cfg$seed + 10L + K)
  m <- poLCA(polca_f, data = polca_df, nclass = K, nrep = 5,
             maxiter = 5000, na.rm = TRUE, verbose = FALSE)
  tibble(K = K, poLCA_BIC = m$bic)
}, .options = furrr::furrr_options(seed = TRUE, packages = "poLCA"))
polca_K <- polca_bic |> slice_min(poLCA_BIC, n = 1) |> pull(K) |> first()

# --- validation: weighted EM with all weights = 1 should match poLCA ---
# fit_u (Section 5) IS the unit-weight fit, already aligned to fit_w; reusing it
# avoids a second multi-start refit of the same model.
fit_unit <- fit_u
max_pi_gap  <- max(abs(fit_unit$pi - pl_aligned$pi))
max_rho_gap <- max(map2_dbl(fit_unit$rho, pl_aligned$rho, ~ max(abs(.x - .y))))

# The trust check: at unit weights our EM and poLCA maximize the SAME likelihood,
# so any gap beyond optimizer tolerance is an implementation problem, not a
# modeling result. The segments script reprints the prevalence verdict alongside
# the weighted estimates.
cat(sprintf(paste0("VALIDATION (unit weights, our EM vs poLCA):\n",
                   "  max |prevalence gap|            = %.5f\n",
                   "  max |conditional-probability gap| = %.5f\n  -> %s\n",
                   "poLCA's own BIC-minimizing K over %s: %d",
                   " (evidence, not a decision; K stays cfg$K_force = %d).\n"),
            max_pi_gap, max_rho_gap,
            ifelse(max(max_pi_gap, max_rho_gap) < 0.005,
                   "MATCH: implementations agree.",
                   "MISMATCH: investigate before trusting the weighted fit."),
            paste0(min(cfg$K_range), "-", max(cfg$K_range)), polca_K, K_star))




ll_ours <- fit_unit$ll; ll_polca <- pl$llik
cat(sprintf("log-likelihood  ours = %.4f | poLCA = %.4f | difference = %+.6f\n",
            ll_ours, ll_polca, ll_ours - ll_polca))
cat(sprintf("convergence     ours = %s | poLCA used %d iterations\n",
            ifelse(isTRUE(fit_unit$converged), "converged", "NOT converged"),
            pl$numiter))

# Positional indexing on BOTH lists: our EM's rho is unnamed while the poLCA
# side is named, so name-based lookup silently returns NULL. arrayInd() finds
# the largest gap without assuming which() returns a matrix.
gaps <- map_dfr(seq_along(cfg$items), function(j) {
  a <- as.matrix(fit_unit$rho[[j]]); b <- as.matrix(pl_aligned$rho[[j]])
  d <- abs(a - b); idx <- arrayInd(which.max(d), dim(d))
  tibble(item = cfg$items[j], category = idx[1], segment = idx[2],
         ours = a[idx[1], idx[2]], polca = b[idx[1], idx[2]], gap = max(d))
}) |> arrange(desc(gap))
print(head(gaps, 5))
cat("segment sizes:", paste(sprintf("%.3f", fit_unit$pi), collapse = ", "), "\n")
```

# Saving the objects for the segments script

Tables go to parquet (arrow), model objects (lists) to rds; `survey_dat` itself is NOT saved, both scripts rebuild it identically from `survey_data_config.R`, which keeps one source of truth for the data.

```{r save-objects}
#| eval: !expr K_ready
saveRDS(fit_w,       file.path(cfg$out_dir, "fit_weighted.rds"))
saveRDS(fit_unit,    file.path(cfg$out_dir, "fit_unweighted.rds"))
saveRDS(pl_aligned,  file.path(cfg$out_dir, "polca_aligned.rds"))
arrow::write_parquet(enum,    file.path(cfg$out_dir, "enumeration.parquet"))
arrow::write_parquet(ci_tbl,  file.path(cfg$out_dir, "ci_tbl.parquet"))
arrow::write_parquet(bvr_tbl, file.path(cfg$out_dir, "bvr_tbl.parquet"))
cat("Saved fit and tables for K =", K_star, "to", cfg$out_dir, "\n")
```
