---
title: "Design-Weighted LCA: Data and Model Selection"
subtitle: "Institutional trust in Mexico: LAPOP AmericasBarometer 2023"
author: "Kevin Linares"
date: today
format:
  pdf:
    toc: false
    number-sections: false
    fig-width: 7
    fig-height: 4.5
    fig-align: center
    df-print: kable
    geometry: margin=1in
    fig-crop: false
execute:
  warning: false
  message: false
---

# Purpose

Presents the data and the recode audit, searches over K with the design-weighted pseudo-likelihood, and, once `cfg$K_force` is set, produces the fit diagnostics and the design-based intervals, then saves the fit for the segments script. Nothing here is interpretive.

```{r setup}
#| include: false
library(matrixStats); library(clue); library(survey); library(furrr)
library(poLCA); library(here); library(tidyverse)   # tidyverse LAST (masking)
source("survey_lca_source.R")
source("survey_data_config.R")
init_parallel(cfg)
K_ready <- !is.null(cfg$K_force)
```

# The data

Grey is the missingness the analysis rule removes. The recode audit shows what every source label became; anything that failed to match would have stopped the config.

```{r data-missingness}
#| fig-height: 5
#| fig-cap: "All respondents, all items, missingness shown explicitly (grey)."
plot_item_stack(survey_dat_full, items, "Items as read, with missingness")

cat(sprintf("%d of %d respondents enter the analysis.\n",
            nrow(cfg$data), nrow(survey_dat_full)))
```

```{r recode-audit}
lca_table(recode_audit,
          col.names = c("Variable", "Source label", "Recoded to", "n"),
          widths = c("6em", "21em", "9em", "3em"),
          caption = "Recode audit: how each source label was mapped.")
```

```{r data-dictionary}
dictionary |>
  mutate(responses = map_chr(responses, \(x) paste(x, collapse = " | "))) |>
  lca_table(col.names = c("Analysis name", "Source variable",
                          "Survey question", "Response labels"),
            widths = c("8em", "5em", "16em", "14em"),
            caption = "Item dictionary: analysis names, source variables, question wording, and the response labels the segment labeler reads.")
```

```{r prepare}
dat_prepared <- cfg$data
w_vec <- dat_prepared[[cfg$weight]]
n_obs <- nrow(dat_prepared)
n_eff <- sum(w_vec)^2 / sum(w_vec^2)
scale_ic <- n_obs / sum(w_vec)

cat(sprintf("Weights: CV = %.3f, unequal-weighting effect = %.2f, effective n = %.0f of %d.\n",
            sd(w_vec) / mean(w_vec), 1 + (sd(w_vec) / mean(w_vec))^2, n_eff, n_obs))
```

The unequal-weighting effect is what design weighting costs in variance. When it is close to 1, the weighted and unweighted fits will differ little and the gap reported later should be read as noise rather than correction.

# Searching over K

Every candidate K is fit from `cfg$n_starts` seeded starts. Seeds are generated in this session and passed to the workers, so sequential and parallel plans give identical results. `BIC` uses the achieved sample size and `BIC_eff` the effective sample size, which is the design-aware penalty; where they disagree, weight `BIC_eff` more heavily on a clustered sample.

```{r enumerate}
enum <- future_map(cfg$K_range, function(K) {
  fit <- fit_lca(dat_prepared, w_vec, cats, cfg$items, K, start_seeds(cfg, K))
  ll_n <- fit$ll * scale_ic
  d <- df_k(K, cats)
  tibble(K = K, logLik = ll_n, df = d,
         AIC = -2 * ll_n + 2 * d,
         BIC = -2 * ll_n + d * log(n_obs),
         BIC_eff = -2 * ll_n + d * log(n_eff),
         entropy = entropy_R2(fit$post, w_vec, K),
         converged = fit$converged)
}, .options = furrr_options(seed = NULL,
     packages = c("matrixStats", "clue", "purrr", "dplyr", "tidyr"))) |>
  list_rbind()

enum |>
  mutate(across(c(logLik, AIC, BIC, BIC_eff), \(x) round(x, 1)),
         entropy = round(entropy, 3)) |>
  lca_table(align = "r",
            caption = "Weighted fit statistics by number of segments. Nothing is selected here. Read BIC_eff and entropy together, then set cfg$K_force.")
```

```{r fig-enum}
#| fig-cap: "Enumeration diagnostics across K. BIC_eff penalizes on the effective sample size; entropy indexes classification sharpness. Neither decides alone."
enum |>
  select(K, BIC, BIC_eff, entropy) |>
  pivot_longer(-K, names_to = "criterion", values_to = "value") |>
  mutate(criterion = recode(criterion, BIC = "BIC (n)",
                            BIC_eff = "BIC (effective n)",
                            entropy = "Relative entropy")) |>
  ggplot(aes(K, value)) +
  geom_line(linewidth = 0.6, color = viridisLite::viridis(1, begin = 0.4)) +
  geom_point(size = 2.4, color = viridisLite::viridis(1, begin = 0.4)) +
  facet_wrap(~ criterion, scales = "free_y") +
  scale_x_continuous(breaks = cfg$K_range) +
  labs(x = "Number of segments (K)", y = NULL) +
  theme_lca()
```

```{r k-gate}
#| results: asis
if (!K_ready) cat("**`cfg$K_force` is `NULL`.** Review the evidence above, set a candidate K in `survey_data_config.R`, and re-render.\n")
```

# The chosen model

```{r fit-final}
#| eval: !expr K_ready
K_star <- as.integer(cfg$K_force)

fit_w <- fit_lca(dat_prepared, w_vec, cats, cfg$items, K_star,
                 start_seeds(cfg, K_star), maxit = 5000L, tol = 1e-10)
fit_u <- align_to(fit_lca(dat_prepared, rep(1, n_obs), cats, cfg$items, K_star,
                          start_seeds(cfg, K_star), maxit = 5000L, tol = 1e-10),
                  fit_w)

inp_full <- make_inputs(dat_prepared, cfg$items, cats)
post_w <- posterior_of(fit_w$pi, fit_w$rho, inp_full$Y)
colnames(post_w) <- paste0("post_segment", seq_len(K_star))
modal_int <- max.col(post_w, ties.method = "first")

dat_prepared <- bind_cols(dat_prepared, as_tibble(post_w))

cat(sprintf(paste0("Fitted K = %d on %d cases. Converged: %s.\n",
                   "Weighted log-likelihood (sum-to-n scale): %.1f | weighted entropy: %.3f\n"),
            K_star, n_obs,
            if (isTRUE(fit_w$converged)) "yes" else "NO, inspect before trusting",
            fit_w$ll * scale_ic, entropy_R2(post_w, w_vec, K_star)))

tibble(Segment = paste("Segment", seq_len(K_star)),
       `Weighted prevalence` = round(fit_w$pi, 3),
       `Unweighted prevalence` = round(fit_u$pi, 3),
       `Modal-assignment share` = round(as.numeric(prop.table(
         table(factor(modal_int, levels = seq_len(K_star))))), 3)) |>
  lca_table(align = "lrrr", widths = c("12em", "11em", "11em", "11em"),
            caption = "Segment prevalences at the chosen K. The modal-assignment share differs from the model prevalence whenever classification is uncertain.")
```

## Which items carry the segmentation

```{r item-discrimination}
#| eval: !expr K_ready
item_discrim <- item_discrimination(fit_w, cfg$items)

item_discrim |>
  mutate(discrimination = round(discrimination, 3)) |>
  lca_table(col.names = c("Item", "Discrimination"), align = "lr",
            caption = "Item discrimination: mean over segment pairs of the total variation distance between their response distributions. Bounded in [0, 1]. Zero means every segment answers alike.")
```

```{r fig-discrimination}
#| eval: !expr K_ready
#| fig-cap: "Item discrimination. Low bars are candidates for removal in a sensitivity refit."
item_discrim |>
  mutate(item = factor(item, levels = rev(item))) |>
  ggplot(aes(discrimination, item)) +
  geom_col(fill = viridisLite::viridis(1, begin = 0.45), width = 0.7) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "grey55") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "Discrimination (mean pairwise total variation distance)", y = NULL) +
  theme_lca()
```

```{r item-screen}
#| eval: !expr K_ready && cfg$run_item_screen
# Does each item earn its segment-dependent parameters? Compare the fitted model
# against one where item j is segment-independent. Under local independence that
# restricted model factorises into an LCA on the other items times item j's
# marginal, so the two BICs are comparable. Negative delta_bic means BIC prefers
# dropping the item's segment-dependence.
ll_full <- fit_w$ll * scale_ic
df_full <- df_k(K_star, cats)

marginal_ll <- function(item) {
  y <- as.integer(dat_prepared[[item]])
  p <- as.numeric(xtabs(w_vec ~ y))
  sum(w_vec * log((p / sum(p))[y]))
}

item_screen <- item_discrim |>
  mutate(delta_bic = future_map_dbl(item, function(j) {
    keep_items <- setdiff(cfg$items, j)
    fit_wo <- fit_lca(dat_prepared, w_vec, cats[keep_items], keep_items, K_star,
                      start_seeds(cfg, K_star))
    ll_restr <- (fit_wo$ll + marginal_ll(j)) * scale_ic
    df_restr <- df_k(K_star, cats[keep_items]) + (cats[[j]] - 1)
    (-2 * ll_restr + df_restr * log(n_eff)) - (-2 * ll_full + df_full * log(n_eff))
  }, .options = furrr_options(seed = NULL,
       packages = c("matrixStats", "clue", "purrr", "dplyr", "tidyr")))) |>
  arrange(delta_bic)

item_screen |>
  mutate(discrimination = round(discrimination, 3), delta_bic = round(delta_bic, 1)) |>
  lca_table(align = "lrr", col.names = c("Item", "Discrimination", "Delta BIC"),
            widths = c("14em", "10em", "8em"),
            caption = "Indicator screen, most droppable first. Delta BIC is BIC(item made segment-independent) minus BIC(full model), on the effective sample size. Negative flags the item. This is a ranking under a pseudo-likelihood, not a Bayes factor, so the usual BIC-difference thresholds do not apply.")
```

## Local independence: bivariate residuals

```{r bvr}
#| eval: !expr K_ready
bvr_tbl <- bvr_pairs(dat_prepared, w_vec, cfg$items, fit_w)

bvr_tbl |>
  mutate(bvr = round(bvr, 3)) |>
  slice_head(n = 10) |>
  lca_table(align = "llr", col.names = c("Item A", "Item B", "BVR"),
            widths = c("14em", "14em", "6em"),
            caption = "Largest bivariate residuals. BVR is the total variation distance between the weighted observed two-way table and the model-implied table, bounded in [0, 1] and zero under exact local independence. No valid reference distribution applies here, so this ranks rather than tests.")
```

```{r bvr-heatmap}
#| eval: !expr K_ready
#| fig-height: 6
#| fig-cap: "Bivariate residuals as total variation distance. Darker cells are pairs whose association the segments explain least well. Persistent dark pairs argue for dropping or merging an item, or trying K+1."
bvr_tbl |>
  mutate(item_a = factor(item_a, levels = cfg$items),
         item_b = factor(item_b, levels = cfg$items)) |>
  ggplot(aes(item_a, item_b, fill = bvr)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", bvr)), size = 2.0) +
  scale_fill_viridis_c(name = "BVR", direction = -1) +
  labs(x = NULL, y = NULL) +
  theme_lca() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Design-based confidence intervals

```{r jackknife-ci}
#| eval: !expr K_ready
design <- build_rep_design(dat_prepared, cfg)
des <- design$des
rep_des <- design$rep_des
design_df <- degf(rep_des)

cat(sprintf("Design: %d replicates, %d degrees of freedom.\n",
            ncol(rep_des$repweights), design_df))

# Parameter bookkeeping, in the order flatten_fit() produces.
param_meta <- bind_rows(
  tibble(kind = "pi", item = NA_character_, category = NA_integer_,
         segment = seq_len(K_star)),
  imap(cats, function(Cj, item_name)
    expand_grid(segment = seq_len(K_star), category = seq_len(Cj)) |>
      transmute(kind = "rho", item = item_name, category, segment)) |>
    list_rbind())

flatten_fit <- function(fit) c(fit$pi, unlist(map(fit$rho, as.vector)))

# Refit on each replicate, warm-started at the full-sample solution and aligned
# to it, at the same tolerance as the point estimate.
theta_fun <- function(w_rep) {
  fit <- em_run(inp_full$Y, inp_full$OH, cats, w_rep, K_star,
                init = fit_w[c("pi", "rho")], maxit = 1000L, tol = 1e-10)
  flatten_fit(align_to(fit, fit_w))
}

est_vec <- flatten_fit(fit_w)
se_vec <- sqrt(diag(replicate_variance(rep_des, theta_fun, est_vec)))
```

```{r ci-table}
#| eval: !expr K_ready
# Intervals on the logit scale by the delta method, back-transformed, with the
# design t critical value. Estimates at a boundary have no meaningful logit SE
# and fall back to Wald truncated to [0, 1].
crit <- qt(0.975, df = design_df)

ci_tbl <- param_meta |>
  mutate(estimate = as.numeric(est_vec), se = as.numeric(se_vec),
         boundary = estimate < 1e-3 | estimate > 1 - 1e-3,
         p_c = pmin(pmax(estimate, 1e-6), 1 - 1e-6),
         gse = se / (p_c * (1 - p_c)),
         lo = if_else(boundary, pmax(0, estimate - crit * se),
                      plogis(qlogis(p_c) - crit * gse)),
         hi = if_else(boundary, pmin(1, estimate + crit * se),
                      plogis(qlogis(p_c) + crit * gse))) |>
  select(-p_c, -gse)
```

# Unweighted benchmark

At unit weights the weighted EM and `poLCA` maximize the same function, so any gap beyond optimizer tolerance is an implementation problem rather than a modeling result.

```{r polca}
#| eval: !expr K_ready
polca_df <- dat_prepared |> select(all_of(cfg$items)) |> as.data.frame()
polca_f <- as.formula(paste0("cbind(", paste(cfg$items, collapse = ", "), ") ~ 1"))

set.seed(cfg$seed + 7L)
pl <- poLCA(polca_f, data = polca_df, nclass = K_star, nrep = 10,
            maxiter = 5000, na.rm = TRUE, verbose = FALSE)
pl_aligned <- align_to(list(pi = as.numeric(pl$P), rho = map(pl$probs, t),
                            post = pl$posterior), fit_w)

# Third candidate: our EM warm-started at poLCA's solution, so the comparison is
# between optima rather than stopping rules.
fit_from_pl <- em_run(inp_full$Y, inp_full$OH, cats, rep(1, n_obs), K_star,
                      init = list(pi = pl_aligned$pi, rho = pl_aligned$rho),
                      maxit = 5000L, tol = 1e-10)
fit_unit <- if (fit_from_pl$ll > fit_u$ll) align_to(fit_from_pl, fit_w) else fit_u

validation <- tibble(
  max_pi_gap = max(abs(fit_unit$pi - pl_aligned$pi)),
  max_rho_gap = max(map2_dbl(fit_unit$rho, pl_aligned$rho, \(a, b) max(abs(a - b)))))
validation$match <- max(validation$max_pi_gap, validation$max_rho_gap) < 0.005

cat(sprintf(paste0("VALIDATION (unit weights, our EM vs poLCA):\n",
                   "  max |prevalence gap|              = %.5f\n",
                   "  max |conditional-probability gap| = %.5f\n",
                   "  log-likelihood ours = %.4f | poLCA = %.4f\n  -> %s\n"),
            validation$max_pi_gap, validation$max_rho_gap, fit_unit$ll, pl$llik,
            if (validation$match) "MATCH: implementations agree."
            else "MISMATCH: investigate before trusting the weighted fit."))
```

```{r save}
#| eval: !expr K_ready
saveRDS(list(fit_w = fit_w, pl_aligned = pl_aligned, validation = validation,
             dat_prepared = dat_prepared, post_w = post_w, modal_int = modal_int,
             ci_tbl = ci_tbl, bvr_tbl = bvr_tbl, enum = enum,
             w_vec = w_vec, n_eff = n_eff, K = K_star),
        file.path(cfg$out_dir, "model_fit.rds"))
cat("Saved model_fit.rds for K =", K_star, "to", cfg$out_dir, "\n")
```
