---
title: "Design-Weighted LCA: Segments, Prediction, and Handoff"
subtitle: "Institutional trust in Mexico: LAPOP AmericasBarometer 2023"
author: "Kevin"
date: today
format:
  pdf:
    toc: false
    number-sections: true
    fig-width: 7
    fig-height: 4.5
    fig-align: center
    df-print: kable
    geometry: margin=1in
    fig-crop: false   # MiKTeX's pdfcrop is fragile; cropping is cosmetic only
execute:
  warning: false
  message: false
---

# Purpose

This script is the fast, interpretive half of the pipeline. It reads the fitted objects the modeling script saved to `{cfg$out_dir}`, drafts (or reads) the segment labels, presents the measurement model with design-based intervals alongside the unweighted `poLCA` benchmark and its validation verdict, predicts segment membership back to every respondent who answered enough items, profiles the segments on the configured demographics (one page each), and writes the analyst handoff: a labelled `.sav` plus CSVs. Throughout, a latent class is called a **segment**, a word-choice preference; the statistical object is unchanged.

```{r setup}
#| include: false
library(matrixStats); library(clue); library(survey); library(furrr)
library(arrow); library(tidyverse)   # tidyverse LAST (masking)
source("survey_lca_source.R")
source("survey_data_config.R")
stopifnot(!is.null(cfg$K_force))     # this script only runs on a chosen K
fit_w      <- readRDS(file.path(cfg$out_dir, "fit_weighted.rds"))
fit_unw    <- readRDS(file.path(cfg$out_dir, "fit_unweighted.rds"))
pl_aligned <- readRDS(file.path(cfg$out_dir, "polca_aligned.rds"))
ci_tbl     <- arrow::read_parquet(file.path(cfg$out_dir, "ci_tbl.parquet"))
validation <- arrow::read_parquet(file.path(cfg$out_dir, "validation.parquet"))
K_star     <- length(fit_w$pi)
prep_fit     <- prepare_items(cfg)      # rebuilt identically from the config script
dat_prepared <- prep_fit$dat_prepared
cats         <- prep_fit$cats
w_vec        <- prep_fit$w_vec

# Posteriors are recomputed here from the SAVED parameters (they lived in the
# modeling session, not on disk). Deterministic: no refitting, no seed needed.
inp_full  <- make_inputs(dat_prepared, cfg$items, cats)
post_w    <- posterior_of(fit_w$pi, fit_w$rho, inp_full$Y)
colnames(post_w) <- paste0("post_class", seq_len(K_star))
modal_int <- max.col(post_w, ties.method = "first")
dat_prepared <- dat_prepared |>
  bind_cols(as_tibble(post_w)) |>
  mutate(modal_class   = factor(modal_int, levels = seq_len(K_star)),
         max_posterior = post_w[cbind(seq_len(n()), modal_int)])

# The sampling design and its JKn replicates are rebuilt from cfg the same way
# the modeling script built them, so every interval here rests on the design.
options(survey.lonely.psu = "adjust")
des     <- svydesign(ids = as.formula(paste0("~", cfg$psu)),
                     strata = as.formula(paste0("~", cfg$strata)),
                     weights = as.formula(paste0("~", cfg$weight)),
                     data = dat_prepared, nest = TRUE)
rep_des <- as.svrepdesign(des, type = "JKn")
```

# What the segment labeler is asked

Before any label is trusted, here is the instrument verbatim: the system prompt (the persona) and one real user prompt, built for segment 1 from this survey's own wording and probabilities. Three properties to note while reading: the prompt carries only ONE segment (isolated calls, an empirical design: joint prompts confused near-neighbor segments); every probability is shown against the item's actual response labels; and rule 1 fences the survey context to referent resolution, so nothing outside the shown probabilities may be attributed to a segment. This instrument, including the segment terminology, is certified by `llm_label_obedience_experiment.R`; any edit to it obliges a re-run.

```{r prompt-example}
meta_preview <- item_meta(survey_dat, cfg$items, cfg$na_codes, questions)
cat("=========== SYSTEM PROMPT (persona) ===========\n")
cat(lca_persona(), "\n\n")
cat("=========== USER PROMPT (segment 1) ===========\n")
cat(prompt_class_label(fit_w, 1, meta_preview, cfg$items, cfg$survey_context), "\n")
```

# Naming the segments

One rule: `segment_labels.csv` in `{cfg$out_dir}` is used when it exists (validated against K); otherwise the LLM drafts once, one call per segment, and writes it, with any harmonizer edit recorded in `Label_draft`. **Editing that file is taking over the naming**; the labels below are drafts to verify against the profile heatmap.

```{r segment-labels}
# get_class_labels() implements the one rule above (survey_lca_source.R,
# Section 4). The cache file freezes the LLM output for reproducibility.
# cfg$data, NOT dat_prepared: recode_consecutive() strips attributes, so a
# recoded copy would silently cost the prompts all wording and labels.
segment_labels <- get_class_labels(fit_w, survey_dat, cfg$items, cfg,
                                 questions = set_names(dictionary$question,
                                                       dictionary$item))

knitr::kable(segment_labels,
             caption = "Segment labels and descriptions in use (segment_labels.csv in cfg$out_dir). Verify each against its response profile before quoting; edit the file to take over naming.")

# Display labels for every downstream table and figure: short, wrapped, fixed
# class order so factor levels are stable everywhere.
# Three display tiers, because one label format cannot serve every surface:
#   lab_of    wrapped factor, for FIGURE legends and facets (newlines are fine there)
#   lab_line  single line "Ck: Label", for SMALL tables (a newline inside a kable
#             cell breaks Typst layout)
#   lab_short "Ck" only, for LARGE tables; the class-label table above is the legend
lab_lev   <- stringr::str_wrap(paste0("S", segment_labels$K, ": ", segment_labels$Label), 22)
lab_of    <- function(k) factor(lab_lev[k], levels = lab_lev)
lab_line  <- function(k) paste0("S", k, ": ", segment_labels$Label[k])
lab_short <- function(k) paste0("S", k)
```

# The measurement model: inclusion probabilities, benchmark, and validation

One table carries the design-weighted segment prevalences with their JKn intervals, the unweighted `poLCA` prevalences, and the gap. The printed validation line is the trust check: at unit weights our pseudo-EM and `poLCA` maximize the same likelihood, so their prevalences must match to optimization tolerance before the weighted output deserves belief.

```{r tbl-inclusion}
# The verdict is READ from validation.parquet, frozen by the modeling script,
# so both documents print the same criterion on the same numbers: max of the
# prevalence gap and the conditional-probability gap, threshold 0.005.
cat(sprintf(paste0("VALIDATION (unit weights, pseudo-EM vs poLCA): ",
                   "max |pi gap| = %.5f, max |rho gap| = %.5f -> %s\n"),
            validation$max_pi_gap, validation$max_rho_gap,
            ifelse(validation$match, "MATCH (implementations agree)",
                   "MISMATCH - investigate before trusting weighted results")))

# ci_tbl columns (from the modeling script): kind, item, category, class,
# estimate, se, lo, hi. Renamed here to short names for the display code.
pi_ci <- ci_tbl |> dplyr::filter(kind == "pi") |>
  transmute(K = class, est = estimate, lo, hi)
incl_tbl <- pi_ci |>
  mutate(Segment = lab_line(K),
         `Weighted pi [95% CI]` = sprintf("%.3f [%.3f, %.3f]", est, lo, hi),
         `poLCA (unweighted)`   = sprintf("%.3f", pl_aligned$pi[K]),
         Gap = sprintf("%+.3f", est - pl_aligned$pi[K])) |>
  dplyr::select(Segment:Gap)
knitr::kable(incl_tbl, align = "lrrr",
             caption = "Segment inclusion probabilities: design-weighted estimates with JKn intervals, the unweighted poLCA benchmark, and the weighting gap. The validation line above certifies the implementations agree at unit weights.")
```

```{r fig-inclusion}
#| fig-cap: "Weighted segment prevalences with 95% JKn intervals (points and bars) against the unweighted poLCA benchmark (open diamonds). Gaps are what the design corrects."
pi_ci |>
  mutate(Segment = lab_of(K)) |>
  ggplot(aes(Segment, est)) +
  geom_pointrange(aes(ymin = lo, ymax = hi), linewidth = 0.5) +
  geom_point(aes(y = pl_aligned$pi[K]), shape = 5, size = 2.6) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = NULL, y = "Inclusion probability") +
  theme_lca() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

# Segment profiles: conditional response probabilities

```{r fig-profiles}
#| fig-width: 8
#| fig-height: 7.5
#| fig-cap: "P(response | segment) for every item. Panel headers carry the drafted segment labels; verify each label against its panel before quoting it."
ci_tbl |>
  dplyr::filter(kind == "rho") |>
  mutate(Segment = lab_of(class),
         item = factor(item, levels = cfg$items),
         est = estimate) |>
  ggplot(aes(factor(category), item, fill = est)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", est)), size = 2.1) +
  scale_fill_viridis_c(name = "P", limits = c(0, 1)) +
  facet_wrap(~ Segment, ncol = min(3, K_star)) +
  labs(x = "Response category", y = NULL) +
  theme_lca() +
  theme(axis.text.y = element_text(size = 7))
```

# Predicting segment membership

Prediction runs on the FULL preserved frame: local independence lets a missing item drop out of the within-segment product, so partial responders are predictable, with `cfg$min_items_predict` as the evidence floor (below it, no prediction rather than a guess). Items must be coded exactly as fitted; the config script guarantees that here.

```{r predict}
pred <- predict_segments(survey_dat_original, fit_w, cfg$items,
                         min_items = cfg$min_items_predict)
survey_scored <- bind_cols(survey_dat_original, pred)
cat(sprintf(
  "Predicted %d complete and %d incomplete cases of %d total (floor: %d items); mean max posterior among predicted = %.3f.\n",
  sum(!is.na(pred$segment) & survey_scored$keep),
  sum(!is.na(pred$segment) & !survey_scored$keep),
  nrow(survey_scored), cfg$min_items_predict,
  mean(pred$max_posterior, na.rm = TRUE)))
```

# Segment composition by demographics

BCH-corrected segment prevalence within each demographic level (Bolck, Croon & Hagenaars 2004; Vermunt 2010): modal assignments are reweighted by the inverse of the design-weighted classification-error matrix, so the cross-tabs are unbiased for true-segment composition rather than diluted by classification error. Intervals are JKn on the logit scale with the design t critical value; cells outside (0, 1), possible because BCH weights can be negative, fall back to truncated Wald and are footnoted. Step-1 parameter uncertainty is not propagated, the standard treatment; with entropy 0.907 the understatement is minor (Bakk, Oberski & Vermunt 2014). One demographic per page, table above plot, so nothing crowds.

```{r domain-setup}
# The segment-membership target is the soft posterior attached in the setup
# chunk; post_cols are the K posterior columns in segment order. des and rep_des
# (the design and its JKn replicates) were also built in setup.
post_cols <- paste0("post_class", seq_len(K_star))

# Fixed layout for the flattened statistic: every (demographic, level, class) cell,
# levels in factor order, class varying fastest within a level. dom_meta names and
# orders the estimate vector so nothing is recovered by string parsing later.
dom_meta <- map_dfr(cfg$aux, function(v)
  expand_grid(plevel = levels(dat_prepared[[v]]), class = seq_len(K_star)) |>
    mutate(pvar = v)) |>
  dplyr::select(pvar, plevel, class) |>
  mutate(idx = row_number())

# The statistic survey::withReplicates evaluates on each replicate: the BCH
# three-step correction. post and modal are FIXED at the full-sample
# measurement model (step-1 uncertainty not propagated, standard practice);
# the classification-error matrix D and the correction weights u are rebuilt
# inside every replicate from the replicate weights, so the intervals carry
# the design variance of the classification table and the cross-tab jointly.
# For each demographic: a level-indicator matrix, then the weighted mean of
# each BCH-corrected membership column within each level,
# crossprod(M, w * U) / crossprod(M, w), flattened to ONE numeric vector in
# dom_meta's order (level-major, class-minor) for the replicate covariance.
domain_theta <- function(w_rep, data) {
  P  <- as.matrix(data[, post_cols, drop = FALSE])      # n x K, fixed across replicates
  U  <- bch_weights(P, max.col(P, ties.method = "first"), w_rep)  # n x K BCH
  wU <- w_rep * U                                       # n x K, weighted corrections
  flat <- map(cfg$aux, function(v) {
    f   <- factor(data[[v]], levels = levels(dat_prepared[[v]]))
    M   <- model.matrix(~ f - 1)                        # n x L, one column per level
    num <- crossprod(M, wU)                             # L x K: corrected sums per level
    den <- as.numeric(crossprod(M, w_rep))              # L: weighted total per level
    as.vector(t(num / den))                             # level-major, class-minor
  })
  unlist(flat, use.names = FALSE)
}

dom_rep <- withReplicates(rep_des, domain_theta)        # design-based covariance of every cell
dom_est <- domain_theta(weights(des), dat_prepared)     # full-sample point estimates
dom_V   <- vcov(dom_rep)                                # full covariance, for the contrasts below
dom_se  <- sqrt(diag(dom_V))
stopifnot(length(dom_est) == nrow(dom_meta))
```

```{r domain-prevalence}
# Per-level BCH-corrected prevalence with a design-based interval: logit-scale
# delta method (se_logit = se / (p (1 - p))), back-transformed, with the design
# t critical value. BCH point estimates can fall outside (0, 1); the logit
# transform is undefined there, so those cells fall back to Wald truncated to
# [0, 1] and are flagged in `wald_fallback` for the table footnote.
crit <- qt(0.975, df = survey::degf(rep_des))
dom_tbl <- dom_meta |>
  mutate(p = as.numeric(dom_est), se = as.numeric(dom_se)) |>
  mutate(inside = p > 0 & p < 1,
         p_c = pmin(pmax(p, 1e-6), 1 - 1e-6),
         gse = se / (p_c * (1 - p_c)),
         lo  = if_else(inside, plogis(qlogis(p_c) - crit * gse),
                       pmax(0, p - crit * se)),
         hi  = if_else(inside, plogis(qlogis(p_c) + crit * gse),
                       pmin(1, p + crit * se)),
         wald_fallback = !inside) |>
  dplyr::select(pvar, plevel, class, p, se, lo, hi, wald_fallback)
```

```{r domain-pages}
#| results: asis
#| fig-height: 3.6
walk(cfg$aux, function(v) {
  d <- dom_tbl |> dplyr::filter(pvar == v)
  d |>
    transmute(Level = plevel, Segment = lab_short(class),
              cell = sprintf("%.2f [%.2f, %.2f]", p, lo, hi)) |>
    pivot_wider(names_from = Segment, values_from = cell) |>
    # format = "pipe" + cat(): guaranteed emission under results: asis, and a
    # Pandoc table that wraps instead of overflowing (the kable |> print()
    # pattern is silently swallowed by some render paths).
    knitr::kable(format = "pipe", align = str_c("l", strrep("r", K_star)),
                 caption = paste0("BCH-corrected segment prevalence within ", v,
                                  " (95% JKn intervals, logit scale; rows sum to one across segments; S1..S",
                                  K_star, " named in the segment-label table",
                                  if (any(d$wald_fallback))
                                    "; cells outside (0, 1) use truncated Wald intervals" else "",
                                  ").")) |>
    cat(sep = "\n")
  cat("\n\n")
  p <- d |>
    mutate(Segment = lab_of(class)) |>
    ggplot(aes(p, plevel, color = Segment)) +
    geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.35, linewidth = 0.4,
                    position = position_dodge(width = 0.6)) +
    scale_color_viridis_d(begin = 0.1, end = 0.8) +
    scale_x_continuous(limits = c(0, NA)) +
    labs(x = "Prevalence of segment membership", y = NULL, color = NULL,
         title = v) +
    theme_lca()
  print(p)
  cat("\n\n\\newpage\n\n")
})
```

# The handoff

Three artifacts: the labelled `.sav` for analysts (numeric `segment` with the LLM/analyst labels as value labels, `NA` below the prediction floor, plus `segment_label` as text and the diagnostics from prediction), and CSVs of the measurement model and the demographic results.

```{r handoff}
seg_labelled <- haven::labelled(
  survey_scored$segment,
  labels = set_names(segment_labels$K, segment_labels$Label),
  label  = "Segment (design-weighted LCA)")
out_sav <- survey_scored |>
  mutate(segment = seg_labelled,
         segment_label = ifelse(is.na(segment), NA_character_,
                                segment_labels$Label[survey_scored$segment]))
haven::write_sav(out_sav, file.path(cfg$out_dir, "survey_with_segments.sav"))

readr::write_csv(ci_tbl |>
                   mutate(segment_label = segment_labels$Label[class]),
                 file.path(cfg$out_dir, "measurement_model_with_CIs.csv"))
readr::write_csv(dom_tbl |>
                   transmute(demographic = pvar, level = plevel,
                             segment = class,
                             segment_label = segment_labels$Label[class],
                             prevalence = p, lo, hi),
                 file.path(cfg$out_dir, "segment_demographics.csv"))
cat("Wrote survey_with_segments.sav (", sum(!is.na(out_sav$segment)), "assigned ),",
    "measurement_model_with_CIs.csv, segment_demographics.csv to", cfg$out_dir, "\n")
```
