---
title: "Design-Weighted LCA: Segments, Profiling, and Handoff"
subtitle: "Institutional trust in Mexico: LAPOP AmericasBarometer 2023"
author: "Kevin Linares"
date: today
format:
  pdf:
    toc: false
    number-sections: true
    fig-width: 7
    fig-height: 4.5
    fig-align: center
    df-print: kable
    geometry: margin=1in
    fig-crop: false
execute:
  warning: false
  message: false
---

# Purpose

Reads the fit saved by the modeling script, drafts or reads the segment labels, presents the measurement model, scores every respondent who answered enough items, profiles the segments on demographics against the design, and writes the handoff. A latent class is called a **segment** throughout; the statistical object is unchanged.

The order matters. The measurement model was estimated on item-complete cases in the modeling script. Scoring happens next and covers every respondent above `cfg$min_items`, which is a larger set. All domain estimates are then computed on that larger set, which is standard three-step practice and matters here because item nonresponse on a trust battery is not random: it skews older, less educated, and more rural, so restricting the domain estimates to complete responders would bias the composition in a predictable direction.

```{r setup}
#| include: false
library(matrixStats); library(clue); library(survey); library(furrr)
library(here); library(tidyverse)   # tidyverse LAST (masking)
source("survey_lca_source.R")
source("survey_data_config.R")
init_parallel(cfg)

fit <- readRDS(file.path(cfg$out_dir, "model_fit.rds"))
fit_w <- fit$fit_w
pl_aligned <- fit$pl_aligned
ci_tbl <- fit$ci_tbl
K_star <- fit$K
post_cols <- paste0("post_segment", seq_len(K_star))
```

# What the segment labeler is asked

The instrument, verbatim: the system prompt and one real user prompt built from this survey's own wording and probabilities. Each call carries one segment only, because joint prompts confuse near-neighbor segments. Rule 1 fences the survey context to referent resolution, so nothing outside the shown probabilities may be attributed to a segment.

```{r prompt-example}
cat("=========== SYSTEM PROMPT ===========\n")
cat(wrap_text(lca_persona()), "\n\n")
cat("=========== USER PROMPT (segment 1) ===========\n")
cat(wrap_text(prompt_segment_label(fit_w, 1, dictionary, cfg$items,
                                   cfg$survey_context)), "\n")
```

# Naming the segments

`segment_labels.csv` in the output directory is used when it exists. Otherwise the model drafts once per segment and writes it, with any harmonizer edit recorded in `Label_draft`. Editing that file is taking over the naming. The labels below are drafts to verify against the profile heatmap.

```{r segment-labels}
segment_labels <- get_segment_labels(fit_w, dictionary, cfg$items, cfg)

lca_table(segment_labels, widths = c("2em", "11em", "28em"),
          caption = "Segment labels and descriptions in use. Verify each against its response profile before quoting.")

lab_lev <- str_wrap(paste0("S", segment_labels$K, ": ", segment_labels$Label), 22)
lab_of <- function(k) factor(lab_lev[k], levels = lab_lev)
lab_line <- function(k) paste0("S", k, ": ", segment_labels$Label[k])
lab_short <- function(k) paste0("S", k)
```

# The measurement model

Intervals here are the JKn intervals computed in the modeling script on the estimation sample. They describe the measurement model, not the scored sample.

```{r tbl-prevalence}
cat(sprintf(paste0("VALIDATION (unit weights, pseudo-EM vs poLCA): max |pi gap| = %.5f, ",
                   "max |rho gap| = %.5f -> %s\n"),
            fit$validation$max_pi_gap, fit$validation$max_rho_gap,
            if (fit$validation$match) "MATCH" else "MISMATCH, investigate"))

pi_ci <- ci_tbl |>
  filter(kind == "pi") |>
  transmute(K = segment, est = estimate, lo, hi)

pi_ci |>
  mutate(Segment = lab_line(K),
         `Weighted prevalence [95% CI]` = sprintf("%.3f [%.3f, %.3f]", est, lo, hi),
         `poLCA (unweighted)` = sprintf("%.3f", pl_aligned$pi[K]),
         Gap = sprintf("%+.3f", est - pl_aligned$pi[K])) |>
  select(Segment:Gap) |>
  lca_table(align = "lrrr", widths = c("13em", "13em", "9em", "5em"),
            caption = "Segment prevalences: design-weighted estimates with JKn intervals, the unweighted poLCA benchmark, and the gap. Read the gap against the unequal-weighting effect reported in the modeling script; when that is near 1 the gap is noise rather than correction.")
```

```{r fig-prevalence}
#| fig-cap: "Weighted segment prevalences with 95% JKn intervals against the unweighted poLCA benchmark (open diamonds)."
pi_ci |>
  mutate(Segment = lab_of(K)) |>
  ggplot(aes(Segment, est)) +
  geom_pointrange(aes(ymin = lo, ymax = hi), linewidth = 0.5) +
  geom_point(aes(y = pl_aligned$pi[K]), shape = 5, size = 2.6) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = NULL, y = "Prevalence") +
  theme_lca() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r fig-profiles}
#| fig-width: 8
#| fig-height: 7.5
#| fig-cap: "P(response | segment) for every item. Panel headers carry the drafted labels; verify each label against its panel before quoting it."
ci_tbl |>
  filter(kind == "rho") |>
  mutate(Segment = lab_of(segment), item = factor(item, levels = cfg$items)) |>
  ggplot(aes(factor(category), item, fill = estimate)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", estimate)), size = 2.1) +
  scale_fill_viridis_c(name = "P", limits = c(0, 1)) +
  facet_wrap(~ Segment, ncol = min(3, K_star)) +
  labs(x = "Response category", y = NULL) +
  theme_lca() +
  theme(axis.text.y = element_text(size = 7))
```

# Scoring segment membership

Local independence lets a missing item drop out of the within-segment product, so a respondent who answered `cfg$min_items` or more is scorable even though they were not in the estimation sample. Everything downstream runs on the scored frame.

```{r score}
pred <- predict_segments(survey_dat_full, fit_w, cfg$items, cfg$min_items)
survey_scored <- bind_cols(survey_dat_full, pred)

scored <- survey_scored |>
  filter(!is.na(segment)) |>
  mutate(segment = factor(segment, levels = seq_len(K_star)))

design <- build_rep_design(scored, cfg)
des <- design$des
rep_des <- design$rep_des
design_df <- degf(rep_des)
crit <- qt(0.975, df = design_df)

cat(sprintf(paste0("Estimation sample: %d item-complete cases.\n",
                   "Scored sample: %d of %d respondents (floor: %d items), ",
                   "a gain of %d cases for the domain estimates.\n",
                   "Domain design: %d replicates, %d degrees of freedom.\n"),
            nrow(fit$dat_prepared), nrow(scored), nrow(survey_scored),
            cfg$min_items, nrow(scored) - nrow(fit$dat_prepared),
            ncol(rep_des$repweights), design_df))
```

```{r score-overall}
# Overall segment prevalence on the scored frame, with the design effect reported
# here rather than per demographic level. deff is the design variance over the
# variance a simple random sample of the same size would give, computed directly
# so it does not depend on a finite-population correction; with self-weighting
# data the sum of weights equals n and survey's deff = TRUE returns Inf.
overall <- svymean(~ segment, rep_des)

overall_tbl <- tibble(K = seq_len(K_star),
                      p = as.numeric(coef(overall)),
                      se = as.numeric(SE(overall))) |>
  mutate(lo = pmax(0, p - crit * se), hi = pmin(1, p + crit * se),
         deff = se^2 / (p * (1 - p) / nrow(scored)))

overall_tbl |>
  transmute(Segment = lab_line(K),
            Prevalence = sprintf("%.3f (%.3f)", p, se),
            `95% CI` = sprintf("[%.3f, %.3f]", lo, hi),
            Deff = sprintf("%.2f", deff)) |>
  lca_table(align = "lrrr", widths = c("14em", "10em", "11em", "5em"),
            caption = "Overall segment prevalence on the scored frame, with JKn standard errors and design effects. Deff above 1 means clustering and weighting cost effective sample size relative to a simple random sample of the same size; below 1 means stratification bought some back.")
```

```{r score-quality}
scored |>
  group_by(n_items_answered) |>
  summarise(n = n(), mean_max_posterior = round(mean(max_posterior), 3),
            .groups = "drop") |>
  lca_table(align = "rrr", widths = c("10em", "6em", "12em"),
            col.names = c("Items answered", "n", "Mean max posterior"),
            caption = "Assignment quality by number of items answered. Respondents scored from fewer items carry flatter posteriors and so more classification error, which the domain estimates below pool over. If the mean falls sharply at the low end, raise cfg$min_items.")
```

# What the design costs

The design effect is the ratio of the variance under this design to the variance a simple random sample of the same size would give. A design effect of 2 means the design cost half the effective sample size; below 1 means it bought some. The table reports it twice: once for the full design and once for a clustering-only design that ignores the strata, so the two sources can be separated.

Three patterns are worth reading for.

**A design effect near the mean cluster size is the ceiling, and it is a specification check.** A variable that is constant within a sampling unit has an intracluster correlation of 1, so its design effect equals the mean cluster size exactly. Any variable measured at the cluster level rather than the respondent level should land there. If it does, the cluster identifier is specified correctly and the replicate variance is computing what it should. If a respondent-level variable lands there, that variable is effectively a cluster attribute and its precision claims are much weaker than its sample size suggests.

**A design effect below 1 means the variance is smaller than random sampling would produce**, which requires the composition of each cluster to be controlled rather than sampled. That is the signature of a quota at the final selection stage: if interviewers must fill a fixed number of respondents in some category per sampling point, the between-cluster variance for that category is suppressed. Precision on those variables is a fieldwork artifact, not information, so read their standard errors with that in mind and do not use them to demonstrate design-based variance estimation.

**If the two design effect columns agree, stratification is contributing nothing** and clustering accounts for all of the difference from simple random sampling. Coarse strata relative to the number of clusters will do this.

```{r design-effects}
srs <- svydesign(ids = ~1, weights = reformulate(cfg$weight), data = scored)
clus <- svydesign(ids = reformulate(cfg$psu), weights = reformulate(cfg$weight),
                  data = scored)

psu_n <- n_distinct(scored[[cfg$psu]])

deff_tbl <- map(cfg$aux, function(v) {
  f <- reformulate(v)
  full <- svymean(f, rep_des, na.rm = TRUE)
  tibble(variable = v,
         level = str_remove(names(coef(full)), paste0("^", v)),
         se_design = as.numeric(SE(full)),
         se_clus = as.numeric(SE(svymean(f, clus, na.rm = TRUE))),
         se_srs = as.numeric(SE(svymean(f, srs, na.rm = TRUE))))
}) |>
  list_rbind() |>
  mutate(deff = (se_design / se_srs)^2, deff_nostrata = (se_clus / se_srs)^2)

cat(sprintf(paste0("%d clusters, mean cluster size %.1f. A variable constant ",
                   "within a cluster has design effect %.1f; that is the ceiling.\n"),
            psu_n, nrow(scored) / psu_n, nrow(scored) / psu_n))

deff_tbl |>
  transmute(Variable = variable, Level = level,
            `SE design` = sprintf("%.4f", se_design),
            `SE SRS` = sprintf("%.4f", se_srs),
            Deff = sprintf("%.2f", deff),
            `Deff no strata` = sprintf("%.2f", deff_nostrata)) |>
  lca_table(align = "llrrrr",
            widths = c("8em", "9em", "7em", "7em", "4em", "6em"),
            caption = "Design effects for the profiling variables. SE design is the JKn standard error, SE SRS the standard error a simple random sample of the same size would give. Deff is their squared ratio; Deff no strata drops the stratification and keeps the clustering, so a match between the two columns means stratification contributes nothing.")
```

# Segment composition by demographics

One page per demographic: its marginal distribution, the segment prevalences within each of its levels, and a figure. Everything uses the scored frame.

**Design-based** applies the sampling weights and takes its variance from the JKn replicates. It uses hard modal assignment, so it is attenuated: differences between demographic levels are understated by roughly the classification error rate. **BCH-corrected** additionally reweights modal assignment by the inverse of the design-weighted classification error matrix (Bolck, Croon and Hagenaars 2004; Vermunt 2010), removing that attenuation at the cost of wider intervals. Step-1 parameter uncertainty is not propagated, the standard treatment.

The gap between them is what classification error costs. Attenuation always biases toward the marginal, so it always understates group differences: a difference that survives under modal assignment is real, but a null under modal assignment may be a real difference that attenuation erased. Read the design-based numbers as primary and the corrected numbers when the claim is that groups do not differ.

```{r domain-estimates}
# 1. marginal distribution of each demographic, among those who answered it.
# The design effect is reported once for the overall estimates in the scoring
# section, not per level.
marginals <- map(cfg$aux, function(v) {
  m <- svymean(reformulate(v), rep_des, na.rm = TRUE)
  raw <- scored |>
    filter(!is.na(.data[[v]])) |>
    count(.data[[v]], name = "n") |>
    set_names(c("level", "n"))
  tibble(variable = v,
         level = str_remove(names(coef(m)), paste0("^", v)),
         weighted = as.numeric(coef(m)),
         se = as.numeric(SE(m))) |>
    left_join(mutate(raw, level = as.character(level)), by = "level")
}) |>
  list_rbind()

# 2. segment prevalence within each level, straight from the survey package.
# svyby splits the design by the demographic and runs svymean(~segment) inside
# each level. Columns are read by position (by-variable, then K estimates, then
# K SEs) because svyby's naming for a factor outcome varies by version.
domain_marginals <- map(cfg$aux, function(v) {
  sb <- as.data.frame(svyby(~ segment, reformulate(v), rep_des, svymean,
                            na.rm = TRUE))
  vals <- sb[, -1, drop = FALSE]
  stopifnot(ncol(vals) == 2 * K_star)
  tibble(variable = v,
         level = rep(as.character(sb[[1]]), times = K_star),
         segment = rep(seq_len(K_star), each = nrow(sb)),
         p = as.numeric(as.matrix(vals[, seq_len(K_star)])),
         se = as.numeric(as.matrix(vals[, K_star + seq_len(K_star)])))
}) |>
  list_rbind() |>
  filter(!is.na(level)) |>
  mutate(lo = pmax(0, p - crit * se), hi = pmin(1, p + crit * se))

# 3. BCH-corrected version of the same quantity. Posteriors and modal assignment
# are fixed at the scored values; the error matrix and correction weights are
# rebuilt inside every replicate, so the intervals carry the design variance of
# the classification table and the cross-tab together.
dom_meta <- map(cfg$aux, function(v)
  expand_grid(plevel = levels(scored[[v]]), segment = seq_len(K_star)) |>
    mutate(pvar = v)) |>
  list_rbind() |>
  select(pvar, plevel, segment) |>
  mutate(idx = row_number())

P_fixed <- as.matrix(scored[, post_cols, drop = FALSE])
modal_fixed <- as.integer(scored$segment)

domain_theta <- function(w_rep) {
  wU <- w_rep * bch_weights(P_fixed, modal_fixed, w_rep)
  unlist(map(cfg$aux, function(v) {
    # indicator matrix with missing demographics contributing to no level
    M <- outer(as.character(scored[[v]]), levels(scored[[v]]), "==") + 0
    M[is.na(M)] <- 0
    as.vector(t(crossprod(M, wU) / as.numeric(crossprod(M, w_rep))))
  }), use.names = FALSE)
}

dom_est <- domain_theta(weights(des, "sampling"))
dom_V <- replicate_variance(rep_des, domain_theta, dom_est)
stopifnot(length(dom_est) == nrow(dom_meta))

dom_tbl <- dom_meta |>
  mutate(p = as.numeric(dom_est), se = sqrt(diag(dom_V)),
         inside = p > 0 & p < 1,
         p_c = pmin(pmax(p, 1e-6), 1 - 1e-6),
         gse = se / (p_c * (1 - p_c)),
         lo = if_else(inside, plogis(qlogis(p_c) - crit * gse), pmax(0, p - crit * se)),
         hi = if_else(inside, plogis(qlogis(p_c) + crit * gse), pmin(1, p + crit * se))) |>
  select(pvar, plevel, segment, p, se, lo, hi)

# Both estimators in one long frame, uncorrected then corrected.
est_levels <- c("Design-based", "BCH-corrected")

domain_all <- bind_rows(
  transmute(domain_marginals, variable, level, segment, p, se, lo, hi,
            estimator = "Design-based"),
  transmute(dom_tbl, variable = pvar, level = as.character(plevel), segment,
            p, se, lo, hi, estimator = "BCH-corrected")) |>
  mutate(estimator = factor(estimator, levels = est_levels))
```

```{r domain-pages}
#| results: asis
#| fig-height: 5.0
walk(cfg$aux, function(v) {
  tag <- str_replace_all(v, "_", "-")
  cat("\n\n##", v, "\n\n")

  marginals |>
    filter(variable == v) |>
    transmute(Level = level, n, Share = sprintf("%.3f (%.3f)", weighted, se)) |>
    lca_table(align = "lrr", widths = c("12em", "4em", "12em"),
              label = paste0("comp-", tag),
              caption = paste0("Composition by ", v,
                               ". Share is the design-weighted proportion among ",
                               "respondents who answered the item, with its JKn ",
                               "standard error.")) |>
    cat(sep = "\n")
  cat("\n\n")

  domain_all |>
    filter(variable == v) |>
    transmute(Level = level, Estimator = as.character(estimator),
              Segment = lab_short(segment), cell = sprintf("%.3f", p)) |>
    pivot_wider(names_from = Segment, values_from = cell) |>
    arrange(Level, match(Estimator, est_levels)) |>
    lca_table(align = str_c("ll", strrep("r", K_star)),
              widths = c("9em", "9em", rep("4em", K_star)),
              label = paste0("est-", tag),
              caption = paste0("Segment prevalence within each level of ", v,
                               ", uncorrected and BCH-corrected. Rows sum to one ",
                               "across segments. Intervals are in the figure below ",
                               "and the exported CSV.")) |>
    cat(sep = "\n")
  cat("\n\n")

  p <- domain_all |>
    filter(variable == v) |>
    mutate(Segment = lab_of(segment)) |>
    ggplot(aes(p, level, color = estimator, shape = estimator)) +
    geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.3, linewidth = 0.4,
                    position = position_dodge(width = 0.7)) +
    facet_wrap(~ Segment, ncol = min(3, K_star)) +
    scale_color_viridis_d(begin = 0.1, end = 0.75, direction = -1) +
    scale_shape_manual(values = c(`Design-based` = 16, `BCH-corrected` = 5)) +
    scale_x_continuous(limits = c(0, NA)) +
    labs(x = "Prevalence of segment membership", y = NULL, color = NULL,
         shape = NULL,
         caption = "95% intervals from JKn replicate variance and the design t. Filled circles: design-based. Open diamonds: BCH-corrected.") +
    theme_lca()
  print(p)
  cat("\n\n{{< pagebreak >}}\n\n")
})
```

## Do the segments differ across demographic levels?

Each level compared against the first level of its variable, within segment, using the replicate covariance so the difference carries its own standard error rather than being read off overlapping intervals. Holm adjustment within each demographic variable.

```{r contrasts}
contrasts <- dom_meta |>
  group_by(pvar, segment) |>
  mutate(ref_idx = idx[1], ref_level = plevel[1]) |>
  ungroup() |>
  filter(plevel != ref_level) |>
  mutate(diff = dom_est[idx] - dom_est[ref_idx],
         se = sqrt(dom_V[cbind(idx, idx)] + dom_V[cbind(ref_idx, ref_idx)] -
                     2 * dom_V[cbind(idx, ref_idx)]),
         p_raw = 2 * pt(-abs(diff / se), df = design_df)) |>
  group_by(pvar) |>
  mutate(p_adj = p.adjust(p_raw, method = "holm")) |>
  ungroup()

contrasts |>
  filter(p_adj < 0.05) |>
  arrange(pvar, p_adj) |>
  transmute(Variable = pvar, Comparison = paste(plevel, "vs", ref_level),
            Segment = lab_short(segment),
            Difference = sprintf("%+.3f", diff),
            `95% CI` = sprintf("[%+.3f, %+.3f]", diff - crit * se, diff + crit * se),
            `Adj. p` = sprintf("%.4f", p_adj)) |>
  lca_table(align = "lllrrr",
            widths = c("6em", "13em", "4em", "6em", "10em", "5em"),
            caption = "Segment composition differences reaching adjusted significance. Each row compares one demographic level against that variable's reference level within one segment. An empty table means no level difference survives adjustment.")
```

# The handoff

The `.sav` carries the segment assignment, the posterior probabilities, and the BCH correction weights for every scored respondent. The posteriors and correction weights are exported deliberately: cross-tabulating the modal assignment alone attenuates every association by the classification error rate, which is what the BCH tables above correct for.

```{r handoff}
U_full <- bch_weights(P_fixed, modal_fixed, weights(des, "sampling"))
colnames(U_full) <- paste0("bch_segment", seq_len(K_star))

out_sav <- survey_scored |>
  left_join(bind_cols(select(scored, all_of(cfg$id)), as_tibble(U_full)),
            by = cfg$id) |>
  mutate(segment_label = if_else(is.na(segment), NA_character_,
                                 segment_labels$Label[segment]),
         segment = haven::labelled(
           segment,
           labels = set_names(segment_labels$K, segment_labels$Label),
           label = paste("Segment (design-weighted LCA). Modal assignment:",
                         "cross-tabs on this variable alone are attenuated by",
                         "classification error. Use the post_segment columns or",
                         "the bch_segment weights for unbiased composition.")))

haven::write_sav(out_sav, file.path(cfg$out_dir, "survey_with_segments.sav"))

write_csv(mutate(ci_tbl, segment_label = segment_labels$Label[segment]),
          file.path(cfg$out_dir, "measurement_model_with_CIs.csv"))
write_csv(transmute(domain_all, demographic = variable, level, estimator, segment,
                    segment_label = segment_labels$Label[segment],
                    prevalence = p, se, lo, hi),
          file.path(cfg$out_dir, "segment_demographics.csv"))
write_csv(transmute(contrasts, demographic = pvar, level = plevel,
                    reference = ref_level, segment, difference = diff, se, p_adj),
          file.path(cfg$out_dir, "segment_contrasts.csv"))

cat(sprintf("Wrote survey_with_segments.sav (%d scored) and three CSVs to %s\n",
            nrow(scored), cfg$out_dir))
```
