---
title: "Public comments 2000-2025: text mining, EDA, and topic models"
subtitle: "Script 1 of 2: paragraph corpus, EDA, LDA, STM (searchK), topic ranking"
format:
  typst:
    toc: true
    fig-width: 9
    fig-height: 5.5
    fig-format: png    # Typst mishandles complex SVG figures; raster sidesteps it
    fig-dpi: 200
execute:
  warning: false
  message: false
---

```{=typst}
// Captioned tables render inside #figure blocks, which Typst treats as
// unbreakable by default; a table taller than the remaining page then
// overflows the bottom margin. This show rule lets figures break across
// pages, so tall tables continue on the next page instead of overlapping.
#show figure: set block(breakable: true)
```

```{=html}
<!--
PIPELINE MAP (two scripts)
  Script 1 (this file): raw comments -> paragraph corpus -> EDA -> LDA comparator
    -> STM searchK -> final STM -> ranked topics + paragraph-level theta
    -> optional LLM topic-label drafting (one call per topic, analyst-verified,
    frozen to disk, never fed back into any model), all saved to outputs/
    (parquet for tables, rds for model objects).
  Script 2 (llm_stance.qmd): reads outputs/, draws the PPS-by-size sample within
    selected-topic strata, runs the LLM stance/sentiment calls in 300-call
    checkpointed batches, produces design-weighted estimates.

PRODUCTION RUN CHECKLIST (simulation -> real data; per-section notes inline)
  1. Delete the SIMULATED DATA section below and replace it with one line:
       corpus_raw <- arrow::read_parquet("D:/path/to/comments.parquet")
     Columns required: atom_id (comment id), Year (integer 2000-2025),
     body_english (full comment text). Use an absolute path (here() anchors
     to the OneDrive project root, not the repo). The "Corpus validation"
     chunk stays: it stops the render with a named error if a column is
     missing or malformed, cheaper than a silent failure five chunks later.
  2. Read the "Newline structure diagnostic" output and apply its decision
     rules; if paragraphs are blank-line separated, set tx$para_split to
     "\\r?\\n{2,}". If most comments have NO newlines, stop and raise it
     (sentence-chunking fallback is a design change, not a config edit).
  3. Delete the outputs/ folder ONCE after the swap (caches track files, not
     code; anything left over is simulated results reloading silently).
  4. Render with tx$run_searchK = TRUE (the long run: hours, not minutes),
     read the four-panel diagnostics and the frontier plot, set tx$K, set
     tx$run_searchK = FALSE.
  5. Re-render (cheap now), read the FREX naming sheet and the representative
     paragraphs; optionally set tx$run_llm_labels = TRUE (OPENROUTER_API_KEY
     in .Renviron) for one drafting call per topic, then verify/edit drafts
     into outputs/topic_labels_verified.csv by hand.
  6. Set tx$selected_topics to the topics going into the stance work and
     re-render. Script 2 reads outputs/ from there.
  SIMULATION-ONLY SETTINGS to revisit in production: tx$K (placeholder 15),
  tx$selected_topics (NULL), tx$run_searchK, tx$run_llm_labels, and possibly
  tx$para_split / tx$para_min_tokens / tx$heading_max_tokens after the
  diagnostics. Everything else is mode-independent.
  Heavy steps are checkpointed: they run once and reload from outputs/ after.
  Delete the relevant outputs/ file to force a re-fit.
-->
```

What it does: defines `tx`, the single list every later chunk reads its settings from, attaches packages (tidyverse deliberately last so dplyr verbs win any masking contest), and defines the two checkpoint helpers. Why one config list: every analyst decision (splitter, thresholds, K, topic selection, LLM flags) is a line here, so adapting the script never means editing analysis code. Why checkpoints: `cache_rds()`/`cache_parquet()` compute an expression once, save it under `outputs/`, and reload it on every later render; the cost is that caches track files, not code, so anything edited upstream of a checkpoint requires deleting that file (or all of `outputs/`) once.
```{r config}
# ---------------------------------------------------------------------------
# CONFIGURATION (the only block you edit)
# ---------------------------------------------------------------------------
tx <- list(
  seed            = 2026L,
  out_dir         = "outputs",
  # paragraph atomization
  para_split      = "\\r?\\n+",     # newline-run splitter (CRLF or LF, single
                                    # or double breaks). PRODUCTION: if the
                                    # newline diagnostic shows single newlines
                                    # are soft wraps INSIDE paragraphs, change
                                    # to "\\r?\\n{2,}" (blank-line separated)
  para_min_tokens = 15L,            # body fragments shorter than this merge
                                    # into the PRECEDING paragraph
  heading_max_tokens = 12L,         # blocks this short WITHOUT sentence-ending
                                    # punctuation are treated as headings and
                                    # merged FORWARD into the paragraph they
                                    # introduce (see the atomization section)
  # searchK
  run_searchK     = TRUE,           # expensive; set FALSE after choosing K
  K_grid          = seq(4L, 80L, by = 3L),
  heldout_prop    = 0.10,           # share of docs with held-out words
  # final models
  K               = 15L,            # PLACEHOLDER for the simulation; set from
                                    # the searchK diagnostics on real data
  spline_df       = 5L,             # Year spline df for topical prevalence
  # parallelism
  workers         = 6L,
  # LLM topic-label drafting (one call per topic; drafts only, analyst verifies)
  run_llm_labels  = FALSE,          # TRUE once the API key is in .Renviron
  llm_model       = "openrouter/meta-llama/llama-4-maverick",
                                    # WORK SWAP: this ONE line. Any string not
                                    # starting with "openrouter/" is routed to
                                    # the OpenAI-compatible work endpoint (see
                                    # make_label_chat in the labeling section).
  # topic selection for Script 2 (set AFTER naming topics from the FREX sheet)
  selected_topics = NULL            # e.g. c(3L, 7L, 11L)
)

suppressPackageStartupMessages({
  library(arrow)        # parquet outputs
  library(quanteda)     # dfm construction and conversion to stm format
  library(topicmodels)  # LDA comparator
  library(stm)          # structural topic model
  library(tidytext)     # tidy tokenization and tidiers for LDA
  library(furrr)        # parallel map for the searchK grid
  library(tidyverse)    # loaded LAST so dplyr verbs win
})
theme_set(theme_minimal(base_size = 12))
set.seed(tx$seed)
dir.create(tx$out_dir, showWarnings = FALSE)

# Checkpoint helpers: compute once, then reload. Delete the file to re-fit.
# IMPORTANT: caches track FILES, not code. After editing anything upstream of a
# checkpoint (the simulator, the atomizer, the dfm), delete that checkpoint file,
# or delete the whole outputs/ folder to rebuild everything from scratch.
cache_rds <- function(path, expr) {
  if (file.exists(path)) return(readRDS(path))
  x <- force(expr); saveRDS(x, path); x
}
cache_parquet <- function(path, expr) {
  if (file.exists(path)) return(arrow::read_parquet(path))
  x <- force(expr); arrow::write_parquet(x, path); x
}
p_out <- function(f) file.path(tx$out_dir, f)
```

# Simulated data (DELETE THIS SECTION AT WORK)

PRODUCTION: delete this whole section and replace it with the one-line read, `corpus_raw <- arrow::read_parquet("D:/path/to/comments.parquet")` (absolute path; `here()` anchors to the OneDrive root, not the repo). Nothing after the validation gate knows or cares which source filled `corpus_raw`.

The block below fabricates \~1,200 multi-paragraph comments over 2000-2025 so the pipeline runs end to end. Each simulated comment mixes 2-4 of six themes across its paragraphs, mimicking the real problem: long comments spanning multiple topics. At work, delete everything in this section and read your real table into `corpus_raw` with columns `atom_id`, `Year`, `body_english`.

```{r simulate-data}
theme_vocab <- list(
  health    = c("vaccine", "hospital", "clinic", "disease", "outbreak", "screening",
                "nurses", "patients", "epidemic", "prevention", "mortality", "insurance",
                "treatment", "wellness", "sanitation", "immunization", "pandemic", "care"),
  economy   = c("jobs", "wages", "inflation", "budget", "taxes", "employment", "trade",
                "industry", "investment", "growth", "deficit", "subsidies", "tariffs",
                "market", "workers", "manufacturing", "exports", "recession"),
  security  = c("police", "crime", "border", "defense", "surveillance", "terrorism",
                "safety", "enforcement", "military", "threat", "protection", "patrol",
                "intelligence", "violence", "weapons", "emergency", "gangs", "courts"),
  environment = c("climate", "emissions", "pollution", "wildlife", "rivers", "forests",
                "renewable", "solar", "drought", "flooding", "conservation", "recycling",
                "pesticides", "wetlands", "carbon", "warming", "habitat", "cleanup"),
  education = c("schools", "teachers", "students", "curriculum", "tuition", "literacy",
                "classrooms", "funding", "graduation", "testing", "scholarships",
                "universities", "training", "libraries", "enrollment", "tutoring"),
  housing   = c("rent", "landlords", "eviction", "zoning", "homeless", "mortgage",
                "affordable", "construction", "tenants", "shelters", "development",
                "neighborhood", "property", "vacancy", "permits", "gentrification")
)
filler <- c("the", "a", "of", "and", "to", "in", "we", "our", "is", "that", "for",
            "should", "must", "will", "not", "more", "less", "very", "public",
            "community", "policy", "government", "people", "city", "council",
            "issue", "concern", "support", "oppose", "believe", "think", "need")

# Era-specific vocabulary so the corpus drifts over time the way real comments
# do; without it every term appears in every era, all idf terms are log(1) = 0,
# and the era TF-IDF view is degenerate by construction.
era_vocab <- list(
  c("y2k", "dialup", "anthrax", "millennium", "modem"),          # 2000-2004
  c("foreclosure", "stimulus", "hurricane", "bailout", "blog"),  # 2005-2009
  c("austerity", "fracking", "broadband", "drone", "hashtag"),   # 2010-2014
  c("opioid", "wildfire", "streaming", "rideshare", "vaping"),   # 2015-2019
  c("lockdown", "telework", "quarantine", "misinformation", "chatbot") # 2020-2025
)
era_of <- function(year) findInterval(year, c(2000, 2005, 2010, 2015, 2020))

make_paragraph <- function(theme, year) {
  n <- sample(40:120, 1)
  words <- sample(c(theme_vocab[[theme]], filler), n, replace = TRUE,
                  prob = c(rep(3, length(theme_vocab[[theme]])), rep(1, length(filler))))
  extra <- sample(era_vocab[[era_of(year)]], sample(2:5, 1), replace = TRUE)
  str_c(str_c(sample(c(words, extra)), collapse = " "), ".")
}
make_comment <- function(id) {
  yr       <- sample(2000:2025, 1, prob = 0.5 + seq(0, 1, length.out = 26))
  themes_i <- sample(names(theme_vocab), sample(2:4, 1))
  n_para   <- sample(2:8, 1)
  paras    <- map_chr(sample(themes_i, n_para, replace = TRUE), make_paragraph, year = yr)
  tibble(atom_id      = sprintf("A%05d", id),
         Year         = yr,
         body_english = str_c(paras, collapse = "\n\n"))
}
corpus_raw <- map_dfr(seq_len(1200), make_comment)
```

# Corpus validation (KEEP THIS AT WORK)

This gate runs on whatever `corpus_raw` holds, simulated or real. Every check names its failure, so a bad real-data read stops here with a diagnosis instead of surfacing as an opaque error downstream. Reading the real table is one line before this chunk, e.g. `corpus_raw <- arrow::read_parquet("path/to/comments.parquet")` or `readr::read_csv(..., show_col_types = FALSE)`; if the source columns have different names, rename them here once (`dplyr::rename(atom_id = ..., Year = ..., body_english = ...)`) so nothing downstream changes.

```{r validate-corpus}
required_cols <- c("atom_id", "Year", "body_english")
missing_cols  <- setdiff(required_cols, names(corpus_raw))
if (length(missing_cols) > 0)
  stop("corpus_raw is missing column(s): ", str_c(missing_cols, collapse = ", "))

stopifnot(
  "atom_id must be non-missing"            = !anyNA(corpus_raw$atom_id),
  "atom_id must be unique (one row per comment)" =
    !anyDuplicated(corpus_raw$atom_id),
  "Year must be integer-valued"            =
    is.numeric(corpus_raw$Year) && all(corpus_raw$Year == as.integer(corpus_raw$Year), na.rm = TRUE),
  "body_english must be character"         = is.character(corpus_raw$body_english)
)

# Non-fatal quality report: rows dropped here are removed with a count, not
# silently. NA years cannot enter the STM prevalence spline; empty bodies
# cannot atomize.
n_before   <- nrow(corpus_raw)
corpus_raw <- corpus_raw |>
  dplyr::filter(!is.na(Year), !is.na(body_english), str_squish(body_english) != "")
cat("Validation: kept", nrow(corpus_raw), "of", n_before, "comments;",
    "Year range", min(corpus_raw$Year), "-", max(corpus_raw$Year), "\n")
if (any(corpus_raw$Year < 2000 | corpus_raw$Year > 2025))
  warning("Years outside 2000-2025 present; confirm this is expected before ",
          "interpreting the Year spline.")
```

# Newline structure diagnostic

What it does: counts, per comment, how many line breaks the text carries and how many blank-line separations, then prints three comments with every break made visible as `<NL>`. Why: the atomizer can only split on whatever the source uses to mark paragraphs, and Excel-origin text (in-cell Alt+Enter breaks survive `readxl` and parquet as `\n`) can carry three different conventions that need three different responses. Decision rules, applied once on the real corpus:

1. `zero_newlines` small and the rest have breaks: keep the default `tx$para_split`. Break-free comments simply stay single paragraphs, which is correct.
2. `has_blank_lines` dominates AND the `<NL>` sample shows single breaks used as soft wraps inside paragraphs: paragraphs are blank-line separated; set `tx$para_split = "\\r?\\n{2,}"` in the config. Only a human reading the sample can tell soft wraps from true breaks, which is why the counts alone do not decide.
3. `zero_newlines` is most of the corpus: there is nothing to split on and the fallback is sentence-boundary chunking to a token budget, a design change to the measurement unit that must be decided deliberately, not patched in.

```{r newline-diagnostic}
newline_report <- corpus_raw |>
  mutate(body_lf  = str_replace_all(body_english, "\r\n", "\n"),
         n_breaks = str_count(body_lf, "\n"),
         n_blank  = str_count(body_lf, "\n{2,}")) |>
  summarise(comments        = n(),
            zero_newlines   = sum(n_breaks == 0),
            single_only     = sum(n_breaks > 0 & n_blank == 0),
            has_blank_lines = sum(n_blank > 0),
            median_breaks   = median(n_breaks))
knitr::kable(newline_report,
             caption = "Line-break conventions in the corpus; apply the decision rules in the prose above before the long render.")

corpus_raw |>
  dplyr::slice_sample(n = 3) |>
  pull(body_english) |>
  str_replace_all("\r?\n", "<NL>\n") |>
  cat(sep = "\n\n==== NEXT COMMENT ====\n\n")
```

# Paragraph atomization

What it does: splits each comment on `tx$para_split`, then reassembles short blocks into full paragraphs with a **direction-aware** merge, and assigns each paragraph a stable id. Why the paragraph: the estimand downstream is stance on a topic, but a comment can span several topics; LDA and STM define a "document" as whatever unit the document-term matrix is built over, and here that unit is the paragraph. Each paragraph keeps its parent `atom_id` and `Year`, and gains a `para_id` nested within the comment.

The merge runs in two passes because short blocks come in two kinds that belong in opposite directions:

- **Headings and subtitles merge FORWARD.** A block of at most `r tx$heading_max_tokens` tokens with no sentence-ending punctuation (so "Housing Concerns", "1. Introduction", "Background:") is a heading: it announces the content that follows, so its words are attached to the *next* body paragraph. Merging it backward, as a naive short-fragment rule would, contaminates the *previous* paragraph's topic signal with the next section's vocabulary, exactly the wrong assignment for a topic model. A heading with nothing after it falls back to the last body paragraph rather than being dropped.
- **Body fragments merge BACKWARD.** A block under `r tx$para_min_tokens` tokens that does end like a sentence ("No.", "I strongly oppose this.") is a continuation or an emphatic coda of what precedes it, so it joins the preceding paragraph, the original rule. Salutations and signature lines are absorbed the same way instead of surviving as junk paragraphs.

The heading heuristic is deliberately simple (length plus terminal punctuation) so its behavior is auditable; after the first real render, check it two ways: `quantile(paragraphs$n_tokens)` for the length distribution, and a read of a few dozen paragraphs for headings glued to the wrong side. Both `tx$heading_max_tokens` and `tx$para_min_tokens` are config knobs.

```{r atomize}
count_tokens <- function(x) str_count(x, "\\S+")

paragraphs <- cache_parquet(p_out("paragraphs.parquet"), {
  # PASS 1 (forward): attach headings to the paragraph they introduce.
  # rev(cumsum(rev(!heading))) gives every body block a distinct id that
  # DECREASES down the document, and gives a heading the same id as the next
  # body block, so grouping on it merges the heading forward. Trailing
  # headings get id 0 and pmax(1L) folds them into the last body block.
  blocks <- corpus_raw |>
    mutate(para_raw = str_split(body_english, tx$para_split)) |>
    dplyr::select(atom_id, Year, para_raw) |>
    unnest(para_raw) |>
    mutate(para_raw = str_squish(para_raw)) |>
    dplyr::filter(para_raw != "") |>
    group_by(atom_id, Year) |>
    mutate(heading = count_tokens(para_raw) <= tx$heading_max_tokens &
                     !str_detect(para_raw, "[.!?]['\")\\]]*$"),
           fwd_block = rev(cumsum(rev(!heading))) |> pmax(1L)) |>
    group_by(atom_id, Year, fwd_block) |>
    summarise(para_raw = str_c(para_raw, collapse = " "), .groups = "drop") |>
    dplyr::arrange(atom_id, Year, dplyr::desc(fwd_block))  # restore doc order

  # PASS 2 (backward): merge remaining short body fragments into the
  # preceding paragraph; a fragment gets the running index of the last
  # full-size paragraph before it.
  blocks |>
    group_by(atom_id, Year) |>
    mutate(full  = count_tokens(para_raw) >= tx$para_min_tokens,
           block = cummax(cumsum(full) * full) |> pmax(1L)) |>
    group_by(atom_id, Year, block) |>
    summarise(text = str_c(para_raw, collapse = " "), .groups = "drop_last") |>
    mutate(para_id  = row_number(),
           para_uid = str_c(atom_id, "-", sprintf("%02d", para_id)),
           n_tokens = count_tokens(text)) |>
    ungroup() |>
    dplyr::filter(n_tokens >= tx$para_min_tokens) |>
    dplyr::select(para_uid, atom_id, para_id, Year, n_tokens, text)
})
glimpse(paragraphs)
```

The corpus atomizes to `r nrow(paragraphs)` paragraphs from `r n_distinct(paragraphs$atom_id)` comments, a median of `r paragraphs |> dplyr::count(atom_id) |> pull(n) |> median()` paragraphs per comment.

# Exploratory analysis

What it does: counts comments and paragraphs per year. Why: this is the denominator picture for everything time-related downstream; thin years mean the STM Year spline is interpolating there and single-year claims are weak, and a sudden paragraph-per-comment shift flags a change in how posts were written or ingested rather than in what they discuss.
```{r eda-volume}
#| fig-cap: "Corpus volume over time: comments and paragraphs per year."
paragraphs |>
  group_by(Year) |>
  summarise(paragraphs = n(), comments = n_distinct(atom_id), .groups = "drop") |>
  pivot_longer(c(comments, paragraphs), names_to = "unit", values_to = "n") |>
  ggplot(aes(Year, n, fill = unit)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(end = 0.7) +
  labs(x = NULL, y = "count", fill = NULL,
       title = "Volume by year")
```

Tokenization uses tidytext: one row per word per paragraph, lowercased, with stop words removed for the frequency views (topic models below get the same treatment). `custom_stops` is the domain stop list: it starts empty and is filled at work with corpus boilerplate that carries no topical signal but would otherwise dominate every topic (e.g. the agency's own name, "comment", "docket", form-letter phrases). Fill it iteratively: render, read the top-terms plot and the topic word sheets, add offenders, delete `outputs/`, re-render.

```{r eda-tokens}
# Domain stop words: terms so ubiquitous in THIS corpus that they behave like
# function words. Empty for the simulation; populate from the real top-terms
# plot and FREX sheets. Removing a word here removes it from the topic models
# too (they consume the same `tokens` table).
custom_stops <- tibble(word = character())

tokens <- paragraphs |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>
  anti_join(custom_stops, by = "word") |>
  dplyr::filter(str_detect(word, "[a-z]"))

top_terms <- tokens |> dplyr::count(word, sort = TRUE)
```

What it does: the raw frequency view after stop-word removal. Why: this is the fastest read on what dominates the corpus, and it is the primary hunting ground for `custom_stops`; a term at the top that carries no topical meaning in this domain (the agency name, "comment", form-letter phrases) belongs in the domain stop list before the topic models see it.
```{r eda-topterms}
#| fig-cap: "Most frequent terms after stop-word removal."
top_terms |>
  slice_head(n = 25) |>
  mutate(word = fct_reorder(word, n)) |>
  ggplot(aes(n, word, fill = n)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c() +
  labs(x = "occurrences", y = NULL, title = "Top 25 terms")
```

TF-IDF weighs a term by how concentrated it is in a subset of the corpus. For term $t$ in document $d$,

$$\mathrm{tf\text{-}idf}(t, d) = \mathrm{tf}(t, d) \times \log \frac{N}{n_t},$$

where $\mathrm{tf}(t,d)$ is the term's share of document $d$'s tokens, $N$ is the number of documents, and $n_t$ the number of documents containing $t$. Terms frequent everywhere score near zero; terms frequent in few places score high. Using the Year as the "document" surfaces era-specific vocabulary. Terms present in every era carry $\log(N/n_t) = \log 1 = 0$ and are filtered out: an empty or thin panel here means the vocabulary barely drifts across eras, which is itself a finding, not an error.

```{r eda-tfidf}
#| fig-cap: "Highest TF-IDF terms by five-year era (Year as the document unit)."
era_tfidf <- tokens |>
  mutate(era = str_c(Year %/% 5 * 5, "-", Year %/% 5 * 5 + 4)) |>
  dplyr::count(era, word) |>
  bind_tf_idf(word, era, n) |>
  dplyr::filter(tf_idf > 0) |>   # terms in every era have idf = log(1) = 0 by definition
  group_by(era) |>
  slice_max(tf_idf, n = 8, with_ties = FALSE) |>
  ungroup()

# An empty result means no term is era-distinctive (every term appears in every
# era). That is a statement about the corpus, so report it; do not crash on it.
if (nrow(era_tfidf) == 0) {
  cat("No era-distinctive terms: every term appears in every era, so all idf",
      "values are log(1) = 0. If this happens on cached simulated data, delete",
      "the outputs/ folder to rebuild the corpus.\n")
} else {
  p_era <- era_tfidf |>
    mutate(word = reorder_within(word, tf_idf, era)) |>
    ggplot(aes(tf_idf, word, fill = era)) +
    geom_col(show.legend = FALSE) +
    scale_y_reordered() +
    scale_fill_viridis_d() +
    facet_wrap(~era, scales = "free_y") +
    labs(x = "TF-IDF", y = NULL, title = "Era-distinctive vocabulary")
  print(p_era)
}
```

Lexicon sentiment gives a cheap corpus-level trend line before any modeling. It is a coarse instrument (word-match counts, no negation handling) and is used here for description only; the defensible sentiment measurement happens in Script 2.

```{r eda-sentiment}
#| fig-cap: "Bing lexicon sentiment by year: share of matched words that are positive."
tokens |>
  inner_join(get_sentiments("bing"), by = "word") |>
  dplyr::count(Year, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(share_positive = positive / (positive + negative)) |>
  ggplot(aes(Year, share_positive)) +
  geom_line(color = viridis::viridis(1, begin = 0.3), linewidth = 1) +
  geom_point(color = viridis::viridis(1, begin = 0.3)) +
  geom_hline(yintercept = 0.5, linetype = 2, color = "grey50") +
  labs(x = NULL, y = "share positive", title = "Lexicon sentiment over time")
```

# Document-term matrix

Both topic models consume the same paragraph-by-term count matrix. Rare terms (fewer than 5 paragraphs) are dropped to stabilize estimation; the quanteda dfm converts directly to stm's input format, keeping `para_uid` alignment.

Adapting at work: `min_docfreq = 5` is calibrated to a corpus of a few thousand paragraphs. On the real corpus, check the vocabulary size after trimming (`ndoc(dfm_para)` and `nfeat(dfm_para)`); a vocabulary in the low thousands is comfortable for STM at this scale. If `nfeat` is much larger, raise the floor (or switch to a proportional floor, `dfm_trim(min_docfreq = 0.001, docfreq_type = "prop")`); if topics look starved of distinctive words, lower it. Also confirm no paragraph became empty after trimming: `quanteda::convert` drops empty documents, and if any are dropped the `meta` construction below still aligns because it is built from the surviving `rownames(dfm_para)`, but the paragraph counts reported earlier will no longer match the modeled corpus, which is worth a sentence in the write-up.

```{r dtm}
dfm_para <- tokens |>
  dplyr::count(para_uid, word) |>
  cast_dfm(para_uid, word, n) |>
  dfm_trim(min_docfreq = 5)

stm_input <- quanteda::convert(dfm_para, to = "stm")
# carry the covariate in dfm row order
meta <- tibble(para_uid = rownames(dfm_para)) |>
  left_join(paragraphs |> dplyr::select(para_uid, atom_id, Year), by = "para_uid")
stm_input$meta <- as.data.frame(meta)
```

# LDA comparator

Latent Dirichlet Allocation is the baseline generative model: each document $d$ has topic proportions $\theta_d \sim \mathrm{Dirichlet}(\alpha)$, each topic $k$ a term distribution $\beta_k \sim \mathrm{Dirichlet}(\eta)$, and each token is drawn by picking a topic from $\theta_d$ then a word from that topic's $\beta_k$. It is fit here at the same $K$ as the final STM as a sanity comparator: if STM's topics have no recognizable LDA counterparts, the STM structure (not the data) is doing the work, which is worth knowing.

```{r lda}
lda_fit <- cache_rds(p_out("lda_fit.rds"), {
  topicmodels::LDA(
    quanteda::convert(dfm_para, to = "topicmodels"),
    k = tx$K, control = list(seed = tx$seed)
  )
})
```

What it does: plots each LDA topic's highest-probability words (beta). Why: this is the comparison sheet against the STM topics later; recognizable one-to-one counterparts mean the topic structure is in the data, while STM topics with no LDA analogue are being shaped by the prevalence model and deserve extra scrutiny before selection.
```{r lda-terms}
#| fig-cap: "LDA: top terms per topic (beta, the topic-word distribution)."
tidy(lda_fit, matrix = "beta") |>
  group_by(topic) |>
  slice_max(beta, n = 7, with_ties = FALSE) |>
  ungroup() |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  scale_fill_viridis_d() +
  facet_wrap(~topic, scales = "free") +
  labs(x = expression(beta), y = NULL, title = str_c("LDA top terms, K = ", tx$K))
```

# STM: choosing K with a held-out grid

The structural topic model replaces LDA's global $\theta$ prior with a regression: topical prevalence depends on covariates, here a smooth function of Year, so topic attention can drift over 2000-2025 and that drift is estimated rather than averaged away.

`searchK`'s internal parallelism forks, which is unavailable on Windows, so the grid is parallelized explicitly with furrr: hold out a random 10% of word tokens, fit each $K$ on the remainder, and score four diagnostics.

Adapting at work: this is the long chunk. The grid has `r length(tx$K_grid)` values of $K$ and each fit's cost grows with $K$, the number of paragraphs, and the vocabulary, so budget hours, not minutes, for the first real-data run. Each furrr worker holds its own copy of the documents, so `tx$workers` is bounded by RAM as much as by cores; if the machine swaps, lower it. The run is checkpointed to `searchK_metrics.parquet`, so a crash or an aborted render costs only the incomplete grid, and once `tx$K` is chosen, set `tx$run_searchK = FALSE` so re-renders skip this entirely. Reading the panels: pick the smallest $K$ region where held-out likelihood has flattened and residual dispersion has stopped falling, then use the frontier plot to choose among candidates; if the frontier is still improving at $K = 80$, widen `tx$K_grid` rather than extrapolating.

- **Held-out likelihood**: log probability of the held-out words; higher is better, the most direct predictive criterion.
- **Semantic coherence**: for each topic's top words, how often they co-occur in the same documents, $\sum_{i<j} \log\{(D(v_i, v_j) + 1) / D(v_i)\}$, where $v_1, \dots, v_M$ are the topic's top words in descending probability, $D(v_i)$ is the number of documents containing $v_i$, and $D(v_i, v_j)$ the number containing both; the denominator is the higher-ranked word $v_i$. Higher (less negative) means topics whose words travel together.
- **Exclusivity**: how much a topic's top words belong to it alone. Coherence can be gamed by a few common words, so read the coherence-exclusivity frontier, not either alone.
- **Residuals**: multinomial dispersion of the residuals; values above 1 signal more topics (or more structure) are needed.

```{r searchK}
if (tx$run_searchK) {
  heldout <- cache_rds(p_out("heldout_split.rds"), {
    set.seed(tx$seed)
    stm::make.heldout(stm_input$documents, stm_input$vocab,
                      proportion = tx$heldout_prop, seed = tx$seed)
  })

  searchK_tab <- cache_parquet(p_out("searchK_metrics.parquet"), {
    plan(multisession, workers = tx$workers)
    on.exit(plan(sequential), add = TRUE)
    future_map_dfr(tx$K_grid, function(K) {
      fit <- stm::stm(heldout$documents, heldout$vocab, K = K,
                      prevalence = ~ s(Year, df = tx$spline_df),
                      data = stm_input$meta, init.type = "Spectral",
                      seed = tx$seed, verbose = FALSE)
      # All metrics are computed into plain locals BEFORE tibble(): a tibble
      # column named `heldout` would shadow the env object `heldout` for every
      # later argument in the same call (data masking), turning
      # heldout$documents into $ on an atomic column.
      ho <- stm::eval.heldout(fit, heldout$missing)$expected.heldout
      rd <- stm::checkResiduals(fit, heldout$documents)$dispersion
      sc <- mean(stm::semanticCoherence(fit, heldout$documents))
      ex <- mean(stm::exclusivity(fit))
      tibble(K = K, heldout = ho, residuals = rd, coherence = sc, exclusivity = ex)
    }, .options = furrr_options(seed = TRUE, packages = c("stm", "tibble")))
  })

  searchK_tab |>
    pivot_longer(-K, names_to = "metric", values_to = "value") |>
    ggplot(aes(K, value, color = metric)) +
    geom_line(linewidth = 1) + geom_point(size = 1.4) +
    scale_color_viridis_d(end = 0.85, guide = "none") +
    facet_wrap(~metric, scales = "free_y") +
    labs(x = "K (number of topics)", y = NULL,
         title = "searchK diagnostics over K = 4..80 by 3",
         subtitle = "Pick K at the coherence-exclusivity frontier where held-out flattens; then set tx$K")
}
```

What it does: plots each K's semantic coherence against its exclusivity. Why: the two metrics trade off (a few common words can buy coherence at the cost of exclusivity), so K is chosen at the frontier, up and to the right, among candidates the four-panel diagnostics already screened; a K dominated on both axes is never the right choice.
```{r searchK-frontier}
#| fig-cap: "Coherence-exclusivity frontier: up and to the right dominates."
if (tx$run_searchK) {
  searchK_tab |>
    ggplot(aes(coherence, exclusivity, color = K)) +
    geom_point(size = 2.5) +
    geom_text(aes(label = K), vjust = -0.8, size = 3, show.legend = FALSE) +
    scale_color_viridis_c() +
    labs(title = "Coherence vs exclusivity by K")
}
```

Set `tx$K` from these diagnostics and re-render (searchK reloads from its checkpoint, so the re-render is cheap). The simulation placeholder is K = `r tx$K`.

# Final STM

What it does: fits the final structural topic model at the chosen `tx$K`, with topical prevalence regressed on a smooth of Year (df = `r tx$spline_df`) and spectral initialization. Why the spline: attention to topics drifts over 2000-2025, and putting Year in the prevalence model estimates that drift instead of averaging it away; why spectral: it is deterministic given the data, removing initialization luck from a fit the whole downstream pipeline depends on. Checkpointed: re-renders reload the fit from `outputs/stm_fit.rds`.
```{r stm-final}
stm_fit <- cache_rds(p_out("stm_fit.rds"), {
  stm::stm(stm_input$documents, stm_input$vocab, K = tx$K,
           prevalence = ~ s(Year, df = tx$spline_df),
           data = stm_input$meta, init.type = "Spectral",
           seed = tx$seed, verbose = FALSE)
})
```

## Topic ranking by expected prevalence

A topic's expected corpus prevalence is the mean of its $\theta$ column over paragraphs: the share of corpus attention it commands. This is the ranking used to shortlist topics for the stance work in Script 2.

```{r topic-rank}
#| fig-cap: "Topics ranked by expected corpus prevalence (mean theta)."
theta <- stm_fit$theta
colnames(theta) <- str_c("topic_", seq_len(tx$K))

# labelTopics output goes into `lbl`, NOT `frex`: the tibble below defines a
# column named frex, which would shadow an env object of the same name for the
# prob column (the same data-masking trap fixed in the searchK chunk).
lbl <- stm::labelTopics(stm_fit, n = 7)
topic_summary <- cache_parquet(p_out("topic_summary.parquet"), {
  tibble(topic      = seq_len(tx$K),
         prevalence = colMeans(theta),
         frex       = map_chr(seq_len(tx$K), ~ str_c(lbl$frex[.x, ], collapse = ", ")),
         prob       = map_chr(seq_len(tx$K), ~ str_c(lbl$prob[.x, ], collapse = ", "))) |>
    arrange(desc(prevalence))
})

topic_summary |>
  mutate(label = fct_reorder(str_c("T", topic, ": ", str_trunc(frex, 35)), prevalence)) |>
  ggplot(aes(prevalence, label, fill = prevalence)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c() +
  labs(x = "expected prevalence", y = NULL,
       title = "Topic ranking (name these, then set tx$selected_topics)")
```

What it does: the naming sheet as a table, one row per topic with prevalence, FREX, and highest-probability words. Why FREX for naming: frequent-only words blur topics together, exclusive-only words are rare oddities; FREX balances the two, which is what a human needs to tell neighboring topics apart.
```{r topic-table}
#| tbl-colwidths: [8, 12, 40, 40]
topic_summary |>
  mutate(prevalence = round(prevalence, 3)) |>
  knitr::kable(caption = "Topics ranked by prevalence, with FREX (frequent AND exclusive) and highest-probability words. FREX words are the naming aid: they balance a word's frequency within the topic against its exclusivity to it.")
```

## Representative paragraphs

`findThoughts` pulls the paragraphs with the highest $\theta$ per topic, the fastest way to sanity-check a topic before naming it.

```{r thoughts}
#| tbl-colwidths: [10, 16, 74]
top4 <- topic_summary$topic[1:4]
thoughts <- stm::findThoughts(stm_fit, texts = meta$para_uid, n = 3, topics = top4)
tibble(topic = rep(top4, each = 3),
       para_uid = unlist(thoughts$docs)) |>
  left_join(paragraphs |> dplyr::select(para_uid, text), by = "para_uid") |>
  mutate(text = str_trunc(text, 150)) |>
  knitr::kable(caption = "Three highest-theta paragraphs for the four most prevalent topics.")
```

## LLM topic-label drafting (optional, one call per topic)

Naming topics is analyst judgment; the LLM's role here is strictly clerical: draft a candidate name and a one-sentence factual description from the same evidence the analyst reads (FREX words, highest-probability words, and the highest-$\theta$ paragraphs). Three properties keep this defensible:

- **One call per topic**, so the full run costs `r tx$K` requests, negligible against the API budget.
- **Drafts are frozen to disk and verified by hand.** The drafts land in `outputs/topic_labels_draft.parquet`; the analyst copies them to `outputs/topic_labels_verified.csv`, edits the `label` column, and that verified file is the only naming artifact anything downstream reads.
- **Never fed back.** Labels are display strings; they enter no model, no prompt, and no estimate. The drafting model is also blind to `Year` and to any covariate, so it cannot leak design information into the names.

Adapting at work: set `tx$run_llm_labels = TRUE` and put the key in `.Renviron` (`OPENROUTER_API_KEY` now; the work endpoint's key and base URL later). The provider swap is the ONE config line `tx$llm_model`: any string not starting with `"openrouter/"` is routed by `make_label_chat()` to `ellmer::chat_openai(base_url = Sys.getenv("WORK_LLM_BASE_URL"))`, the OpenAI-compatible pattern for the work API. Structured output (`ellmer::type_object`) constrains the response to the two fields, so there is no free-text parsing step.

```{r llm-labels}
#| tbl-colwidths: [8, 22, 58, 12]
# One structured drafting call per topic. Wrapped in a function so pasting the
# chunk into a console keeps every intermediate local (no stale globals).
make_label_chat <- function(model, system_prompt) {
  if (startsWith(model, "openrouter/")) {
    ellmer::chat_openrouter(
      model         = sub("^openrouter/", "", model),   # slug itself has a "/"
      system_prompt = system_prompt,
      params        = ellmer::params(temperature = 0),
      echo          = "none")
  } else {
    # WORK SWAP target: OpenAI-compatible endpoint; key + base URL from
    # .Renviron, never in code.
    ellmer::chat_openai(
      model         = model,
      base_url      = Sys.getenv("WORK_LLM_BASE_URL"),
      system_prompt = system_prompt,
      params        = ellmer::params(temperature = 0),
      echo          = "none")
  }
}

draft_topic_labels <- function(stm_fit, lbl, meta, paragraphs, K, model) {
  # Representative paragraphs for every topic, in dfm row order so findThoughts
  # indexes align with the fitted model.
  txt_in_dfm_order <- meta |>
    left_join(paragraphs |> dplyr::select(para_uid, text), by = "para_uid") |>
    pull(text)
  th <- stm::findThoughts(stm_fit, texts = txt_in_dfm_order, n = 3,
                          topics = seq_len(K))

  prompts <- map_chr(seq_len(K), function(k) str_c(
    "Topic ", k, " of ", K, " from a topic model of public comments.\n",
    "FREX words (frequent AND exclusive to this topic): ",
    str_c(lbl$frex[k, ], collapse = ", "), "\n",
    "Highest-probability words: ",
    str_c(lbl$prob[k, ], collapse = ", "), "\n",
    "Representative paragraphs:\n",
    str_c("- ", str_trunc(th$docs[[k]], 400), collapse = "\n")))

  schema <- ellmer::type_object(
    label       = ellmer::type_string(
      "Draft topic name, at most 4 words, noun phrase."),
    description = ellmer::type_string(
      "One factual sentence stating what these paragraphs discuss. Describe only what the words and paragraphs show; no speculation about intent, stance, or sentiment."))

  chat <- make_label_chat(model, system_prompt = str_c(
    "You draft names for statistical topics so an analyst can verify them. ",
    "Base the name and description ONLY on the words and paragraphs provided."))

  res <- ellmer::parallel_chat_structured(
    chat = chat, prompts = as.list(prompts), type = schema,
    max_active = 2, rpm = 30, on_error = "continue")

  tibble(topic = seq_len(K)) |>
    bind_cols(as_tibble(res) |>
                dplyr::select(dplyr::any_of(c("label", "description"))))
}

if (isTRUE(tx$run_llm_labels)) {
  labels_draft <- cache_parquet(p_out("topic_labels_draft.parquet"), {
    draft_topic_labels(stm_fit, lbl, meta, paragraphs,
                       K = tx$K, model = tx$llm_model)
  })
  labels_draft |>
    left_join(topic_summary |> dplyr::select(topic, prevalence), by = "topic") |>
    arrange(desc(prevalence)) |>
    knitr::kable(caption = str_c(
      "DRAFT topic labels from ", tx$llm_model,
      ". Verify against the FREX sheet and representative paragraphs, then save the edited table as outputs/topic_labels_verified.csv."))
} else {
  cat("LLM label drafting skipped (tx$run_llm_labels = FALSE). Name topics",
      "manually from the FREX sheet, or set the flag TRUE with the API key",
      "in .Renviron for one drafting call per topic.\n")
}
```

What it does: loads the analyst-verified label file when it exists and shows it; otherwise says so. Why a separate verified file: the LLM drafts are clerical suggestions, and the analyst's edited CSV is the only naming artifact Script 2 or any write-up may use, so verification is a file-existence contract rather than a promise.
```{r llm-labels-verified}
#| tbl-colwidths: [10, 25, 65]
# The verified file is the naming source of record. If your verified CSV kept
# the prevalence column, change the widths to [8, 22, 58, 12] to match. Create it by copying the
# draft parquet to CSV and editing:
#   arrow::read_parquet("outputs/topic_labels_draft.parquet") |>
#     readr::write_csv("outputs/topic_labels_verified.csv")
# then edit the label column by hand and keep only rows you have checked.
verified_path <- p_out("topic_labels_verified.csv")
if (file.exists(verified_path)) {
  topic_labels <- readr::read_csv(verified_path, show_col_types = FALSE)
  topic_labels |>
    knitr::kable(caption = "Verified topic labels (analyst-edited; the file Script 2 and all write-ups should use).")
} else {
  cat("No outputs/topic_labels_verified.csv yet; topics remain numbered.",
      "Drafts, if generated, are NOT used downstream until verified.\n")
}
```

## Prevalence over time

`estimateEffect` regresses each topic's $\theta$ on the Year spline, giving the estimated attention trend with uncertainty from the model's approximate posterior.

```{r effects}
#| fig-cap: "Estimated topical prevalence over time, top four topics."
# estimateEffect subsets its metadata by every variable name appearing in the
# formula, so tx$spline_df inside the formula becomes a lookup for a metadata
# column named "tx" and fails. Build the formula with literal numbers instead.
eff_formula <- as.formula(str_c("1:", tx$K, " ~ s(Year, df = ", tx$spline_df, ")"))
eff <- cache_rds(p_out("stm_year_effects.rds"), {
  stm::estimateEffect(eff_formula, stm_fit,
                      metadata = stm_input$meta, uncertainty = "Global")
})
eff_tab <- map_dfr(top4, function(k) {
  pd <- plot(eff, covariate = "Year", topics = k, method = "continuous",
             printlegend = FALSE, omit.plot = TRUE) |> suppressWarnings()
  tibble(topic = k, Year = pd$x,
         est = as.numeric(pd$means[[1]]),
         lo  = pd$ci[[1]][1, ], hi = pd$ci[[1]][2, ])
})
eff_tab |>
  ggplot(aes(Year, est, color = factor(topic), fill = factor(topic))) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, color = NA) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(end = 0.85, name = "topic") +
  scale_fill_viridis_d(end = 0.85, guide = "none") +
  labs(x = NULL, y = "expected prevalence", title = "Topic attention 2000-2025")
```

# Handoff to Script 2

What it does: assembles the paragraph-level table joined with the STM theta matrix and writes the full contract of files Script 2 reads. Why theta travels with the text: Script 2 assigns comments to topic strata from these paragraph shares and needs ids, Year, token counts, and text in one aligned table; nothing in Script 2 recomputes anything from this script.
```{r handoff}
theta_out <- cache_parquet(p_out("paragraph_theta.parquet"), {
  bind_cols(meta, as_tibble(theta)) |>
    left_join(paragraphs |> dplyr::select(para_uid, para_id, n_tokens, text),
              by = "para_uid")
})
cat("outputs/ written:\n",
    "- paragraphs.parquet          para_uid, atom_id, para_id, Year, n_tokens, text\n",
    "- paragraph_theta.parquet     the above + topic_1..topic_K (STM theta)\n",
    "- topic_summary.parquet       topic, prevalence, frex, prob (naming sheet)\n",
    "- topic_labels_draft.parquet  LLM drafts (only if tx$run_llm_labels)\n",
    "- topic_labels_verified.csv   analyst-verified names (created by hand)\n",
    "- stm_fit.rds, lda_fit.rds, stm_year_effects.rds, searchK_metrics.parquet\n")
```

Script 2 consumes `paragraph_theta.parquet`, `topic_summary.parquet`, and, when present, `topic_labels_verified.csv`. Before running it: name the topics (LLM drafts optional, verification mandatory), choose the ones for stance analysis, set `tx$selected_topics` here, re-render, and write one stance proposition per selected topic there. Script 2's three parts are the LLM API setup (OpenRouter now, the work endpoint by the same one-line swap used above), the sampling design (stratified PPS of comments by token count within selected-topic strata, so the LLM sees a designed sample rather than the full corpus), and the stance measurement with design-weighted estimation.
