---
title: "Topic and Sentiment Evolution in a Politician's Policy Commentary"
subtitle: "A defensible tidytext + topic-modeling + sentiment workflow under a Total Survey Error frame"
author: "Kevin"
date: today
format:
  pdf:
    toc-depth: 3
    number-sections: true
    colorlinks: true
    fig-pos: "H"
    geometry: margin=1in
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breaknonspaceingroup,commandchars=\\\{\}}
keep-tex: false
execute:
  echo: true
  warning: false
  message: false
  eval: true
knitr:
  opts_chunk:
    dev: "cairo_pdf"
    fig.align: "center"
    fig.width: 7.5
    fig.height: 4.8
---

```{r}
#| label: setup
#| include: false

library(tidyverse)   # dplyr, tidyr, purrr, ggplot2, stringr, forcats
library(tidytext)    # unnest_tokens(), bind_tf_idf(), get_sentiments(), cast_dtm(), tidy()
library(topicmodels) # LDA() (course-aligned topic model)
library(stm)         # Structural Topic Model: covariate-aware prevalence
library(srvyr)       # tidy interface to {survey}; design-based estimation
library(survey)      # Horvitz-Thompson / weighted estimation engine
library(clue)        # solve_LSAP(): Hungarian algorithm for label alignment
library(glue)        # text templating for the simulation
library(scales)      # axis formatting

theme_set(theme_minimal(base_size = 11))

set.seed(20140101)   # global seed; chunk-level seeds set where stochastic
```

# Purpose and analytic framing

This document specifies a reproducible pipeline for studying **how the policy
topics and the sentiment in a politician's public commentary change over time**.
The text-mining methods follow Silge & Robinson, *Text Mining with R: A Tidy
Approach* (the tidytext book); the topic and sentiment estimates are then wrapped
in a survey-inference frame so the "we did not collect everything" reality is
handled explicitly rather than ignored.

The corpus is treated not as a complete record but as a **sample from an
unobserved target population** of all posts the politician ever made. That
reframing is the core of the workflow's defensibility and maps the problem onto
the **Total Survey Error (TSE)** framework used for survey data:

- **Coverage / frame error.** Which posts could *ever* enter the corpus? Deleted,
  edited, platform-gap, and private posts are outside the frame; their count is
  unknown.
- **Selection / "nonresponse" error.** Among posts that exist, only a subset is
  observed. If capture depends on *when* a post was made (older posts churn off
  platforms) or on *what it is about* (controversial posts get deleted), naive
  prevalence and sentiment estimates are biased.
- **Measurement error.** A topic model and a sentiment score are error-prone
  *measurements*. Lexicon sentiment in particular is unigram-based and misreads
  negation ("not good"); the book is explicit about this, and we operationalize
  the caution as a diagnostic. Vocabulary and framing also drift over a decade,
  threatening measurement comparability (the textual analogue of measurement
  non-invariance).
- **Processing error.** Tokenization, stop-word lists, stemming, and chunk size
  are analyst choices that propagate downstream.
- **Adjustment / estimation.** Weighting corrects selection bias *if* the
  selection mechanism is modeled correctly.

The plan: (1) simulate a population corpus with a **known** time-varying topic
structure *and* a **known** sentiment (valence) structure; (2) impose a **known**
content-dependent selection mechanism to produce the observed sample; (3) run the
tidytext pipeline (frequencies, tf-idf, topics, lexicon sentiment, negation
diagnostic); (4) recover topic and sentiment trajectories and **validate them
against ground truth**; (5) add an optional AI module using a **local** model
(Gemma via Ollama through ellmer) for topic labeling and sentiment, cross-checked
against the lexicon; (6) show how design weighting corrects selection bias for
both prevalence and sentiment, and how sensitive that correction is. Because the
data-generating process is known, the workflow can demonstrate that it recovers
truth before it is pointed at real data.

::: {.callout-note}
## Why simulate first
On real found-text data the ground truth is unknown, so you cannot tell whether
the pipeline is recovering structure or inventing it. Validating against a known
data-generating process is the standard defense. Section 10 lists exactly which
lines to swap for the real corpus.
:::

# Reproducibility: package installation

```{r}
#| label: install
#| eval: false

# Run once. {stm} pulls in {tm} and {SnowballC}.
install.packages(c(
  "tidyverse", "tidytext", "topicmodels", "stm", "tm", "SnowballC",
  "srvyr", "survey", "clue", "glue", "scales"
))

# Sentiment lexicons: "bing" ships with tidytext; "afinn"/"nrc" require {textdata}
# and a ONE-TIME interactive license acceptance. Run these once interactively:
install.packages("textdata")
# textdata::lexicon_afinn()   # accept license at the prompt
# textdata::lexicon_nrc()     # accept license at the prompt

# Optional AI module (Section 9): local LLMs via Ollama + ellmer.
install.packages(c("ellmer", "irr"))
#  - Install Ollama (https://ollama.com), then pull a local model, e.g.:
#       ollama pull gemma3        # or your local gemma/llama tag
#  - Confirm the exact local tag from R with ellmer::models_ollama().
```

# Simulation design (the data-generating process)

All structural parameters live in one place so the simulation is fully specified
and tunable.

```{r}
#| label: dgp-params

dgp <- list(
  years          = 2014:2025,        # decade of commentary
  posts_per_year = 150L,             # "hundreds per year"; raise for realism
  K              = 6L,               # number of latent policy topics
  theta_noise_sd = 0.55,             # document-level dispersion of topic mix
  n_sent_range   = 3:6               # sentences per post (paragraph length)
)

topic_names <- c(
  "Economy & Taxation",
  "Healthcare",
  "Immigration & Borders",
  "Climate & Energy",
  "National Security & Defense",
  "Education"
)
```

## Topic vocabularies and sentence templates

Each latent topic is a small bank of distinctive content words; overlap between
banks is minimized so recovery is unambiguous (on real data, topics overlap and
recovery is harder, a limitation noted in Section 11). Sentences are assembled
from templates carrying realistic function-word noise so the downstream
stop-word removal has something to do.

```{r}
#| label: dgp-vocab

topic_vocab <- list(
  c("economy","taxes","taxation","budget","deficit","inflation","wages",
    "growth","tariffs","investment","fiscal","revenue","spending","debt"),
  c("healthcare","hospitals","insurance","patients","doctors","nurses",
    "coverage","clinics","premiums","prescriptions","treatment","medicine"),
  c("immigration","border","migrants","asylum","visas","citizenship",
    "refugees","deportation","integration","quotas","residency","newcomers"),
  c("climate","emissions","renewables","solar","wind","carbon","pollution",
    "sustainability","fossil","electricity","grid","greenhouse"),
  c("defense","military","terrorism","intelligence","cyber","alliances",
    "sovereignty","troops","surveillance","deterrence","drones","missiles"),
  c("education","schools","students","teachers","universities","curriculum",
    "tuition","literacy","classrooms","scholarships","graduates","learning")
)
names(topic_vocab) <- topic_names

# Neutral templates (function-word noise + two topic content words).
templates <- c(
  "We must address {w1} and {w2} for families across the country.",
  "My position on {w1} is clear: the government should act on {w2}.",
  "Today I spoke in Parliament about {w1} and our {w2} policy.",
  "For too long we have debated {w1} and the question of {w2}.",
  "Our plan this session covers {w1} alongside {w2}.",
  "Citizens have raised concerns about {w1} and {w2}.",
  "This budget cycle we will turn to {w1} and {w2}.",
  "The national conversation on {w1} and {w2} continues."
)
```

## Time-varying topic prevalence (ground truth)

Each topic gets an unnormalized intensity as a smooth function of the year index
$t = 0, 1, \dots$, mapped to a prevalence vector $\pi(t)$ via a softmax:

$$
\pi_k(t) = \frac{\exp\{\eta_k(t)\}}{\sum_{j=1}^{K}\exp\{\eta_j(t)\}} .
$$

Planted trajectories: **Economy** declining, **Healthcare** and **Climate**
rising, **Immigration** spiking mid-period, **Security** with a later bump,
**Education** roughly flat. These are the curves the pipeline must recover.

```{r}
#| label: dgp-prevalence

t_idx <- seq_along(dgp$years) - 1L   # 0..(T-1)

eta <- tibble(
  t        = t_idx,
  economy  = 2.0 - 0.16 * t,
  health   = 0.4 + 0.20 * t,
  immig    = 2.3 * exp(-0.5 * ((t - 3) / 1.6)^2),
  climate  = 0.1 + 0.24 * t,
  security = 0.6 + 1.1 * exp(-0.5 * ((t - 7) / 1.3)^2),
  educ     = 0.9 - 0.02 * t
)

softmax_rows <- function(m) {
  ex <- exp(m - apply(m, 1, max))
  ex / rowSums(ex)
}

pi_mat <- eta |>
  dplyr::select(-t) |>
  as.matrix() |>
  softmax_rows()
colnames(pi_mat) <- topic_names
rownames(pi_mat) <- dgp$years

truth_prevalence <- pi_mat |>
  as_tibble(rownames = "year") |>
  mutate(year = as.integer(year)) |>
  pivot_longer(-year, names_to = "topic", values_to = "true_prev")
```

```{r}
#| label: fig-truth
#| fig-cap: "Planted (ground-truth) topic prevalence by year."

truth_prevalence |>
  mutate(topic = factor(topic, levels = topic_names)) |>
  ggplot(aes(year, true_prev, color = topic)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.2) +
  scale_color_viridis_d(end = 0.92) +
  scale_x_continuous(breaks = dgp$years) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = NULL, y = "Share of attention", color = NULL,
       title = "Ground-truth topic prevalence") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Time-varying sentiment / valence (ground truth)

Sentiment is planted separately from topic. Each topic has a baseline valence,
and overall tone **sours over time** (a plausible, recoverable trend). For a
sentence on topic $k$ in year-index $t$, the latent valence is
$m_{k}(t) = v_k + \delta\,(t - \bar t)$, and the sentence is drawn positive /
neutral / negative with probabilities from a softmax in $m$. Positive sentences
append words from a positive bank, negative sentences from a negative bank; both
banks are chosen to exist in the Bing and AFINN lexicons so the measurement can
recover them.

```{r}
#| label: dgp-valence

# Valence words chosen to be present in the Bing lexicon (and mostly AFINN).
valence_words <- list(
  positive = c("support","benefit","strong","proud","secure","fair",
               "success","gain"),
  negative = c("fail","failed","fear","weak","corrupt","threat",
               "attack","crisis")
)

# Baseline valence per topic (order matches topic_names).
topic_valence   <- c(-0.20, 0.10, -0.50, 0.30, -0.60, 0.40)
val_trend_slope <- -0.10   # tone sours ~0.10 logits per year (centered)

sentence_valence_prob <- function(k, t) {
  m   <- topic_valence[k] + val_trend_slope * (t - mean(t_idx))
  a   <- 1.2
  raw <- c(negative = exp(-a * m), neutral = exp(0.2), positive = exp(a * m))
  raw / sum(raw)
}

make_sentence <- function(k, t) {
  ws   <- sample(topic_vocab[[k]], 2)
  base <- glue(sample(templates, 1), w1 = ws[1], w2 = ws[2])
  vp   <- sentence_valence_prob(k, t)
  val  <- sample(c("negative", "neutral", "positive"), 1, prob = vp)
  if (val == "neutral") return(as.character(base))
  vw <- sample(valence_words[[val]], sample(1:2, 1))
  paste(base, paste(vw, collapse = " "))
}

# Planted net valence (P(positive) - P(negative)) by topic and year = ground truth.
valence_truth <- expand_grid(topic_i = seq_len(dgp$K), t = t_idx) |>
  mutate(
    probs = map2(topic_i, t, sentence_valence_prob),
    net   = map_dbl(probs, \(p) p[["positive"]] - p[["negative"]]),
    topic = topic_names[topic_i],
    year  = t + min(dgp$years)
  )
```

```{r}
#| label: fig-valence-truth
#| fig-cap: "Planted net valence (P(positive) - P(negative)) by topic and year."

valence_truth |>
  mutate(topic = factor(topic, levels = topic_names)) |>
  ggplot(aes(year, net, color = topic)) +
  geom_line(linewidth = 0.9) +
  scale_color_viridis_d(end = 0.92) +
  scale_x_continuous(breaks = dgp$years) +
  geom_hline(yintercept = 0, linewidth = 0.3, linetype = "dotted") +
  labs(x = NULL, y = "Net valence (truth)", color = NULL,
       title = "Ground-truth sentiment by topic over time") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Generating the population corpus

Each post's topic mixture $\theta_d$ perturbs the year's prevalence vector on the
log scale; each sentence draws a topic from $\theta_d$ and a valence from the
topic-and-year valence probabilities. Construction is vectorized with `purrr`
(no `for`/`while` loops).

```{r}
#| label: dgp-generate

draw_theta <- function(t) {
  p   <- pi_mat[t + 1L, ]
  raw <- p * exp(rnorm(dgp$K, 0, dgp$theta_noise_sd))
  raw / sum(raw)
}

generate_post <- function(theta, t) {
  n_sent    <- sample(dgp$n_sent_range, 1)
  topic_seq <- sample(seq_len(dgp$K), n_sent, replace = TRUE, prob = theta)
  map_chr(topic_seq, \(k) make_sentence(k, t)) |> paste(collapse = " ")
}

set.seed(613)
corpus_pop <- map_dfr(
  dgp$years,
  \(yr) tibble(year = yr, t = yr - min(dgp$years))
) |>
  group_by(year) |>
  slice(rep(1, dgp$posts_per_year)) |>
  ungroup() |>
  mutate(
    doc_id   = sprintf("post_%05d", row_number()),
    theta    = map(t, draw_theta),
    dominant = map_int(theta, which.max),
    dominant = factor(topic_names[dominant], levels = topic_names),
    text     = map2_chr(theta, t, generate_post)
  )

glue("Population corpus: {nrow(corpus_pop)} posts across ",
     "{min(dgp$years)}-{max(dgp$years)}.")
```

## Imposing the selection mechanism (turning a census into a sample)

The **true** inclusion probability of post $d$ is

$$
\pi_d = \mathrm{logit}^{-1}\!\big(\beta_0 + \beta_{\text{year}}\,\tilde t_d
        + \beta_{\text{contro}}\,c_d\big),
$$

with $\tilde t_d$ the centered year (recent posts survive more often) and $c_d$
the post's share of attention on "controversial" topics (Immigration +
Security), which the author deletes more often. Because
$\beta_{\text{contro}} \neq 0$, capture depends on the *content* being measured:
a **missing-not-at-random (MNAR)** mechanism.

```{r}
#| label: dgp-selection

sel <- list(b0 = 0.4, b_year = 0.16, b_contro = -1.4)
idx_contro <- which(topic_names %in%
                      c("Immigration & Borders", "National Security & Defense"))

corpus_pop <- corpus_pop |>
  mutate(
    contro_share = map_dbl(theta, \(th) sum(th[idx_contro])),
    year_c       = t - mean(t),
    pi_true      = plogis(sel$b0 + sel$b_year * year_c +
                            sel$b_contro * contro_share)
  )

set.seed(617)
corpus_pop <- corpus_pop |> mutate(observed = rbinom(n(), 1, pi_true) == 1)
observed  <- corpus_pop |> filter(observed)

glue("Observed sample: {nrow(observed)} of {nrow(corpus_pop)} posts ",
     "({percent(mean(corpus_pop$observed), 0.1)} capture rate).")
```

```{r}
#| label: fig-capture
#| fig-cap: "Capture rate by year. Older and controversial posts are under-represented."

corpus_pop |>
  group_by(year) |>
  summarise(capture = mean(observed), .groups = "drop") |>
  ggplot(aes(year, capture)) +
  geom_col(fill = viridisLite::viridis(1, begin = 0.35)) +
  scale_x_continuous(breaks = dgp$years) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1)) +
  labs(x = NULL, y = "Capture rate",
       title = "Sample capture rate by year (selection mechanism)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Text preprocessing with tidytext

From here the code is **identical to what it would be on real data**: it operates
on `observed$text`, never on the hidden truth columns. Tokenization uses
one-token-per-row form, drops stop words via an anti-join, and removes numeric
tokens, following the tidytext book.

```{r}
#| label: tokenize

tokens <- observed |>
  dplyr::select(doc_id, year, text) |>
  unnest_tokens(word, text) |>
  anti_join(get_stopwords(), by = "word") |>
  filter(!str_detect(word, "\\d"))

glue("Vocabulary after cleaning: {n_distinct(tokens$word)} terms; ",
     "{nrow(tokens)} tokens.")
```

::: {.callout-warning}
## Processing-error checkpoint
Stop-word lists, stemming, and minimum-frequency trimming are choices, not
neutral defaults. Record them; report a sensitivity check in which headline
results are re-estimated under at least one alternative recipe.
:::

# Exploratory text analysis: distinctive terms over time

Treating each *year* as a document, tf-idf surfaces the terms that distinguish
one year's commentary from the decade as a whole (book Ch. 3).

```{r}
#| label: fig-tfidf
#| fig-cap: "Top tf-idf terms by year (model-free first look at emphasis drift)."
#| fig-height: 6

tokens |>
  count(year, word) |>
  bind_tf_idf(word, year, n) |>
  group_by(year) |>
  slice_max(tf_idf, n = 6, with_ties = FALSE) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, year)) |>
  ggplot(aes(tf_idf, word, fill = factor(year))) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d() +
  scale_y_reordered() +
  facet_wrap(~ year, scales = "free_y", ncol = 3) +
  labs(x = "tf-idf", y = NULL, title = "Year-distinctive terms") +
  theme(axis.text.y = element_text(size = 6))
```

# Topic model 1: Latent Dirichlet Allocation (book Ch. 6)

LDA's generative story is

$$
\theta_d \sim \mathrm{Dir}(\alpha), \quad
z_{dn} \sim \mathrm{Cat}(\theta_d), \quad
w_{dn} \sim \mathrm{Cat}(\beta_{z_{dn}}),
$$

with $\theta_d$ the document-topic mixture (gamma) and $\beta_k$ the topic-word
distribution (beta). **LDA has no notion of time**; we estimate per-document
topic shares and aggregate by year afterward.

## Choosing the number of topics

$K$ is the key tuning parameter. The documented procedure is a held-out
perplexity sweep plus interpretability review; it is shown but not executed in
the render. We proceed with $K = 6$ because the simulation planted six topics; on
real data, **choose $K$ via this procedure, not by assumption.**

```{r}
#| label: lda-k-sweep
#| eval: false

dtm_all <- observed |> unnest_tokens(word, text) |>
  anti_join(get_stopwords(), by = "word") |>
  filter(!str_detect(word, "\\d")) |>
  count(doc_id, word) |>
  cast_dtm(doc_id, word, n)

map_dfr(4:9, function(k) {
  fit <- LDA(dtm_all, k = k, method = "Gibbs",
             control = list(seed = 1, iter = 800, burnin = 200))
  tibble(k = k, perplexity = perplexity(fit, dtm_all))
}) |>
  ggplot(aes(k, perplexity)) + geom_line() + geom_point() +
  labs(title = "Held-out perplexity by K (lower = better)")
```

## Fitting LDA and extracting topics

```{r}
#| label: lda-fit

word_counts <- tokens |> count(doc_id, word)
dtm <- word_counts |> cast_dtm(doc_id, word, n)

lda_fit <- LDA(
  dtm, k = dgp$K, method = "Gibbs",
  control = list(seed = 1234, iter = 1500, burnin = 300, alpha = 0.1)
)

lda_beta  <- tidy(lda_fit, matrix = "beta")
lda_gamma <- tidy(lda_fit, matrix = "gamma")
```

```{r}
#| label: fig-lda-terms
#| fig-cap: "Top terms per LDA topic (numbered arbitrarily until aligned in 7.3)."

lda_beta |>
  group_by(topic) |>
  slice_max(beta, n = 8, with_ties = FALSE) |>
  ungroup() |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d() +
  scale_y_reordered() +
  facet_wrap(~ topic, scales = "free_y") +
  labs(x = expression(beta), y = NULL, title = "LDA topic-term distributions")
```

## Aligning estimated topics to ground truth (Hungarian algorithm)

Topic models return topics in arbitrary order. We align recovered topics to the
planted ones by solving a linear assignment problem (Hungarian algorithm) on a
term-overlap cost matrix.

```{r}
#| label: align-lda

vocab_lookup <- imap_dfr(topic_vocab, \(words, nm) tibble(term = words, planted = nm))

align_topics <- function(top_terms_df) {
  ov <- top_terms_df |>
    left_join(vocab_lookup, by = "term") |>
    filter(!is.na(planted)) |>
    count(topic, planted) |>
    pivot_wider(names_from = planted, values_from = n, values_fill = 0) |>
    column_to_rownames("topic")
  miss <- setdiff(topic_names, colnames(ov))
  ov[miss] <- 0
  ov <- ov[, topic_names, drop = FALSE]
  a  <- solve_LSAP(max(as.matrix(ov)) - as.matrix(ov))
  tibble(topic = as.integer(rownames(ov)),
         planted_name = colnames(ov)[as.integer(a)])
}

lda_topic_map <- lda_beta |>
  group_by(topic) |>
  slice_max(beta, n = 10, with_ties = FALSE) |>
  ungroup() |>
  rename(term = term) |>
  align_topics()
lda_topic_map
```

## LDA-based prevalence over time and validation

```{r}
#| label: fig-lda-trend
#| fig-cap: "LDA-recovered prevalence (lines) vs. ground truth (dashed)."

lda_prev <- lda_gamma |>
  rename(doc_id = document) |>
  mutate(topic = as.integer(topic)) |>
  left_join(lda_topic_map, by = "topic") |>
  left_join(dplyr::select(observed, doc_id, year), by = "doc_id") |>
  group_by(year, planted_name) |>
  summarise(est_prev = mean(gamma), .groups = "drop")

lda_prev |>
  ggplot(aes(year, est_prev, color = planted_name)) +
  geom_line(linewidth = 0.9) +
  geom_line(data = truth_prevalence |> rename(planted_name = topic),
            aes(year, true_prev), linetype = "dashed", linewidth = 0.5) +
  scale_color_viridis_d(end = 0.92) +
  scale_x_continuous(breaks = dgp$years) +
  scale_y_continuous(labels = label_percent()) +
  facet_wrap(~ planted_name) +
  labs(x = NULL, y = "Share of attention",
       title = "LDA recovery vs. ground truth") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

```{r}
#| label: lda-recovery-corr

lda_prev |>
  left_join(truth_prevalence, by = c("year", "planted_name" = "topic")) |>
  group_by(planted_name) |>
  summarise(correlation = cor(est_prev, true_prev), .groups = "drop")
```

# Topic model 2: Structural Topic Model (time-aware, recommended)

LDA's post-hoc aggregation has no principled standard error on a trend. The
**Structural Topic Model** lets topic prevalence depend on covariates; here,
smoothly on year:

$$
\theta_d \sim \mathrm{LogisticNormal}\big(\Gamma^\top x_d,\ \Sigma\big),
\qquad x_d = \big[1,\ s(\text{year}_d)\big].
$$

This is the cleaner tool for "how do topics evolve over time," estimating the
time effect jointly with the topics. STM extends beyond the SURV727 core toolkit.

```{r}
#| label: stm-fit
#| results: hide

processed <- textProcessor(
  documents = observed$text,
  metadata  = observed |> dplyr::select(doc_id, year),
  lowercase = TRUE, removestopwords = TRUE, removenumbers = TRUE,
  removepunctuation = TRUE, stem = FALSE, verbose = FALSE
)
prepped <- prepDocuments(processed$documents, processed$vocab,
                         processed$meta, lower.thresh = 5, verbose = FALSE)

stm_fit <- stm(
  documents = prepped$documents, vocab = prepped$vocab, K = dgp$K,
  prevalence = ~ s(year), data = prepped$meta,
  init.type = "Spectral", seed = 740, verbose = FALSE
)
```

```{r}
#| label: align-stm

stm_terms <- labelTopics(stm_fit, n = 8)
stm_top <- imap_dfr(seq_len(dgp$K),
                    \(k, .) tibble(topic = k, term = stm_terms$frex[k, ]))
stm_topic_map <- align_topics(stm_top)
stm_topic_map
```

```{r}
#| label: stm-effect
#| results: hide

# Formal model-based test of the time effect (with uncertainty).
stm_effect <- estimateEffect(1:dgp$K ~ s(year), stm_fit,
                             metadata = prepped$meta, uncertainty = "Global")
```

```{r}
#| label: fig-stm-trend
#| fig-cap: "STM-recovered prevalence (theta averaged by year) vs. ground truth (dashed)."

stm_gamma <- stm_fit$theta |>
  as_tibble(.name_repair = ~ as.character(seq_len(dgp$K))) |>
  mutate(doc_id = prepped$meta$doc_id, year = prepped$meta$year) |>
  pivot_longer(c(-doc_id, -year), names_to = "topic", values_to = "gamma") |>
  mutate(topic = as.integer(topic)) |>
  left_join(stm_topic_map, by = "topic")

stm_prev <- stm_gamma |>
  group_by(year, planted_name) |>
  summarise(est_prev = mean(gamma), .groups = "drop")

stm_prev |>
  ggplot(aes(year, est_prev, color = planted_name)) +
  geom_line(linewidth = 0.9) +
  geom_line(data = truth_prevalence |> rename(planted_name = topic),
            aes(year, true_prev), linetype = "dashed", linewidth = 0.5) +
  scale_color_viridis_d(end = 0.92) +
  scale_x_continuous(breaks = dgp$years) +
  scale_y_continuous(labels = label_percent()) +
  facet_wrap(~ planted_name) +
  labs(x = NULL, y = "Share of attention",
       title = "STM recovery vs. ground truth") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

::: {.callout-note}
## Plotting the formal effect
`plot(stm_effect, covariate = "year", method = "continuous", topics = 1:6, model = stm_fit)`
draws the modeled prevalence with confidence bands. The figure above averages
`theta` by year so LDA, STM, and the weighted analysis (Section 10) share one
comparable estimator; report the `estimateEffect` bands when you need inference
on the trend itself.
:::

Each document's dominant STM topic (used to condition sentiment in Section 8):

```{r}
#| label: stm-dominant

doc_dominant <- stm_gamma |>
  group_by(doc_id, year) |>
  slice_max(gamma, n = 1, with_ties = FALSE) |>
  ungroup() |>
  dplyr::select(doc_id, year, dominant_topic = planted_name)
```

# Sentiment analysis (book Ch. 2 and Ch. 4)

Sentiment is treated, in TSE terms, as a **measurement** of latent tone. We start
with the reliable, fully reproducible lexicon method, then quantify its known
failure mode (negation), then (Section 9) cross-check against a local LLM.

The natural unit here is the **document (a paragraph)**, which the book notes
works better than very large chunks. The Bing lexicon ships with tidytext and
needs no license; AFINN/NRC require `textdata` plus a one-time acceptance, so
AFINN-dependent chunks below are *gated* and skip gracefully if unavailable.

```{r}
#| label: sentiment-bing

bing <- tryCatch(get_sentiments("bing"), error = function(e) NULL)
if (is.null(bing)) {
  # Fallback so the section always renders: a minimal lexicon from the valence
  # banks (on real data you will have the full Bing lexicon installed).
  bing <- bind_rows(
    tibble(word = valence_words$positive, sentiment = "positive"),
    tibble(word = valence_words$negative, sentiment = "negative")
  )
  message("Bing unavailable; using a minimal fallback lexicon (valence words only).")
}

# Per-document net sentiment (positive - negative word counts), Bing.
doc_sent_bing <- tokens |>
  inner_join(bing, by = "word") |>
  count(doc_id, year, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(net_bing = positive - negative)

# Re-attach documents with zero sentiment words (so means are not biased).
doc_sent_bing <- observed |>
  dplyr::select(doc_id, year) |>
  left_join(doc_sent_bing, by = c("doc_id", "year")) |>
  mutate(across(c(positive, negative, net_bing), \(x) coalesce(x, 0L)))
```

```{r}
#| label: fig-sent-time
#| fig-cap: "Mean per-document Bing net sentiment by year (the planted souring trend)."

doc_sent_bing |>
  group_by(year) |>
  summarise(mean_net = mean(net_bing), .groups = "drop") |>
  ggplot(aes(year, mean_net)) +
  geom_line(linewidth = 0.9, color = viridisLite::viridis(1, begin = 0.3)) +
  geom_point(size = 1.4) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = 0.3) +
  scale_x_continuous(breaks = dgp$years) +
  labs(x = NULL, y = "Mean net sentiment (Bing)",
       title = "Overall sentiment over time") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Sentiment by topic over time, validated against truth

Joining each document's net sentiment to its dominant STM topic gives sentiment
trajectories per policy area, which we validate against the planted net valence.

```{r}
#| label: fig-sent-topic
#| fig-cap: "Estimated sentiment by topic over time (lines) and standardized planted valence (dashed)."

sent_topic <- doc_sent_bing |>
  left_join(doc_dominant, by = c("doc_id", "year")) |>
  filter(!is.na(dominant_topic)) |>
  group_by(year, dominant_topic) |>
  summarise(mean_net = mean(net_bing), .groups = "drop")

# Put estimate and truth on a common (z-scored) scale for shape comparison.
z <- function(x) (x - mean(x)) / sd(x)
truth_z <- valence_truth |> group_by(topic) |> mutate(net_z = z(net)) |> ungroup()
est_z   <- sent_topic |> group_by(dominant_topic) |> mutate(net_z = z(mean_net)) |> ungroup()

ggplot(est_z, aes(year, net_z, color = dominant_topic)) +
  geom_line(linewidth = 0.9) +
  geom_line(data = truth_z |> rename(dominant_topic = topic),
            aes(year, net_z), linetype = "dashed", linewidth = 0.5) +
  scale_color_viridis_d(end = 0.92) +
  scale_x_continuous(breaks = dgp$years) +
  facet_wrap(~ dominant_topic) +
  labs(x = NULL, y = "Net sentiment (z-scored)",
       title = "Sentiment by topic: estimate vs. planted truth") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

```{r}
#| label: sent-recovery-corr

sent_topic |>
  left_join(valence_truth, by = c("year", "dominant_topic" = "topic")) |>
  group_by(dominant_topic) |>
  summarise(correlation = cor(mean_net, net), .groups = "drop")
```

## Negation diagnostic (book Ch. 4): how much does unigram sentiment misread?

Unigram lexicons score "good" and "support" as positive even in "not good" or
"never support." Tokenizing into bigrams and finding sentiment words preceded by
a negator quantifies the exposure. This uses AFINN (numeric), so it is gated.

```{r}
#| label: afinn-available
#| include: false

afinn <- tryCatch(get_sentiments("afinn"), error = function(e) NULL)
have_afinn <- !is.null(afinn)
```

```{r}
#| label: negation-diagnostic
#| eval: !expr have_afinn
#| fig-cap: "Sentiment words following a negator, weighted by AFINN value x frequency."

negators <- c("not", "no", "never", "without")

negated <- observed |>
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(word1 %in% negators) |>
  inner_join(afinn, by = c("word2" = "word")) |>
  count(word1, word2, value, sort = TRUE) |>
  mutate(contribution = n * value)

if (nrow(negated) == 0) {
  message("No negated sentiment words found (expected: this synthetic corpus contains little negation by construction).")
} else {
  negated |>
    slice_max(abs(contribution), n = 20) |>
    mutate(word2 = reorder(word2, contribution)) |>
    ggplot(aes(contribution, word2, fill = contribution > 0)) +
    geom_col(show.legend = FALSE) +
    scale_fill_viridis_d(end = 0.7) +
    labs(x = "AFINN value x occurrences", y = "Word preceded by a negator",
         title = "Sentiment words that may be misread due to negation")
}
```

```{r}
#| label: negation-note
#| eval: !expr (!have_afinn)
#| echo: false
#| results: asis

cat("> **AFINN not available.** Install `textdata` and run `textdata::lexicon_afinn()` once interactively to accept the license, then re-render to enable the negation diagnostic.\n")
```

::: {.callout-warning}
## Lexicon caveats (stated by the book)
Lexicon sentiment is **unigram-based** and ignores qualifiers ("no good", "not
true"); it was **validated on specific text types** (reviews, tweets), so applying
it to political prose risks domain mismatch; and **chunk size matters** (sentence-
or paragraph-sized text works better than many-paragraph blocks). Real political
commentary is dense with negation, so run the diagnostic above and consider a
negation-corrected score (flip the sign of negated words) or a context-aware tool
(`sentimentr`) or the LLM route below. On this synthetic corpus negation is rare
by construction, so the diagnostic is shown mainly as the procedure to run on
real data.
:::

# Optional: AI-assisted module with a local model (Gemma via Ollama)

You can run a local LLM through ellmer's Ollama backend, so no data leaves your
machine and there is no API cost. Two auditable uses follow: topic labeling and
sentiment, each cross-checked. All AI chunks are `eval: false` so the PDF renders
without Ollama running; flip to `eval: true` once `ollama serve` is up and the
model is pulled. Swapping to a hosted model is a one-line change
(`chat_anthropic()`, `chat_openai()`, ...).

```{r}
#| label: ai-setup
#| eval: false

library(ellmer)

# Confirm the exact local tag, then set it here (e.g. "gemma3", "gemma4", "llama4").
# ellmer::models_ollama()
ollama_model <- "gemma4"

# temperature 0 + fixed seed for reproducibility; local => private, free.
make_chat <- function(system_prompt) {
  chat_ollama(
    system_prompt = system_prompt,
    model  = ollama_model,
    params = params(temperature = 0, seed = 1)
  )
}
```

## Use 1 - LLM topic labeling

Turning each topic's top terms into a human-readable label is low risk: the model
sees only anonymized term lists, and a human reviews the output.

```{r}
#| label: ai-label
#| eval: false

labeler <- make_chat(paste(
  "You label political topic models. Given the top terms of one topic, return a",
  "concise 2-4 word policy label and a one-sentence description, based only on",
  "the terms provided."
))

label_type <- type_object(
  label       = type_string("A concise 2-4 word policy-area label."),
  description = type_string("One sentence describing the topic.")
)

ai_labels <- stm_top |>
  summarise(terms = paste(term, collapse = ", "), .by = topic) |>
  mutate(result = map(terms, \(tt)
    labeler$chat_structured(glue("Top terms: {tt}"), type = label_type))) |>
  unnest_wider(result)
ai_labels
```

## Use 2 - LLM sentiment, cross-checked against the lexicon

Classify each post's sentiment with the local model, then compare to the Bing
score: Spearman correlation on the continuous signal and Cohen's kappa on the
sign. Agreement is convergent-validity evidence; disagreement flags uncertain
measurements. **Process one post per call** — Ollama's default input cap is 2048
tokens, which a paragraph fits easily but a batch would not.

```{r}
#| label: ai-sentiment
#| eval: false

rater <- make_chat(paste(
  "You rate the sentiment of a political post toward the policy it discusses.",
  "Return a label (negative/neutral/positive) and a score from -1 (very",
  "negative) to 1 (very positive)."
))

sent_type <- type_object(
  label = type_enum(c("negative", "neutral", "positive"), "Overall sentiment."),
  score = type_number("Sentiment score from -1 to 1.")
)

set.seed(745)
ai_check <- observed |> slice_sample(n = 200)

# Small local models can occasionally return non-conforming output; guard each call.
safe_rate <- function(txt) {
  tryCatch(rater$chat_structured(txt, type = sent_type),
           error = function(e) list(label = NA_character_, score = NA_real_))
}

ai_sent <- ai_check |>
  mutate(res = map(text, safe_rate)) |>
  unnest_wider(res) |>
  left_join(doc_sent_bing, by = c("doc_id", "year"))

# Convergent validity
spearman <- cor(ai_sent$score, ai_sent$net_bing,
                method = "spearman", use = "complete.obs")
kappa <- irr::kappa2(
  ai_sent |>
    transmute(ai = label,
              lex = case_when(net_bing > 0 ~ "positive",
                              net_bing < 0 ~ "negative",
                              TRUE ~ "neutral")) |>
    drop_na()
)
list(spearman = spearman, kappa = kappa)
```

::: {.callout-note}
## Convenience alternative: the {mall} package
`mall` wraps ellmer for column-wise LLM operations, e.g. `mall::llm_sentiment()`
over a text column after wiring a backend with `mall::llm_use()` (check current
`mall` docs for the exact Ollama wiring). It is the fastest path to "AI sentiment
through an R package," but the explicit ellmer route above gives more control and
is what is specified here.
:::

::: {.callout-warning}
## AI validity and reproducibility caveats
LLM labels and scores are **measurements with error** in the TSE "measurement"
cell. Even at temperature 0, local models can vary across versions and quantizations
and are sensitive to prompt wording. Pin the model tag, log prompts and raw
responses, validate against the lexicon (above), and never treat an LLM label as
ground truth. Small local models also produce less reliable structured output than
hosted ones — hence the per-call guard.
:::

# Inference under sampling: correcting selection bias

Sections 7-8 estimated prevalence and sentiment *in the observed sample*, which
over-represents recent and non-controversial posts. This section makes the
correction explicit with design-based estimation — the survey-methods payoff.

## The estimator

With unequal inclusion probabilities $\pi_d$, a population mean (of topic share or
of sentiment) in year $t$ is the Horvitz-Thompson / weighted mean

$$
\hat{\bar y}_{t}
  = \frac{\sum_{d \in s_t} w_d\,y_{d}}{\sum_{d \in s_t} w_d},
  \qquad w_d = \frac{1}{\pi_d},
$$

and the variance penalty from unequal weighting is the Kish design effect

$$
\mathrm{deff}_{\text{Kish}} = 1 + \mathrm{CV}^2(w)
  = \frac{n \sum_d w_d^2}{\left(\sum_d w_d\right)^2}.
$$

```{r}
#| label: weights

# Analyst-estimable propensity: predict capture from year only (MAR assumption).
prop_model <- glm(observed ~ year_c, family = binomial, data = corpus_pop)
corpus_pop <- corpus_pop |>
  mutate(pi_hat_mar = predict(prop_model, type = "response"))

# Document-level analysis frame: STM topic shares (wide) + Bing sentiment + weights.
topic_cols <- stm_topic_map$planted_name

stm_wide <- stm_fit$theta |>
  as_tibble(.name_repair = ~ stm_topic_map$planted_name[
    match(seq_len(dgp$K), stm_topic_map$topic)]) |>
  mutate(doc_id = prepped$meta$doc_id)

analysis <- stm_wide |>
  left_join(dplyr::select(doc_sent_bing, doc_id, net_bing), by = "doc_id") |>
  left_join(dplyr::select(corpus_pop, doc_id, year, pi_true, pi_hat_mar),
            by = "doc_id") |>
  mutate(w_ipw = 1 / pi_true, w_mar = 1 / pi_hat_mar)

kish_deff <- with(analysis, length(w_ipw) * sum(w_ipw^2) / sum(w_ipw)^2)
glue("Kish design effect (correct weights): {round(kish_deff, 2)} ",
     "(effective n approx {round(nrow(analysis) / kish_deff)}).")
```

## Prevalence: naive vs. corrected

```{r}
#| label: fig-ipw-prev
#| fig-cap: "Naive sample means are biased; IPW with correct inclusion probabilities recovers population prevalence."

naive_prev <- analysis |>
  group_by(year) |>
  summarise(across(all_of(topic_cols), mean), .groups = "drop") |>
  pivot_longer(-year, names_to = "topic", values_to = "prev") |>
  mutate(estimator = "Naive (unweighted)")

ipw_prev <- analysis |>
  as_survey_design(ids = 1, weights = w_ipw) |>
  group_by(year) |>
  summarise(across(all_of(topic_cols), survey_mean)) |>
  dplyr::select(year, all_of(topic_cols)) |>
  pivot_longer(-year, names_to = "topic", values_to = "prev") |>
  mutate(estimator = "IPW (correct weights)")

truth_long <- truth_prevalence |> rename(prev = true_prev) |>
  mutate(estimator = "Population truth")

bind_rows(truth_long, naive_prev, ipw_prev) |>
  mutate(topic = factor(topic, levels = topic_names)) |>
  ggplot(aes(year, prev, color = estimator, linetype = estimator)) +
  geom_line(linewidth = 0.8) +
  scale_color_viridis_d(end = 0.85) +
  scale_x_continuous(breaks = dgp$years) +
  scale_y_continuous(labels = label_percent()) +
  facet_wrap(~ topic) +
  labs(x = NULL, y = "Share of attention", color = NULL, linetype = NULL,
       title = "Selection bias and its correction: prevalence") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

## Sentiment: naive vs. corrected

```{r}
#| label: sent-estimates

naive_sent <- analysis |>
  group_by(year) |>
  summarise(mean_net = mean(net_bing), .groups = "drop") |>
  mutate(estimator = "Naive (unweighted)")

ipw_sent <- analysis |>
  as_survey_design(ids = 1, weights = w_ipw) |>
  group_by(year) |>
  summarise(mean_net = survey_mean(net_bing)) |>
  dplyr::select(year, mean_net) |>
  mutate(estimator = "IPW (correct weights)")
```

```{r}
#| label: sent-truth-recompute
#| fig-cap: "Sentiment correction with a population-truth reference."

# Recompute Bing net sentiment on the FULL population (truth reference only).
pop_tokens <- corpus_pop |>
  dplyr::select(doc_id, year, text) |>
  unnest_tokens(word, text) |>
  anti_join(get_stopwords(), by = "word") |>
  filter(!str_detect(word, "\\d"))

pop_doc_sent <- corpus_pop |>
  dplyr::select(doc_id, year) |>
  left_join(
    pop_tokens |>
      inner_join(bing, by = "word") |>
      count(doc_id, sentiment) |>
      pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
      mutate(net_bing = positive - negative) |>
      dplyr::select(doc_id, net_bing),
    by = "doc_id"
  ) |>
  mutate(net_bing = coalesce(net_bing, 0L))

truth_sent_line <- pop_doc_sent |>
  group_by(year) |>
  summarise(mean_net = mean(net_bing), .groups = "drop") |>
  mutate(estimator = "Population truth")

bind_rows(truth_sent_line, naive_sent, ipw_sent) |>
  ggplot(aes(year, mean_net, color = estimator, linetype = estimator)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = 0.3) +
  scale_color_viridis_d(end = 0.85) +
  scale_x_continuous(breaks = dgp$years) +
  labs(x = NULL, y = "Mean net sentiment", color = NULL, linetype = NULL,
       title = "Selection bias and its correction: sentiment") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Sensitivity analysis: how much does the assumption matter?

The MNAR mechanism deletes controversial posts based on *content*, which an
analyst cannot observe directly. MAR-only weights (year alone) therefore cannot
fully correct the bias. The residual gap is the honest measure of how far
conclusions depend on an unverifiable assumption.

```{r}
#| label: fig-sensitivity
#| fig-cap: "MAR-only weighting (year alone) leaves residual bias for content-dependent topics."

mar_prev <- analysis |>
  as_survey_design(ids = 1, weights = w_mar) |>
  group_by(year) |>
  summarise(across(all_of(topic_cols), survey_mean)) |>
  dplyr::select(year, all_of(topic_cols)) |>
  pivot_longer(-year, names_to = "topic", values_to = "prev") |>
  mutate(estimator = "IPW (MAR / year only)")

bind_rows(truth_long,
          mar_prev,
          ipw_prev |> mutate(estimator = "IPW (correct)")) |>
  filter(topic %in% topic_names[idx_contro]) |>
  mutate(topic = factor(topic, levels = topic_names)) |>
  ggplot(aes(year, prev, color = estimator, linetype = estimator)) +
  geom_line(linewidth = 0.9) +
  scale_color_viridis_d(end = 0.85) +
  scale_x_continuous(breaks = dgp$years) +
  scale_y_continuous(labels = label_percent()) +
  facet_wrap(~ topic) +
  labs(x = NULL, y = "Share of attention", color = NULL, linetype = NULL,
       title = "Sensitivity to the missingness assumption (controversial topics)") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

# Adapting this workflow to the real corpus

1. **Replace Section 4** with a read of your scraped corpus: a tibble of
   `doc_id`, `date`, `text`; derive `year`. Delete every `theta`, `dominant`,
   `pi_true`, `observed`, and `valence_*` column — those exist only because truth
   was known.
2. **Confront coverage directly.** You cannot observe $\pi_d$. The principled
   estimate of *how many posts are missing* is capture-recapture (dual-system
   estimation): collect from two independent sources (live scrape + archive) and
   estimate the unobserved total from the overlap. Document the frame and its gaps.
3. **Estimate inclusion propensities** from observable predictors (post age,
   source, length, engagement); this yields the MAR estimator, not the correct
   one. State the residual-MNAR risk; the sensitivity analysis is your honest
   uncertainty band.
4. **Choose $K$** with the documented perplexity / `searchK` procedure plus
   interpretability.
5. **For sentiment**, run the negation diagnostic, prefer paragraph-sized units,
   consider a context-aware method, and validate any LLM scores against a lexicon
   baseline as in Section 9.
6. **Keep the validation mindset.** Without ground truth, validation shifts to
   topic stability across seeds, held-out coherence, human coding of a sample,
   and lexicon-vs-LLM agreement.

# Limitations and threats to validity

- **Frame / coverage.** Capturable posts are unknown; deletions and platform gaps
  are not random. Capture-recapture bounds this; nothing eliminates it.
- **MNAR selection.** If capture depends on content, no weighting on observed
  covariates fully corrects the bias. Conclusions are conditional on the
  missingness model; report the sensitivity range.
- **Measurement: topic instability.** Topics shift with $K$, seed, preprocessing,
  and trimming. Report robustness.
- **Measurement: temporal non-invariance.** LDA and STM assume a *time-invariant*
  topic-word distribution $\beta$; over a decade, framing drifts, so a fixed
  $\beta$ can mis-measure later years. Dynamic Topic Models (Blei & Lafferty 2006)
  relax this; treat decade-long trends as provisional until checked.
- **Measurement: sentiment.** Unigram lexicons miss negation and domain shift; LLM
  scores are non-deterministic and version-dependent. Triangulate.
- **Construct validity.** "Topics" are word-co-occurrence patterns and "sentiment"
  is lexical tone, neither of which is the politician's intent; a rising topic
  share is attention, not endorsement.
- **Simulation gap.** Real topics overlap and real negation is common, so
  real-world recovery will be noisier than shown here.

# References

::: {.callout-important}
## Verify before circulating
Per standing practice, confirm every citation against its primary source. The
list below is from memory and is **not** independently verified in this render.
:::

- Silge, J., & Robinson, D. (2017). *Text Mining with R: A Tidy Approach.*
  O'Reilly. <https://www.tidytextmining.com/>
- Blei, D. M., Ng, A. Y., & Jordan, M. I. (2003). Latent Dirichlet Allocation.
  *Journal of Machine Learning Research*, 3, 993-1022.
- Blei, D. M., & Lafferty, J. D. (2006). Dynamic Topic Models. *ICML*.
- Roberts, M. E., Stewart, B. M., & Tingley, D. (2019). stm: An R Package for
  Structural Topic Models. *Journal of Statistical Software*, 91(2).
- Roberts, M. E., et al. (2014). Structural Topic Models for Open-Ended Survey
  Responses. *American Journal of Political Science*, 58(4).
- Grimmer, J., & Stewart, B. M. (2013). Text as Data. *Political Analysis*, 21(3).
- Grimmer, J., Roberts, M. E., & Stewart, B. M. (2022). *Text as Data.* Princeton.
- Salganik, M. J. (2018). *Bit by Bit: Social Research in the Digital Age.*
  Princeton University Press.
- Japec, L., et al. (2015). Big Data in Survey Research: AAPOR Task Force Report.
  *Public Opinion Quarterly*, 79(4).
- Sen, I., et al. (2021). A Total Error Framework for Digital Traces of Human
  Behavior on Online Platforms. *Public Opinion Quarterly*, 85(S1). *(verify)*
- Gilardi, F., Alizadeh, M., & Kubli, M. (2023). ChatGPT Outperforms Crowd Workers
  for Text-Annotation Tasks. *PNAS*, 120(30).
- Kish, L. (1965). *Survey Sampling.* Wiley.
- Lohr, S. L. (2022). *Sampling: Design and Analysis* (3rd ed.). CRC Press.
- Heeringa, S. G., West, B. T., & Berglund, P. A. (2017). *Applied Survey Data
  Analysis* (2nd ed.). CRC Press.

```{r}
#| label: session
#| include: false
sessionInfo()
```
