---
title: "Public comments: paragraph corpus and structural topic model"
subtitle: "Script 1 of 2: text mining, STM, modal topic assignment"
format:
  typst:
    toc: false
    fig-width: 9
    fig-height: 5.5
    fig-format: png
    fig-dpi: 200
execute:
  warning: false
  message: false
---

```{=typst}
#show figure: set block(breakable: true)
```
```{=html}
<!--
PRODUCTION CHECKLIST (simulation -> real data)
  1. Delete the SIMULATED CORPUS section and replace it with one line:
       corpus_raw <- readr::read_csv(here(tx$proj, "input", "YOUR_FILE.csv"))
     Create text_mining/input/ and drop the CSV there; the file name is the
     only blank. Required columns: atom_id (comment id), Year (integer),
     body_english. Any other source is fine too: the validation gate below
     runs on whatever filled corpus_raw.
  2. Read the newline diagnostic; set tx$para_split if paragraphs are
     blank-line separated. Delete outputs/ ONCE (caches track files, not code).
  3. Render with tx$run_searchK = TRUE (hours), read the diagnostics, set
     tx$K, set tx$run_searchK = FALSE, re-render.
  4. Script 2 reads outputs/. Nothing else here changes between projects.
OUTPUTS: tabular -> .csv, model objects and lists -> .rds, all in outputs/.
-->
```

```{r config}
# The only block you edit. Every analyst decision is a line here.
tx <- list(
  seed               = 2026L,
  proj               = "text_mining",  # project folder under the here() root;
                                       # the one string to change when reusing
                                       # this workflow on another corpus
  para_split         = "\\r?\\n+",  # blank-line separated? "\\r?\\n{2,}"
  para_min_tokens    = 15L,         # body fragments below this merge BACKWARD
  heading_max_tokens = 12L,         # headings merge FORWARD (see atomization)
  min_docfreq        = 5L,          # DTM vocabulary floor
  run_searchK        = TRUE,       # TRUE for the long K-selection run
  K_grid             = seq(10, 40, 3),
  K                  = 25L,         # set from the searchK diagnostics
  spline_df          = 5L,
  facets_per_page    = 9L,          # topic trend panels per page
  workers            = 6L
)

suppressPackageStartupMessages({
  library(here); library(stm); library(quanteda); library(tidytext)
  library(splines)
  library(furrr); library(viridis)
  library(tidyverse)   # tidyverse LAST so its verbs win
})
theme_set(theme_minimal(base_size = 12))
# Tables must not silently lose rows in the rendered PDF: knitr truncates long
# tables by default, and a 25-topic summary is past that limit.
options(knitr.table.max.print = Inf, knitr.kable.max_rows = Inf,
        max.print = 1e6)
set.seed(tx$seed)
# All paths are anchored with here(), so chunks run the same whether you
# render from the project root or run them interactively from code/.
# Layout: <here()>/text_mining/{code, input, outputs}.
p_out <- function(f) here::here(tx$proj, "outputs", f)
dir.create(here::here(tx$proj, "outputs"), showWarnings = FALSE,
           recursive = TRUE)

# Checkpoints: compute once, reload after. Delete the file to re-fit.
cache_rds  <- function(path, expr) {
  if (file.exists(path)) return(readRDS(path))
  x <- force(expr); saveRDS(x, path); x
}
cache_csv <- function(path, expr) {
  if (file.exists(path)) return(readr::read_csv(path, show_col_types = FALSE))
  x <- force(expr); readr::write_csv(x, path); x
}
```

# SIMULATED CORPUS (delete this section at work)

Plants 25 topics with distinct vocabularies and per-topic year trajectories, so the pipeline can be scored on recovery of structure it was not shown. Script 2 reuses the planted labels to validate its estimators.

```{r simulate}
sim <- local({
  K <- 25L; years <- 2000:2025
  vocab <- purrr::map(seq_len(K), function(k)
    str_c("t", k, "term", seq_len(12)))
  filler <- str_c("common", seq_len(40))
  # topic prevalence drifts linearly in year, each topic with its own slope
  slope <- seq(-1, 1, length.out = K)
  make_comment <- function(i) {
    yr  <- sample(years, 1)
    pr  <- exp(slope * (yr - 2012) / 12); pr <- pr / sum(pr)
    n_p <- sample(4:12, 1)
    ks  <- sample.int(K, n_p, replace = TRUE, prob = pr)
    body <- purrr::map_chr(ks, function(k) str_c(
      str_c(sample(vocab[[k]], 14, replace = TRUE), collapse = " "), " ",
      str_c(sample(filler, 10, replace = TRUE), collapse = " ")))
    tibble(atom_id = sprintf("A%05d", i), Year = yr,
           body_english = str_c(body, collapse = "\n"),
           true_topics = str_c(ks, collapse = ","))
  }
  purrr::map(seq_len(1040), make_comment) |> purrr::list_rbind()
})
corpus_raw <- sim |> dplyr::select(atom_id, Year, body_english)
readr::write_csv(sim |> dplyr::select(atom_id, true_topics),
                 p_out("sim_truth_comments.csv"))
```

# Corpus validation (keep this at work)

Stops the render with a named error on a bad read, which is cheaper than debugging a silent failure five chunks later.

```{r validate}
miss <- setdiff(c("atom_id", "Year", "body_english"), names(corpus_raw))
if (length(miss)) stop("corpus_raw missing column(s): ",
                       str_c(miss, collapse = ", "))
stopifnot(
  "atom_id must be unique"     = !anyDuplicated(corpus_raw$atom_id),
  "Year must be integer-valued" = all(corpus_raw$Year ==
                                        as.integer(corpus_raw$Year)),
  "body_english must be character" = is.character(corpus_raw$body_english))
n0 <- nrow(corpus_raw)
corpus_raw <- corpus_raw |>
  dplyr::filter(!is.na(Year), !is.na(body_english),
                str_squish(body_english) != "")
glue::glue("Validation: kept {nrow(corpus_raw)} of {n0} comments, ",
           "{min(corpus_raw$Year)}-{max(corpus_raw$Year)}")
```

# Newline diagnostic

The atomizer can only split on what the source uses to mark paragraphs, and Excel-origin text (in-cell breaks survive as `\n`) carries three conventions needing three responses. Rules: mostly line breaks present, keep the default; blank lines dominant **and** the visible sample shows single breaks used as soft wraps, set `tx$para_split = "\\r?\\n{2,}"`; almost no newlines, stop, since sentence chunking changes the measurement unit and is a design decision.

```{r newline-diagnostic}
corpus_raw |>
  dplyr::mutate(b = str_replace_all(body_english, "\r\n", "\n")) |>
  dplyr::summarise(comments = dplyr::n(),
                   zero_newlines   = sum(str_count(b, "\n") == 0),
                   has_blank_lines = sum(str_count(b, "\n{2,}") > 0),
                   median_breaks   = median(str_count(b, "\n"))) |>
  knitr::kable(caption = "Line-break conventions; apply the decision rules above before the long render.")

corpus_raw |> dplyr::slice_sample(n = 2) |> dplyr::pull(body_english) |>
  str_replace_all("\r?\n", "<NL>\n") |>
  str_c(collapse = "\n\n==== NEXT ====\n\n") |> writeLines()
```

# Paragraph atomization

The unit of analysis is the paragraph: a comment spans several topics, and a topic model's "document" is whatever unit the DTM rows are. Merging is **direction-aware**, because short blocks come in two kinds that belong in opposite directions. Headings (at most `r tx$heading_max_tokens` tokens, no sentence-ending punctuation) merge **forward** into the paragraph they introduce, since a naive rule would glue a subtitle's words onto the *preceding* paragraph and misfile its topic signal; a trailing heading falls back to the last body paragraph. Short body fragments that do end like sentences merge **backward** into what they continue, absorbing codas, salutations, and signature lines.

```{r atomize}
count_tokens <- function(x) str_count(x, "\\S+")

paragraphs <- cache_csv(p_out("paragraphs.csv"), {
  # PASS 1 (forward): rev(cumsum(rev(!heading))) gives each body block an id
  # that decreases down the document and gives a heading the id of the NEXT
  # body block; trailing headings get 0, folded in by pmax(1L).
  corpus_raw |>
    dplyr::mutate(para_raw = str_split(body_english, tx$para_split)) |>
    dplyr::select(atom_id, Year, para_raw) |>
    tidyr::unnest(para_raw) |>
    dplyr::mutate(para_raw = str_squish(para_raw)) |>
    dplyr::filter(para_raw != "") |>
    dplyr::group_by(atom_id, Year) |>
    dplyr::mutate(
      heading = count_tokens(para_raw) <= tx$heading_max_tokens &
                !str_detect(para_raw, "[.!?]['\")\\]]*$"),
      fwd     = rev(cumsum(rev(!heading))) |> pmax(1L)) |>
    dplyr::group_by(atom_id, Year, fwd) |>
    dplyr::summarise(para_raw = str_c(para_raw, collapse = " "),
                     .groups = "drop") |>
    dplyr::arrange(atom_id, Year, dplyr::desc(fwd)) |>
    # PASS 2 (backward): short body fragments join the preceding paragraph.
    dplyr::group_by(atom_id, Year) |>
    dplyr::mutate(full = count_tokens(para_raw) >= tx$para_min_tokens,
                  blk  = cummax(cumsum(full) * full) |> pmax(1L)) |>
    dplyr::group_by(atom_id, Year, blk) |>
    dplyr::summarise(text = str_c(para_raw, collapse = " "),
                     .groups = "drop_last") |>
    dplyr::mutate(para_id  = dplyr::row_number(),
                  para_uid = str_c(atom_id, "-", sprintf("%02d", para_id)),
                  n_tokens = count_tokens(text)) |>
    dplyr::ungroup() |>
    dplyr::filter(n_tokens >= tx$para_min_tokens) |>
    dplyr::select(para_uid, atom_id, para_id, Year, n_tokens, text)
})
glue::glue("{nrow(paragraphs)} paragraphs from ",
           "{dplyr::n_distinct(paragraphs$atom_id)} comments")
```

# Corpus description

```{r eda}
#| fig-cap: "Paragraph volume by year: the denominator for every time-indexed estimate. Thin years mean the prevalence spline is interpolating and single-year claims are weak."
paragraphs |>
  dplyr::count(Year) |>
  ggplot(aes(Year, n)) +
  geom_col(fill = viridis(1, begin = 0.35)) +
  labs(x = NULL, y = "paragraphs")

tokens <- paragraphs |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::get_stopwords(), by = "word") |>
  dplyr::filter(str_detect(word, "[a-z]"))
```

# Document-term matrix

Rare terms (under `r tx$min_docfreq` paragraphs) are trimmed to stabilize estimation. `meta` is built from the surviving dfm rownames, so alignment with the model survives any dropped document. At work, check `nfeat(dfm_para)`: a vocabulary in the low thousands is comfortable at this scale, and a much larger one argues for a higher floor.

```{r dtm}
dfm_para <- tokens |>
  dplyr::count(para_uid, word) |>
  tidytext::cast_dfm(para_uid, word, n) |>
  quanteda::dfm_trim(min_docfreq = tx$min_docfreq)
stm_in <- quanteda::convert(dfm_para, to = "stm")
# as.data.frame, not a tibble: stm subsets metadata with base [ , ] semantics
# when building the prevalence model matrix, and a tibble returns a column
# rather than a vector there ("Error creating model matrix").
meta   <- tibble(para_uid = rownames(dfm_para)) |>
  dplyr::left_join(paragraphs, by = "para_uid") |>
  as.data.frame()
stopifnot("every dfm row must match a paragraph" = !anyNA(meta$Year),
          "meta must align with the stm documents" =
            nrow(meta) == length(stm_in$documents))

# Spline basis as ORDINARY COLUMNS, so no function call ever appears inside a
# formula. stm builds its prevalence model matrix by evaluating the formula
# against the metadata, and a call there (s(), ns(), or anything namespaced)
# is the documented cause of "Error creating model matrix". prev_form then
# reads ~ yr1 + yr2 + ..., which is also what estimateEffect uses below.
bs <- splines::ns(meta$Year, df = tx$spline_df)
colnames(bs) <- str_c("yr", seq_len(ncol(bs)))
meta <- cbind(meta, as.data.frame(bs))
prev_rhs  <- str_c(colnames(bs), collapse = " + ")   # "yr1 + yr2 + ..."
prev_form <- as.formula(str_c("~ ", prev_rhs))
```

# Choosing K

stm's own parallelism forks, which is unavailable on Windows, so the held-out grid is parallelized explicitly with furrr. Read the panels together: take the smallest K where held-out likelihood has flattened and residual dispersion has stopped falling, then break ties on the coherence-exclusivity frontier. If the frontier is still improving at the top of the grid, widen `tx$K_grid` rather than extrapolating. Semantic coherence is $\sum_{i<j}\log\{(D(v_i,v_j)+1)/D(v_i)\}$ over each topic's top words, with the denominator the higher-ranked word.

```{r searchK}
if (isTRUE(tx$run_searchK)) {
  future::plan(future::multisession, workers = tx$workers)
  ho <- stm::make.heldout(stm_in$documents, stm_in$vocab, seed = tx$seed)
  metrics <- cache_csv(p_out("searchK_metrics.csv"), {
    furrr::future_map(tx$K_grid, function(k) {
      fit <- stm::stm(ho$documents, ho$vocab, K = k, init.type = "Spectral",
                      prevalence = prev_form, data = meta, verbose = FALSE)
      tibble(K = k,
             heldout   = stm::eval.heldout(fit, ho$missing)$expected.heldout,
             dispersion = stm::checkResiduals(fit, ho$documents)$dispersion,
             coherence = mean(stm::semanticCoherence(fit, ho$documents)),
             exclusivity = mean(stm::exclusivity(fit)))
    }, .options = furrr::furrr_options(seed = tx$seed)) |> purrr::list_rbind()
  })
  future::plan(future::sequential)
  metrics |>
    tidyr::pivot_longer(-K) |>
    ggplot(aes(K, value)) + geom_line(color = viridis(1, begin = 0.35)) +
    geom_point() + facet_wrap(~ name, scales = "free_y") + labs(y = NULL)
} else {
  if (is.null(tx$K))
    stop("tx$K is NULL and tx$run_searchK is FALSE. Re-run the config chunk, ",
         "or set tx$K before continuing.")
  glue::glue("searchK skipped (tx$run_searchK = FALSE); K fixed at {tx$K}")
}
```

# Structural topic model

Year enters the prevalence model as a spline (df `r tx$spline_df`), so topic attention is allowed to drift over the period instead of being averaged away; spectral initialization is deterministic given the data, removing initialization luck from a fit everything downstream depends on.

```{r stm-fit}
stm_fit <- cache_rds(p_out("stm_fit.rds"),
  stm::stm(stm_in$documents, stm_in$vocab, K = tx$K, init.type = "Spectral",
           prevalence = prev_form, data = meta, verbose = FALSE))

theta <- stm_fit$theta
colnames(theta) <- str_c("topic_", seq_len(tx$K))
paragraph_theta <- meta |> dplyr::bind_cols(tibble::as_tibble(theta))
```

# Modal topic assignment and its diagnostics

Each paragraph is assigned its **modal** topic, so topics partition the corpus (mutually exclusive, collectively exhaustive) and can serve as sampling strata in Script 2. This is the same move as modal class assignment in latent class analysis, and it inherits the same caveat: a modal assignment is an estimate, not an observation, and treating it as observed pushes classification error downstream. Two LCA diagnostics report how much error to expect. **Entropy** $R^2$ summarizes overall assignment sharpness on a 0-1 scale; **AvePP**, the average posterior probability among a topic's own assigned paragraphs, does it per topic, with about 0.7 the conventional floor. Unlike LCA, STM supplies no classification error matrix over modal assignment, so no BCH-style correction is available here; the honest substitutes are the `min_theta` scope floor in Script 2 and its sensitivity analysis.

```{r modal}
#| tbl-colwidths: [7, 11, 9, 8, 33, 32]
paragraph_theta <- paragraph_theta |>
  dplyr::mutate(topic     = max.col(theta, ties.method = "first"),
                theta_max = theta[cbind(dplyr::row_number(), topic)])

ent <- -sum(theta * log(theta + 1e-12))
entropy_r2 <- 1 - ent / (nrow(theta) * log(tx$K))

topic_summary <- tibble(
    topic = seq_len(tx$K),
    prevalence = colMeans(theta),
    frex = apply(stm::labelTopics(stm_fit, n = 8)$frex, 1, str_c,
                 collapse = ", "),
    prob = apply(stm::labelTopics(stm_fit, n = 8)$prob, 1, str_c,
                 collapse = ", ")) |>
  dplyr::left_join(
    paragraph_theta |>
      dplyr::group_by(topic) |>
      dplyr::summarise(n_paras = dplyr::n(), avepp = mean(theta_max)),
    by = "topic")

readr::write_csv(topic_summary, p_out("topic_summary.csv"))
readr::write_csv(paragraph_theta, p_out("paragraph_theta.csv"))

glue::glue("Entropy R2 = {round(entropy_r2, 3)}; topics with AvePP < 0.7: ",
           "{sum(topic_summary$avepp < 0.7)} of {tx$K}")
topic_summary |>
  dplyr::mutate(dplyr::across(c(prevalence, avepp), ~ round(.x, 3))) |>
  dplyr::arrange(dplyr::desc(prevalence)) |>
  knitr::kable(caption = "Topics by prevalence with FREX and highest-probability terms, assigned paragraph counts, and AvePP. Script 2 labels these; the terms here are the evidence a human uses to check those labels.")
```

# Representative paragraphs

```{r thoughts}
#| tbl-colwidths: [10, 16, 74]
top4 <- topic_summary |> dplyr::slice_max(prevalence, n = 4) |>
  dplyr::pull(topic)
th <- stm::findThoughts(stm_fit, texts = meta$para_uid, n = 3, topics = top4)
tibble(topic = rep(top4, each = 3), para_uid = unlist(th$docs)) |>
  dplyr::left_join(paragraphs |> dplyr::select(para_uid, text),
                   by = "para_uid") |>
  dplyr::mutate(text = str_trunc(text, 150)) |>
  knitr::kable(caption = "Three highest-theta paragraphs for the four most prevalent topics.")
```

# Prevalence over time

```{r effects}
stm_year_effects <- cache_rds(p_out("stm_year_effects.rds"),
  stm::estimateEffect(
    as.formula(str_c("1:", tx$K, " ~ ", prev_rhs)),
    stm_fit, metadata = meta, uncertainty = "Global"))
```

Expected topic prevalence is predicted across the observed year range from the fitted effect object, with 95 percent intervals. Two details make the prediction correct rather than merely plausible. The spline basis is rebuilt at the prediction grid with `predict(bs, ...)`, which reuses the **original knots**; calling `ns()` afresh on the grid would place knots differently and mis-predict silently. And the intervals come from the simulation draws `estimateEffect` already stores under `uncertainty = "Global"`, so they carry the topic-model uncertainty rather than regression standard errors alone.

```{r trend-estimates}
# Prediction grid: every observed year, with the basis evaluated at the
# ORIGINAL knots so the fitted coefficients apply.
grid <- tibble(Year = sort(unique(meta$Year)))
Xg   <- cbind(1, predict(bs, newx = grid$Year))

# estimateEffect stores per-topic simulation draws of the coefficient vector
# in $parameters[[k]][[j]]$est. Predicting each draw and taking quantiles
# propagates both the regression and the topic-model uncertainty.
draws_for_topic <- function(k) {
  purrr::map(stm_year_effects$parameters[[k]], function(sim)
    as.numeric(Xg %*% sim$est)) |>
    purrr::reduce(rbind) |> matrix(ncol = nrow(grid))
}
topic_trends <- purrr::map(seq_along(stm_year_effects$topics), function(i) {
  d <- draws_for_topic(i)
  grid |> dplyr::mutate(
    topic = stm_year_effects$topics[i],
    fit   = apply(d, 2, mean),
    lo    = apply(d, 2, quantile, 0.025),
    hi    = apply(d, 2, quantile, 0.975))
}) |> purrr::list_rbind()
readr::write_csv(topic_trends, p_out("topic_trends.csv"))
glue::glue("Predicted prevalence for {dplyr::n_distinct(topic_trends$topic)} \\
            topics across {nrow(grid)} years written to topic_trends.csv")
```

```{r trend-plots}
#| fig-height: 8
# All topics, paged so panels stay legible: 25 facets on one page renders each
# about 1.5 inches wide and the ribbons become unreadable.
trend_panel <- function(d, page, n_pages) {
  ggplot(d, aes(Year, fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2,
                fill = viridis(1, begin = 0.35)) +
    geom_line(linewidth = 0.7, color = viridis(1, begin = 0.35)) +
    facet_wrap(~ topic_label, scales = "free_y", ncol = 3) +
    labs(x = NULL, y = "expected topic proportion",
         title = glue::glue("Topic prevalence over time ({page} of {n_pages})"),
         subtitle = "Model-based estimates with 95 percent intervals")
}
trend_pages <- topic_trends |>
  dplyr::left_join(topic_summary |> dplyr::select(topic, prevalence),
                   by = "topic") |>
  dplyr::mutate(
    topic_label = str_c("Topic ", topic),
    page = (dplyr::dense_rank(dplyr::desc(prevalence)) - 1L) %/%
             tx$facets_per_page + 1L) |>
  dplyr::group_split(page)
purrr::iwalk(trend_pages, function(d, i)
  print(trend_panel(d, i, length(trend_pages))))
```

```{r handoff}
glue::glue("outputs/ written: paragraphs.csv, paragraph_theta.csv (ids, Year, \\
n_tokens, text, topic_1..K, topic, theta_max), topic_summary.csv, \\
stm_fit.rds, stm_year_effects.rds. Script 2 reads these.")
```
