---
title: "Violence Against Civilians and Leader Favorability in Mali"
subtitle: "Multilevel models of cercle-level ACLED events and a three-wave household survey"
author: "Kevin Linares"
date: today
format:
  typst:
    papersize: us-letter
    margin:
      x: 1in
      y: 1in
    fontsize: 11pt
    section-numbering: "1.1"
    fig-format: png
    fig-dpi: 300
    fig-width: 6.5
    fig-height: 3.5
execute:
  echo: false
  warning: false
  message: false
---

```{r config}
# ---- Settings: change these, not the scripts ------------------------------------
# The data file is prepared upstream (see R/prep_data.R for the columns it needs).
cfg <- list(
  data_file = "data/survey_sim.rds",
  pop_file = "data/population_example.csv",   # region, urban, population (adults); replace the example
  violence = "violence_civilian",              # cercle-wave event count used in every model
  violence_label = "Violence against civilians (ACLED)",
  covariates = c("female", "age", "ethnicity", "education"),   # M2a
  covariates_m2b = c("female", "age", "education"),            # M2b: an alternative set to compare
  final = "M2a",                               # covariate set carried into M3: fix on substantive grounds
  event_levels = c(0, 1, 2, 5),                # events per surveyed commune shown in predictions (above the data's maximum are dropped)
  model_dir = "models",                        # fitted models are saved here
  refit = TRUE                                 # set FALSE after the first run; TRUE after changes
)
```

```{r setup}
#| include: false
pacman::p_load(lme4, survey, srvyr, PracTools, marginaleffects, performance, broom.mixed, sf, viridis,
               patchwork, knitr, tidyverse)

source("R/models.R")         # runs R/prep_data.R first
source("R/poststratify.R")
source("R/maps.R")
source("R/export_tables.R")  # writes output/tables/*.csv

theme_set(theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank()))
pal <- viridis::viridis(3, end = 0.8)

# Formatting helpers
pct <- function(x) sprintf("%.1f", 100 * x)
f2 <- function(x) sprintf("%.2f", x)
f3 <- function(x) sprintf("%.3f", x)
pv <- function(p) if (p < 0.001) "< .001" else paste("=", sprintf("%.3f", p))

se_wt <- sqrt(diag(as.matrix(stats::vcov(m3))))[["log_wt_c"]]
b_wt <- lme4::fixef(m3)[["log_wt_c"]]
n_cw <- dplyr::n_distinct(ad$cercle_wave)
n_cercle <- dplyr::n_distinct(ad$cercle_id)
svy_main <- dplyr::filter(violence_rows, model == "Marginal, weighted, cercle clusters (design-based check)")
svy_unweighted <- dplyr::filter(violence_rows, model == "Marginal, unweighted, cercle clusters")
svy_commune <- dplyr::filter(violence_rows, model == "Marginal, weighted, commune clusters")
tau_m3 <- sum(as.data.frame(lme4::VarCorr(m3))$vcov)   # total random-intercept variance
zeger <- sqrt(1 + 0.346 * tau_m3)                       # expected conditional / marginal ratio
pred_at <- function(k) predicted$estimate[which.min(abs(predicted$events - k))]
```

# Question and data

Is violence against civilians in a respondent's cercle associated with how favorably they view the national leader, comparing people in the same region, urban or rural area, and survey wave? Violence is the number of ACLED events against civilians in the cercle during each wave's survey window, counted over the communes surveyed there and divided by their number (section 2.1). About eight in ten respondents view the leader favorably, so any association is expected to be modest.

The analysis has three objectives, and each sets the method (West, 2025, following Hansen, Madow & Tepping, 1983):

- **Association (primary):** how favorability differs with violence for people in the same cercle setting. This is a conditional effect, estimated with a multilevel model (M3).
- **Check:** the same association averaged over the population, estimated design-based with the survey weights. It tests whether the design changes the conclusion.
- **Description:** percent favorable in the adult population, from the survey weights poststratified to population counts by region and area.

The analysis file has `r format(nrow(ad), big.mark = ",")` respondents in `r dplyr::n_distinct(ad$commune_id)` communes and `r n_cercle` cercles. Each wave is a separate sample: communes are drawn within strata (region by urban or rural), with eight respondents per commune. Violence takes one value per cercle and wave, so the violence effect rests on `r n_cw` cercle-waves in `r n_cercle` cercles, not on the number of respondents.

```{r}
#| label: tbl-waves
#| tbl-cap: "Sample by wave: design-weighted percent favorable (design-based SE), cercles, median events per surveyed commune, and the Kish design effect from unequal weights with the effective sample size it implies."
des_sample |>
  dplyr::group_by(wave) |>
  dplyr::summarise(
    fav = srvyr::survey_mean(fav),
    n = srvyr::unweighted(dplyr::n()),
    cercles = srvyr::unweighted(dplyr::n_distinct(cercle_id)),
    events = srvyr::unweighted(stats::median(events_per_commune)),
    deff_k = srvyr::unweighted(PracTools::deffK(wt))
  ) |>
  dplyr::transmute(Wave = wave, Respondents = n, Cercles = cercles,
                   `% favorable` = pct(fav), SE = pct(fav_se), `Median events per commune` = f2(events),
                   `Kish deff` = f2(deff_k), `Effective n` = round(n / deff_k)) |>
  kable(format = "pipe")
```

The Kish design effect, $1 + \text{relvar}(w)$, is the variance inflation from unequal weights alone (Kish, 1965); it ignores clustering, which the models handle.

# Models

## Violence measure

The ACLED count for cercle $k$ in wave $w$ is a sum over the communes surveyed there, so a cercle with more surveyed communes would look more violent. It is divided by the number of surveyed communes, $n_{kw}$:

$$V_{kw} = \log\left(1 + \frac{E_{kw}}{n_{kw}}\right)$$

Zero events gives 0; adding $\log 2 \approx 0.69$ doubles $(1 + E/n)$. Results are also given as predicted favorability at `r toString(event_levels)` events per surveyed commune (the largest value in the data is `r f2(max_events)`). Respondent covariates $\mathbf{x}_i$: M2a uses `r toString(cfg$covariates)`; M2b uses `r toString(cfg$covariates_m2b)`. M3 carries the `r cfg$final` set.

## Model sequence

| Model | Contents | Role |
|----|----|----|
| M1 | cercle and cercle-by-wave random intercepts | how much favorability clusters |
| M2a | M1 + violence, wave, region, urban, covariate set A |  |
| M2b | the same with covariate set B | checks that the violence estimate does not depend on the covariate choice; AIC, BIC, and a likelihood ratio test if one set contains the other |
| M3 | the chosen M2 + centered log survey weight | **final model** for estimates, predictions, and maps |
| Design-based | M3's fixed effects with survey weights, cercles as clusters | check on M3 |
| Sensitivity | M3 plus a commune intercept; M3 with stratum-by-wave indicators; violence-by-weight interaction; violence split into the cercle's usual level and its wave-to-wave change |  |

The multilevel models are logistic (`lme4::glmer`, maximum likelihood with the Laplace approximation), unweighted, on the same respondents. Both random intercepts are part of every model by design, not chosen by a test: violence takes one value per cercle and wave, and the cercle-by-wave intercept is what accounts for respondents sharing that value. Dropping it understates the standard error of the violence effect, and testing whether to keep it would bias the analysis toward finding an effect. Communes are not a random effect in M3: violence does not vary within a cercle-wave, and on the real data the commune variance made models fail to converge. M3 with a commune intercept is reported as a sensitivity check.

## Equations

Respondent $i$ in cercle $k$, wave $w$; $y_{ikw} = 1$ for a favorable view.

**M1.** $\operatorname{logit} \Pr(y = 1) = \gamma_{0} + u_{k} + v_{kw}$, with $u_k \sim N(0, \tau_k)$ and $v_{kw} \sim N(0, \tau_{kw})$.

$u_k$ is the cercle's lasting level; $v_{kw}$ is how that cercle departs from it in one wave. On the latent scale the individual variance is $\pi^2/3 \approx 3.29$, so

$$\rho_{\text{cercle}} = \frac{\tau_k}{\tau_k + \tau_{kw} + \pi^2/3}, \qquad
\rho_{\text{cercle-wave}} = \frac{\tau_k + \tau_{kw}}{\tau_k + \tau_{kw} + \pi^2/3}$$

**M2a and M2b.** $$\operatorname{logit} \Pr(y = 1) = \gamma_0 + \beta V_{kw} + \sum_{w=2}^{3}\delta_w \text{Wave}_w + \sum_{r \ne \text{ref}} \alpha_r \text{Region}_r + \lambda\,\text{Urban} + \mathbf{x}_i'\boldsymbol\theta + u_k + v_{kw}$$

The reference region is `r levels(ad$region)[1]` and the reference wave is wave 1. Region and urban are the survey strata and enter as fixed effects, so violence is compared within region and area. Wave enters as fixed effects (the national shift between waves). M2a and M2b differ only in $\mathbf{x}_i$.

**M3 (final).** The chosen M2 plus $\phi \log \tilde w_i$, the centered log of the survey weight. A weighted multilevel model needs a weight at each level (the commune's selection probability and the respondent's weight given the commune; Pfeffermann et al., 1998; Rabe-Hesketh & Skrondal, 2006). Only the final respondent weight is available, and the commune frame needed to rebuild the commune weight (communes per stratum) is not, so the weight enters as a covariate instead (DuMouchel & Duncan, 1983; Little, 1991). A clearly nonzero $\phi$ would mean people with larger weights differ in favorability beyond the predictors (informative sampling). Communes were selected with equal probability within strata, so their selection depends only on region and area, which are in the model. Including the weight changes what the other coefficients are conditional on, so M2 (without it) is shown alongside M3.

**Design-based check.** The same fixed effects as a marginal logistic model (`survey::svyglm`) with the survey weights (rescaled within wave), cercles as clusters and regions as strata, and Taylor-linearized standard errors (Binder, 1983). Clustering at the cercle is a choice about the estimand, not a description of the sampling: communes were sampled, but violence is assigned per cercle and wave, and for an association meant to hold beyond these particular cercle shocks the uncertainty belongs where the exposure is assigned (Moulton, 1990; Abadie et al., 2023). Clustering at the commune, as the sampling alone would suggest, gives a smaller standard error; both are shown. Urban cannot be a stratum here because a cercle can contain urban and rural communes. With `r n_cercle` cercles in `r dplyr::n_distinct(ad$region)` regions the check has few degrees of freedom. Marginal coefficients are smaller than conditional ones by roughly $\sqrt{1 + 0.346\,\tau}$, where $\tau$ is the total random-intercept variance (Zeger, Liang & Albert, 1988); for M3 that is about `r f2(zeger)`. A larger gap comes from the weights, which an unweighted marginal fit with the same clustering separates out. The marginal rows are therefore compared with M3 on direction and precision, not on size.

**Sensitivity (commune intercept).** M3 plus a commune random intercept, $c_j \sim N(0, \tau_j)$, since communes are the sampled clusters.

**Sensitivity (usual level and change).** $V_{kw}$ is replaced by $\bar V_{k}$, the cercle's mean over the waves it was surveyed, and $V_{kw} - \bar V_{k}$ (Mundlak, 1978). The change coefficient asks whether a cercle is less favorable in a wave when its violence is above its usual level, so differences between cercles that do not change over time cannot explain it. Two cautions: wave effects are national, so trends specific to a region are not removed; and the count covers only the communes surveyed in that wave, so a change can partly reflect which communes were sampled.

**Poststratification.** Within each wave, the survey weights are scaled so that each region-by-area cell $c$ sums to its adult population $N_c$ (Lumley, 2010). Percent favorable and its standard error then come from the poststratified design:

$$\hat\theta_w = \sum_c \frac{N_c}{\sum_{c'} N_{c'}}\, \bar y^{\,w}_{cw}, \qquad \bar y^{\,w}_{cw} = \text{weighted percent favorable in cell } c \text{, wave } w$$

This is design-based rather than model-based (MRP): with counts for region and area only, a model cannot adjust for anything the design does not already cover, and M3's predictions, which include each cercle's own effects, nearly reproduce the observed cell percentages.

# Results

## How much favorability clusters

```{r}
#| label: tbl-icc
#| tbl-cap: "Latent-scale intraclass correlations from M1."
icc |>
  dplyr::transmute(`Two respondents in the` = level, ICC = f3(icc)) |>
  kable(format = "pipe")
```

## Model comparison

```{r}
#| label: tbl-compare
#| tbl-cap: "Fit of the model sequence (same respondents, maximum likelihood). The test compares M2b with M2a when the covariate sets are nested."
m2_test <- if (is.null(test_m2)) "not nested: compare AIC, BIC" else
  sprintf("chi2(%d) = %.2f, p %s", abs(test_m2$Df[2]), test_m2$Chisq[2], pv(p_m2))
comparison |>
  dplyr::mutate(test = dplyr::if_else(model == "M2b", m2_test, "")) |>
  dplyr::transmute(Model = model, `Log-lik.` = f2(log_lik), Parameters = n_par,
                   AIC = f2(aic), BIC = f2(bic), Test = test, Singular = singular) |>
  kable(format = "pipe")
```

M3 uses the `r cfg$final` covariates. The final set is fixed on substantive grounds (variables that could affect both exposure to violence and views of the leader), not by the fit statistics above; the other set is shown to check that the violence estimate does not depend on the choice.

## The violence effect

```{r}
#| label: tbl-violence
#| tbl-cap: "Violence coefficient (log-odds per unit of log(1 + events per commune)). Multilevel rows: z and normal p-value. Marginal rows: t with the design's residual degrees of freedom; their estimates are population-averaged, so compare direction and precision with M3, not size. The commune-cluster row shows the effect of the clustering choice."
violence_rows |>
  dplyr::transmute(Model = model, Estimate = f3(estimate), SE = f3(std.error), `z or t` = f2(stat),
                   df = dplyr::if_else(is.na(df), "", as.character(df)),
                   p = dplyr::if_else(p.value < 0.001, "< .001", sprintf("%.3f", p.value))) |>
  kable(format = "pipe")
```

In M3, a one-unit increase in violence changes the log-odds of a favorable view by `r f3(b3$estimate)` (95% CI `r f3(ci3[1])` to `r f3(ci3[2])`, $p$ `r pv(b3$p.value)`), holding wave, region, urban, and respondent characteristics constant. The design-based check (weighted, cercle clusters) estimates `r f3(svy_main$estimate)`, $t =$ `r f2(svy_main$stat)` on `r svy_main$df` degrees of freedom ($p$ `r pv(svy_main$p.value)`). M3's estimate is `r f2(b3$estimate / svy_main$estimate)` times as large, against about `r f2(zeger)` expected from the conditional-marginal difference alone; the unweighted marginal fit (`r f3(svy_unweighted$estimate)`) shows how much of the rest is due to the weights. Clustering at the commune gives $t =$ `r f2(svy_commune$stat)` on `r svy_commune$df` degrees of freedom; the difference is the cost of treating communes in the same cercle as independent information about violence.

```{r}
#| label: tbl-predicted
#| tbl-cap: "M3: predicted percent favorable if every cercle had the given number of events per surveyed commune, everything else as observed, typical cercle (95% CI). Levels are limited to the range observed in the data."
predicted |>
  dplyr::transmute(`Events per commune` = events, `Predicted % favorable` = pct(estimate),
                   `95% CI` = paste(pct(conf.low), "to", pct(conf.high))) |>
  kable(format = "pipe", align = "rrr")
```

**Size of the association and what the study can detect.** Going from 0 to 5 events per surveyed commune changes predicted favorability by `r pct(diff_0_5$estimate)` percentage points (95% CI `r pct(diff_0_5$conf.low)` to `r pct(diff_0_5$conf.high)`). Changes outside this interval are not well supported by the data. The smallest 0-to-5 change this design would detect with 80% power (two-sided, 5% level) is about `r pct(abs(mde_0_5))` percentage points; an estimate smaller than that is one the study was not designed to resolve, and an interval that includes zero means the data cannot distinguish the association from none, which is not the same as showing there is none.

**Design.** The log-weight coefficient is `r f3(b_wt)` (SE `r f3(se_wt)`), and the violence-by-log-weight interaction is `r f3(wt_slope[["Estimate"]])` (SE `r f3(wt_slope[["Std. Error"]])`). Values far from zero would suggest the weights carry information the unweighted model misses, about favorability or about the violence slope itself (DuMouchel & Duncan, 1983). Values near zero are consistent with sampling that is not informative for this association, but they cannot prove it.

```{r}
#| label: fig-by-variable
#| fig-cap: "M3: predicted percent favorable for each level of each factor in the model, setting everyone to that level and keeping everything else as observed, typical cercle (95% CI)."
#| fig-height: 7
predicted_by_var |>
  dplyr::mutate(level = forcats::fct_inorder(level)) |>
  ggplot(aes(estimate, level)) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = pal[1], size = 0.25) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Predicted % favorable", y = NULL)
```

## Sensitivity: usual level and change

```{r}
#| label: tbl-sensitivity
#| tbl-cap: "Violence split into the cercle's usual level and its wave-to-wave change (log-odds, SE)."
summary(m_split)$coefficients[c("v_cercle", "v_change"), ] |>
  tibble::as_tibble(rownames = "term") |>
  dplyr::transmute(Term = c("Cercle's usual level", "Change from usual level"), Estimate = f3(Estimate),
                   SE = f3(`Std. Error`), z = f2(`z value`)) |>
  kable(format = "pipe")
```

The change coefficient is the localized, over-time comparison; the usual-level coefficient compares cercles with persistently different violence. Their difference is `r f3(split_diff)` (SE `r f3(split_diff_se)`, $p$ `r pv(split_diff_p)`).

## Poststratified favorability

The survey weights are poststratified to the adult population of each region and area, separately by wave. `r if (pop_is_example) "**The population counts used here are placeholder values for testing (data/population_example.csv); replace them with the real counts before reporting.** " else ""`If the weights had been calibrated to these counts, each cell's share of the weights would match its share of the population; the largest gap is `r worst_share$region` `r worst_share$urban` in wave `r worst_share$wave`, where the weights give `r pct(worst_share$weight_share)`% against `r pct(worst_share$population_share)`% of the population. Large gaps mean the weights and the counts describe different populations, and the estimate leans on the counts.

```{r}
#| label: tbl-poststrat
#| tbl-cap: "Percent favorable by wave: sample (unweighted) and poststratified to the region-by-area population (95% CI from the poststratified design)."
ps_national |>
  dplyr::transmute(Wave = wave, `Sample %` = pct(sample), `Poststratified %` = pct(estimate),
                   `95% CI` = paste(pct(conf.low), "to", pct(conf.high))) |>
  kable(format = "pipe")
```

```{r}
#| label: fig-poststrat
#| fig-cap: "(A) Adult population by region and area (poststratification counts). (B) Weighted percent favorable for each region and area by wave (95% CI); dashed line: poststratified national estimate. Cells with a single surveyed commune (Gao urban in the test data) have no design-based interval and are shown as points only."
#| fig-height: 4.2
ps_plot
```

# Diagnostics

```{r}
#| label: tbl-vif
#| tbl-cap: "Collinearity of the M3 predictors (generalized variance inflation factors)."
performance::check_collinearity(m3) |>
  tibble::as_tibble() |>
  dplyr::transmute(Term = Term, VIF = f2(VIF)) |>
  kable(format = "pipe")
```

M3 is `r if (lme4::isSingular(m3)) "singular" else "not singular"`. Convergence messages from lme4: `r if (is.null(m3@optinfo$conv$lme4$messages)) "none" else sub("[.[:space:]]+$", "", gsub("\\s+", " ", toString(m3@optinfo$conv$lme4$messages)))`. lme4 checks the absolute gradient, which large samples can trip without a real problem; the relative gradient, the step the optimizer would still take, is `r if (is.na(rel_gradient)) "not available (lme4 did not store the derivatives)" else format(rel_gradient, digits = 2)` (below about 0.001 means the fit has converged).

```{r}
#| label: fig-checks
#| fig-cap: "(A) M3 calibration: observed percent favorable in 20 bins of expected probability, against the range the model itself produces (gray, 95% of 200 simulated data sets with new cercle effects). (B) M3 cercle random effects against a normal distribution."
#| fig-height: 3
p_a <- ggplot(calibration, aes(expected, observed)) +
  geom_abline(color = "gray40", linetype = "dashed") +
  geom_errorbar(aes(ymin = sim_low, ymax = sim_high), color = "gray70", width = 0) +
  geom_point(color = pal[1], size = 0.9) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "A. Calibration", x = "Expected % favorable", y = "Observed % favorable")
u_hat <- lme4::ranef(m3)$cercle_id |> tibble::as_tibble() |> dplyr::rename(u = `(Intercept)`)
p_b <- ggplot(u_hat, aes(sample = u)) +
  stat_qq(color = pal[2], size = 0.8) + stat_qq_line(color = pal[3]) +
  labs(title = "B. Cercle effects", x = "Normal quantiles", y = "Estimated cercle effect")
p_a | p_b
```

# Maps

Gray cercles had no surveyed commune, so there is neither survey data nor an ACLED count for them.

```{r}
#| label: fig-map-violence
#| fig-cap: "Violence by cercle: events per surveyed commune, averaged over the waves the cercle was surveyed. Bins follow the data (zero, then quartiles of the positive values)."
#| fig-height: 5
map_violence
```

```{r}
#| label: fig-map-predicted
#| fig-cap: "M3 predicted percent favorable by cercle and wave, including each cercle's own effects, averaged over the respondents of its surveyed communes. It describes the surveyed communes, not the whole cercle."
#| fig-height: 4.1
map_predicted
```

# How to explain the result

> Among people surveyed in the same region, area, and survey round, predicted favorability toward the leader is about `r pct(pred_at(0))`% where a cercle had no violence against civilians and about `r pct(pred_at(5))`% where it had 5 events per surveyed commune, a difference of `r pct(diff_0_5$estimate)` percentage points (95% CI `r pct(diff_0_5$conf.low)` to `r pct(diff_0_5$conf.high)`). This is an association, not proof that violence caused the change.

# Changes from the previous version

Two specification errors were corrected; both follow from the level at which violence is measured, not from the results.

- **Cercle-by-wave intercept.** It was previously kept or dropped by a likelihood ratio test. It is now part of every model, because violence takes one value per cercle and wave; dropping it understates the standard error of the violence effect.
- **Design-based check.** It previously used communes as the clusters, which understates the standard error of a cercle-level exposure. It now uses cercles; the commune version is shown for comparison.

# Limitations

- **Few violence values.** Violence varies across `r n_cw` cercle-waves in `r n_cercle` cercles, which limits precision however many people were interviewed.
- **Surveyed communes only.** ACLED counts exist only where a commune was surveyed, and are scaled by the number surveyed; they describe the surveyed parts of each cercle, not the whole cercle.
- **Communes in M3.** Communes are not a random effect in M3; the sensitivity model with a commune intercept shows whether that matters.
- **Unweighted multilevel models.** A weighted multilevel model needs commune selection weights, which are not available. The log-weight terms and the design-based check probe informative sampling but do not replace weighting, and M3 has region and area main effects rather than all stratum indicators (the stratum-by-wave sensitivity row shows the difference).
- **Few clusters in the design-based check.** Bamako is a single cercle and Gao urban has one commune per wave, so both are single-cluster strata; their variance is approximated (`survey.lonely.psu = "adjust"`).
- **Poststratification.** The counts cover region and area only, so within each cell the mix of age, sex, and education follows the weighted sample.
- **Timing.** Events are counted over each wave's survey window, so some may occur after a respondent was interviewed. Counting only events before the interview date would fix this if interview dates are available.
- **Violence measure.** ACLED undercounts remote areas. Because undercounting is tied to where violence happens, its effect on the association can go either way. Perpetrator is not available.
- **Sensitive outcome.** Favorability toward the leader may be over-reported where people fear consequences.
- **Coverage.** Insecure areas are harder to survey; estimates describe the population that could be reached.

# Output files

M3's coefficients, random effects, and predicted probabilities are written to `output/tables/` as CSV files: `m3_fixed_effects.csv`, `m3_random_effects.csv`, `violence_by_model.csv`, `m3_predicted_by_events.csv`, `m3_change_0_to_5.csv`, `m3_predicted_by_variable.csv`, `poststratified_by_cell.csv`, and `poststratified_national.csv`.

# References

Abadie, A., Athey, S., Imbens, G. W., & Wooldridge, J. M. (2023). When should you adjust standard errors for clustering? *Quarterly Journal of Economics, 138*(1), 1-35. https://doi.org/10.1093/qje/qjac038

Binder, D. A. (1983). On the variances of asymptotically normal estimators from complex surveys. *International Statistical Review, 51*(3), 279-292.

DuMouchel, W. H., & Duncan, G. J. (1983). Using sample survey weights in multiple regression analyses of stratified samples. *Journal of the American Statistical Association, 78*(383), 535-543.

Gelman, A. (2007). Struggles with survey weighting and regression modeling. *Statistical Science, 22*(2), 153-164.

Hansen, M. H., Madow, W. G., & Tepping, B. J. (1983). An evaluation of model-dependent and probability-sampling inferences in sample surveys. *Journal of the American Statistical Association, 78*(384), 776-793.

Kish, L. (1965). *Survey sampling*. Wiley.

Little, R. J. A. (1991). Inference with survey weights. *Journal of Official Statistics, 7*, 405-424.

Lumley, T. (2010). *Complex surveys: A guide to analysis using R*. Wiley. (R packages `survey` and, as its tidy interface, `srvyr`.)

Moulton, B. R. (1990). An illustration of a pitfall in estimating the effects of aggregate variables on micro units. *Review of Economics and Statistics, 72*(2), 334-338.

Mundlak, Y. (1978). On the pooling of time series and cross section data. *Econometrica, 46*(1), 69-85.

OCHA (United Nations Office for the Coordination of Humanitarian Affairs). Mali administrative boundaries, levels 0 to 3 (COD-AB). Source: Direction Nationale des Collectivités Territoriales (DNCT), release of 10 November 2021. Humanitarian Data Exchange.

Pfeffermann, D., Skinner, C. J., Holmes, D. J., Goldstein, H., & Rasbash, J. (1998). Weighting for unequal selection probabilities in multilevel models. *Journal of the Royal Statistical Society: Series B, 60*(1), 23-40.

Rabe-Hesketh, S., & Skrondal, A. (2006). Multilevel modelling of complex survey data. *Journal of the Royal Statistical Society: Series A, 169*(4), 805-827.

Raleigh, C., Linke, A., Hegre, H., & Karlsen, J. (2010). Introducing ACLED: An armed conflict location and event dataset. *Journal of Peace Research, 47*(5), 651-660.

Valliant, R., Dever, J. A., & Kreuter, F. (2018). *Practical tools for designing and weighting survey samples* (2nd ed.). Springer. (R package `PracTools`.)

West, B. T. (2025). Accounting for complex sample designs when fitting multilevel and marginal models. Applications of Statistical Modeling, Module 1, Lecture 6 (course slides). University of Michigan.

Zeger, S. L., Liang, K.-Y., & Albert, P. S. (1988). Models for longitudinal data: A generalized estimating equation approach. *Biometrics, 44*(4), 1049-1060.
