---
title: "Violence Against Civilians and Leader Favorability in Mali"
subtitle: "Multilevel models of commune-level ACLED events and a three-wave household survey"
author: "Kevin Linares"
date: today
format:
  typst:
    papersize: us-letter
    margin:
      x: 1in
      y: 1in
    fontsize: 11pt
    section-numbering: "1.1"
    fig-format: png
    fig-dpi: 300
    fig-width: 6.5
    fig-height: 3.5
execute:
  echo: false
  warning: false
  message: false
---

```{r config}
# ---- Settings: change these, not the scripts ------------------------------------
# The data file is prepared upstream (see R/prep_data.R for the columns it needs).
cfg <- list(
  data_file = "data/survey_sim.rds",
  violence = "violence_civilian",          # event count used in every model
  violence_label = "Events of violence against civilians (ACLED)",
  covariates = c("female", "age", "ethnicity", "education"),
  cercle = TRUE,                           # keep the cercle random intercept (read the M1 test)
  strata = FALSE,                          # strata instead of region + urban (read the M2 test)
  event_levels = c(0, 1, 5, 20),           # event counts shown in the predictions
  model_dir = "models",                    # fitted models are saved here
  refit = TRUE                             # set FALSE after the first run; TRUE after changing data, covariates, or switches
)
```

```{r setup}
#| include: false
pacman::p_load(lme4, survey, marginaleffects, performance, broom.mixed, modelsummary, tinytable,
               sf, viridis, patchwork, knitr, tidyverse)

source("R/models.R")   # runs R/prep_data.R first
source("R/maps.R")

theme_set(theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank()))
pal <- viridis::viridis(3, end = 0.8)

# Formatting helpers
pct <- function(x) sprintf("%.1f", 100 * x)
f2 <- function(x) sprintf("%.2f", x)
f3 <- function(x) sprintf("%.3f", x)
pv <- function(p) if (p < 0.001) "< .001" else paste("=", sprintf("%.3f", p))

# Readable coefficient names in the appendix: "regionGao" -> "region: Gao"
term_vars <- c("wave", "region", "urban", "strata", cfg$covariates)
clean_term <- function(x) {
  purrr::map_chr(x, \(t) {
    if (t == "v") return("Violence, log(1 + events)")
    if (t == "log_wt_c") return("Log weight (centered)")
    if (grepl("^SD \\(Intercept ", t)) return(sub("^SD \\(Intercept (.*)_id\\)$", "SD: \\1", t))
    hit = term_vars[startsWith(t, term_vars) & nchar(t) > nchar(term_vars)]
    if (length(hit) == 0) return(t)
    var = hit[which.max(nchar(hit))]
    paste0(var, ": ", substring(t, nchar(var) + 1))
  })
}

ci3 <- b3$estimate + c(-1.96, 1.96) * b3$std.error
se_wt <- sqrt(diag(as.matrix(stats::vcov(m3))))[["log_wt_c"]]
b_wt <- lme4::fixef(m3)[["log_wt_c"]]
w_share <- stats::var(ad$v_within_cercle) / stats::var(ad$v)
```

# Question and data

Is violence against civilians in a respondent's commune associated with how favorably they view the national leader, comparing people in the same region, urban or rural area, and survey wave? Violence is the number of ACLED events in the commune during each wave's survey window. About eight in ten respondents view the leader favorably, so any association is expected to be modest. The model sequence and decision rules were fixed before the violence coefficient was examined.

The analysis file has `r format(nrow(ad), big.mark = ",")` respondents in `r dplyr::n_distinct(ad$commune_id)` communes and `r dplyr::n_distinct(ad$cercle_id)` cercles. Each wave is a separate sample: communes are drawn within `r nlevels(ad$strata)` strata (region by urban or rural), with eight respondents per commune.

```{r}
#| label: tbl-waves
#| tbl-cap: "Sample by wave: design-weighted percent favorable (design-based SE) and median events per respondent's commune."
survey::svyby(~fav, ~wave, des, survey::svymean) |>
  tibble::as_tibble() |>
  dplyr::left_join(ad |> dplyr::summarise(n = dplyr::n(), communes = dplyr::n_distinct(commune_id),
                                          events = stats::median(.data[[cfg$violence]]), .by = wave),
                   by = "wave") |>
  dplyr::transmute(Wave = wave, Respondents = n, Communes = communes,
                   `% favorable` = pct(fav), SE = pct(se), `Median events` = events) |>
  kable(format = "pipe")
```

# Models

## Measures

For commune $c$ in wave $w$ with $E_{cw}$ events, violence is $V_{cw} = \log(1 + E_{cw})$. Zero events gives 0; adding 1 multiplies $(1 + E)$ by $e \approx 2.72$, and adding $\log 2 \approx 0.69$ doubles it. Results are also given as predicted favorability at `r toString(cfg$event_levels)` events. Respondent covariates $\mathbf{x}_i$: `r toString(cfg$covariates)`.

## Model sequence

| Model | Adds | Compared by |
|---|---|---|
| M1a | commune random intercept | |
| M1b | adds a cercle random intercept | M1b vs M1a: likelihood ratio test with boundary correction, AIC, BIC |
| M2a | violence, wave, region, urban, covariates | |
| M2b | the strata in place of region and urban | M2b vs M2a: likelihood ratio test, AIC, BIC |
| M3 | the chosen structure plus the centered log survey weight | final model; design-based check with `svyglm` |
| Sensitivity | violence split into cercle average and within-cercle difference | |

All models are logistic multilevel models (`lme4::glmer`, maximum likelihood with the Laplace approximation), unweighted, on the same respondents, so likelihoods and AIC/BIC are comparable. The cercle level is kept if its test gives $p < .05$, AIC is lower, and the fit is not singular; the strata replace region and urban if $p < .05$ and AIC is lower.

## Equations

Respondent $i$ in commune $c$, cercle $k$, wave $w$; $y_{ickw} = 1$ for a favorable view.

**M1a.** $\operatorname{logit} \Pr(y = 1) = \gamma_{0} + u_{c}$, with $u_c \sim N(0, \tau_c)$.

**M1b.** $\operatorname{logit} \Pr(y = 1) = \gamma_{0} + u_{c} + r_{k}$, with $r_k \sim N(0, \tau_k)$: communes nested in cercles.

A logit model has no residual variance; on the latent scale the individual variance is $\pi^2/3 \approx 3.29$, so

$$
\rho_{\text{cercle}} = \frac{\tau_k}{\tau_k + \tau_c + \pi^2/3}, \qquad
\rho_{\text{commune}} = \frac{\tau_k + \tau_c}{\tau_k + \tau_c + \pi^2/3}
$$

$\rho_{\text{cercle}}$ is the correlation between two people in the same cercle but different communes; $\rho_{\text{commune}}$ between two people in the same commune. The cercle test is $H_0: \tau_k = 0$; because a variance cannot be negative, the p-value is half of the $\chi^2_1$ p-value from `anova()`.

**M2a.**
$$
\operatorname{logit} \Pr(y = 1) = \gamma_0 + \beta V_{cw} + \sum_{w=2}^{3}\delta_w \text{Wave}_w + \sum_{r \ne \text{ref}} \alpha_r \text{Region}_r + \lambda\,\text{Urban} + \mathbf{x}_i'\boldsymbol\theta + u_c + r_k
$$

The reference region is `r levels(ad$region)[1]` and the reference wave is wave 1.

**M2b.** The same, with $\sum_r \alpha_r \text{Region}_r + \lambda\,\text{Urban}$ replaced by $\sum_{s \ne \text{ref}} \kappa_s \text{Stratum}_s$. M2a is nested in M2b, so the test asks whether the urban-rural gap differs by region.

**M3 (final).** The chosen structure plus $\phi \log \tilde w_i$, the centered log of the survey weight. Strata are fixed effects and communes the random intercept; `glmer` cannot use sampling weights, so the weight enters as a covariate. A clearly nonzero $\phi$ would mean people with larger weights differ in favorability beyond the predictors (informative sampling; Pfeffermann et al., 1998; Gelman, 2007).

**Design-based check.** The same fixed effects fit with `survey::svyglm`: weights rescaled within wave, strata within wave, communes as primary sampling units, Taylor-linearized standard errors (Binder, 1983). Its coefficients are population-averaged and smaller than the multilevel ones by roughly $\sqrt{1 + 0.346\,\tau}$ (Zeger, Liang & Albert, 1988), so it is compared by $z$.

**Sensitivity (cercle split).** $V_{cw}$ is replaced by $\bar V_{kw}$, the average across the surveyed communes of cercle $k$ in wave $w$, and $V_{cw} - \bar V_{kw}$ (Mundlak, 1978). The within-cercle coefficient compares communes in the same cercle and wave, so stable cercle context cannot explain it.

# Results

## How much favorability clusters

```{r}
#| label: tbl-icc
#| tbl-cap: "Latent-scale intraclass correlations from the empty models."
icc |>
  dplyr::transmute(Model = model, Level = level, ICC = f3(icc)) |>
  kable(format = "pipe")
```

In M1b, `r pct(icc$icc[3])`% of the variation in the latent propensity to view the leader favorably lies between communes, of which `r pct(icc$icc[2])` percentage points are shared across communes of the same cercle.

## Model comparison

```{r}
#| label: tbl-compare
#| tbl-cap: "Fit of the model sequence (same respondents, maximum likelihood). Tests compare M1b with M1a and M2b with M2a."
tests <- tibble::tibble(
  model = c("M1b", "M2b"),
  test = c(sprintf("chi2(1) = %.2f, p %s (boundary)", test_m1$Chisq[2], pv(p_m1)),
           sprintf("chi2(%d) = %.2f, p %s", test_m2$Df[2], test_m2$Chisq[2], pv(p_m2)))
)
comparison |>
  dplyr::left_join(tests, by = "model") |>
  dplyr::transmute(Model = model, `Log-lik.` = f2(log_lik), Parameters = n_par,
                   AIC = f2(aic), BIC = f2(bic), Test = dplyr::coalesce(test, ""), Singular = singular) |>
  kable(format = "pipe")
```

M3 uses random intercepts for `r if (cfg$cercle) "commune and cercle" else "commune"` and fixed effects for `r paste(place, collapse = " and ")`, as set in the configuration after reading these tests.

## The violence effect

```{r}
#| label: tbl-violence
#| tbl-cap: "Violence coefficient (log-odds per unit of log(1 + events)) across models. The design-based check is population-averaged: compare z, not the estimate."
violence_rows |>
  dplyr::transmute(Model = model, Estimate = f3(estimate), SE = f3(std.error), z = f2(z)) |>
  kable(format = "pipe")
```

In M3, a one-unit increase in $\log(1 + E)$ changes the log-odds of a favorable view by `r f3(b3$estimate)` (95% CI `r f3(ci3[1])` to `r f3(ci3[2])`, $p$ `r pv(p_v)`), holding wave, `r paste(place, collapse = ", ")`, and respondent characteristics constant.

```{r}
#| label: tbl-predicted
#| tbl-cap: "M3: predicted percent favorable if every respondent's commune had the given number of events, everything else as observed, typical commune and cercle (95% CI)."
predicted |>
  dplyr::transmute(`Events in commune` = events, `Predicted % favorable` = pct(estimate),
                   `95% CI` = paste(pct(conf.low), "to", pct(conf.high))) |>
  kable(format = "pipe", align = "rrr")
```

**What the data can and cannot rule out.** Going from 0 to 5 events changes predicted favorability by `r pct(diff_0_5$estimate)` percentage points (95% CI `r pct(diff_0_5$conf.low)` to `r pct(diff_0_5$conf.high)`). Effects outside this interval are unlikely; if it includes zero, the data cannot distinguish the association from no association, which is not the same as showing no effect. The smallest 0-to-5-event change this design would detect with 80% power at the 5% level is about `r pct(abs(mde_0_5))` percentage points (2.8 standard errors of the coefficient, converted with the average slope of the logistic curve).

**Design.** The log-weight coefficient is `r f3(b_wt)` (SE `r f3(se_wt)`); a value more than about two standard errors from zero would suggest informative sampling. The design-based check gives $z =$ `r f2(violence_rows$z[4])` against `r f2(b3$z)` for M3.

```{r}
#| label: fig-by-variable
#| fig-cap: "M3: predicted percent favorable (95% CI). Wave and covariates: everyone set to each level, everything else as observed. Strata: average prediction for the people in each stratum."
#| fig-height: 7
predicted_by_var |>
  dplyr::mutate(level = forcats::fct_inorder(level)) |>
  ggplot(aes(estimate, level)) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = pal[1], size = 0.25) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Predicted % favorable", y = NULL)
```

## Sensitivity: cercle split

```{r}
#| label: tbl-sensitivity
#| tbl-cap: "Violence split into the cercle average and the commune's distance from it (log-odds, SE)."
summary(m_cercle)$coefficients[c("v_cercle", "v_within_cercle"), ] |>
  tibble::as_tibble(rownames = "term") |>
  dplyr::transmute(Term = c("Cercle average", "Within cercle"), Estimate = f3(Estimate),
                   SE = f3(`Std. Error`), z = f2(`z value`)) |>
  kable(format = "pipe")
```

The within-cercle coefficient compares communes with more or less violence than other surveyed communes of the same cercle in the same wave; `r pct(w_share)`% of the variation in violence is of this kind. The cercle-average coefficient compares cercles.

# Diagnostics

```{r}
#| label: tbl-vif
#| tbl-cap: "Collinearity of the M3 predictors (generalized variance inflation factors)."
performance::check_collinearity(m3) |>
  tibble::as_tibble() |>
  dplyr::transmute(Term = Term, VIF = f2(VIF)) |>
  kable(format = "pipe")
```

M3 is `r if (lme4::isSingular(m3)) "singular" else "not singular"`. Convergence messages from lme4: `r if (is.null(m3@optinfo$conv$lme4$messages)) "none" else toString(m3@optinfo$conv$lme4$messages)`.

```{r}
#| label: fig-checks
#| fig-cap: "(A) M3 calibration: observed percent favorable in 20 bins of expected probability, against the range the model itself produces (gray, 95% of 200 simulated data sets with new commune and cercle effects). (B) M3 commune random effects against a normal distribution."
#| fig-height: 3
p_a <- ggplot(calibration, aes(expected, observed)) +
  geom_abline(color = "gray40", linetype = "dashed") +
  geom_errorbar(aes(ymin = sim_low, ymax = sim_high), color = "gray70", width = 0) +
  geom_point(color = pal[1], size = 0.9) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "A. Calibration", x = "Expected % favorable", y = "Observed % favorable")
u_hat <- lme4::ranef(m3)$commune_id |> tibble::as_tibble() |> dplyr::rename(u = `(Intercept)`)
p_b <- ggplot(u_hat, aes(sample = u)) +
  stat_qq(color = pal[2], size = 0.7) + stat_qq_line(color = pal[3]) +
  labs(title = "B. Commune effects", x = "Normal quantiles", y = "Estimated commune effect")
p_a | p_b
```

# Maps

Gray communes were not surveyed in that wave: missing data, not zero.

```{r}
#| label: fig-map-violence
#| fig-cap: "Violence events by commune and wave (surveyed communes only)."
#| fig-height: 3.9
map_violence
```

```{r}
#| label: fig-map-cercle
#| fig-cap: "M3 predicted percent favorable by cercle and wave, including each cercle's own random effect, averaged over the respondents of its surveyed communes. It describes the surveyed communes, not the whole cercle."
#| fig-height: 4.1
map_predicted_cercle
```

The cercle map is the easier one to read: each cercle pools several surveyed communes. The commune map below shows more detail, but with eight respondents per commune each value leans heavily on its cercle and on the model, so single communes should not be read on their own.

```{r}
#| label: fig-map-predicted
#| fig-cap: "M3 predicted percent favorable by commune and wave, including each commune's and cercle's own random effects."
#| fig-height: 4.1
map_predicted
```

# How to explain the result

> Among people surveyed in the same region, area, and survey round, predicted favorability toward the leader is about `r pct(predicted$estimate[1])`% where a commune had no violence against civilians and about `r pct(predicted$estimate[predicted$events == 5])`% where it had 5 events, a difference of `r pct(diff_0_5$estimate)` percentage points (95% CI `r pct(diff_0_5$conf.low)` to `r pct(diff_0_5$conf.high)`). The data mostly compare different communes, so this is an association, not proof that violence caused the change.

# Limitations

- **Mostly between communes.** Violence differs mainly between communes. Stable differences the covariates miss could drive part of any association; the cercle-split model removes stable cercle-level context.
- **Unweighted multilevel models.** The log-weight covariate and the design-based check guard against informative sampling but do not replace weighting; commune-level selection weights were not available for a weighted multilevel model.
- **Violence measure.** Only surveyed communes have counts; ACLED undercounts remote areas, which most likely weakens any association. Perpetrator is not available.
- **Sensitive outcome.** Favorability toward the leader may be over-reported where people fear consequences, compressing differences between communes.
- **Coverage.** Insecure areas are harder to survey; estimates describe the population that could be reached.

# References

Bell, A., & Jones, K. (2015). Explaining fixed effects: Random effects modeling of time-series cross-sectional and panel data. *Political Science Research and Methods, 3*(1), 133-153.

Binder, D. A. (1983). On the variances of asymptotically normal estimators from complex surveys. *International Statistical Review, 51*(3), 279-292.

Gelman, A. (2007). Struggles with survey weighting and regression modeling. *Statistical Science, 22*(2), 153-164.

Mundlak, Y. (1978). On the pooling of time series and cross section data. *Econometrica, 46*(1), 69-85.

OCHA (United Nations Office for the Coordination of Humanitarian Affairs). Mali administrative boundaries, levels 0 to 3 (COD-AB). Source: Direction Nationale des Collectivités Territoriales (DNCT), release of 10 November 2021. Humanitarian Data Exchange.

Pfeffermann, D., Skinner, C. J., Holmes, D. J., Goldstein, H., & Rasbash, J. (1998). Weighting for unequal selection probabilities in multilevel models. *Journal of the Royal Statistical Society: Series B, 60*(1), 23-40.

Raleigh, C., Linke, A., Hegre, H., & Karlsen, J. (2010). Introducing ACLED: An armed conflict location and event dataset. *Journal of Peace Research, 47*(5), 651-660.

Zeger, S. L., Liang, K.-Y., & Albert, P. S. (1988). Models for longitudinal data: A generalized estimating equation approach. *Biometrics, 44*(4), 1049-1060.

# Appendix: full coefficient tables

```{r}
#| label: tbl-full
#| tbl-cap: "Final model M3: all fixed effects (log-odds [95% CI]) and random-effect standard deviations."
modelsummary::modelsummary(
  list(M3 = m3), output = "tinytable",
  estimate = "{estimate} [{conf.low}, {conf.high}]", statistic = NULL,
  coef_rename = clean_term, gof_map = c("nobs", "aic", "bic"), fmt = 2
)
```

```{r}
#| label: tbl-strata
#| tbl-cap: "M2b: stratum effects (log-odds [95% CI]) against the reference stratum."
modelsummary::modelsummary(
  list(M2b = m2b), output = "tinytable",
  estimate = "{estimate} [{conf.low}, {conf.high}]", statistic = NULL,
  coef_omit = "^(?!strata)", coef_rename = clean_term, gof_map = NA, fmt = 2
)
```
