---
title: "Violence Against Civilians and Leader Favorability in Mali"
subtitle: "Multilevel models of commune-level ACLED events and a three-wave household survey"
author: "Kevin Linares"
date: today
format:
  typst:
    papersize: us-letter
    margin:
      x: 1in
      y: 1in
    fontsize: 11pt
    section-numbering: "1.1"
    fig-format: png
    fig-dpi: 300
    fig-width: 6.5
    fig-height: 3.5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
# Packages for the whole pipeline (tidyverse last)
pacman::p_load(lme4, survey, marginaleffects, performance, jtools, huxtable, broom.mixed,
               sf, viridis, patchwork, knitr, glue, tidyverse)

# Everything is computed in plain R scripts; this report only presents results.
#   R/prep_data.R   edit for real data (file path, column names)
#   R/models.R      M1 to M4 and every number below
#   R/maps.R        maps (also saved to output/maps/)
source("R/models.R")
source("R/maps.R")

theme_set(theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank()))
pal <- viridis::viridis(3, end = 0.8)
pp  <- function(x) sprintf("%.1f", 100 * x)        # proportion -> percent / percentage points
f2  <- function(x) sprintf("%.2f", x)
f3  <- function(x) sprintf("%.3f", x)
pv  <- function(p) if (p < 0.001) "< .001" else sprintf("= %.3f", p)

# jtools table as a markdown table (renders in any output format)
md_summs <- function(...) {
  jtools::export_summs(..., digits = 3, error_format = "[{conf.low}, {conf.high}]") |>
    huxtable::to_md() |>
    knitr::asis_output()
}

# Short labels for the coefficient tables
coef_labels <- c("Violence, log(1 + events)" = "v_commune",
                 "Violence, within commune"  = "v_within",
                 "Violence, between communes" = "v_between",
                 "Wave 2" = "wave2", "Wave 3" = "wave3",
                 "Female" = "female", "Age (per decade, centered)" = "age10", "Urban" = "urban")
```

# Question

Is violence against civilians in a respondent's commune associated with how favorably they view the national leader, comparing people in the same region and survey wave? Violence is the number of ACLED events of violence against civilians in the commune during the same window as each wave's survey fieldwork. Favorability is a yes/no item, and most respondents answer yes, so the association is expected to be modest and the analysis is built to estimate it carefully rather than to find it.

# Data and design

```{r}
#| include: false
wave_table <- survey::svyby(~fav, ~wave, des, survey::svymean) |>
  tibble::as_tibble() |>
  dplyr::left_join(
    ad |> dplyr::summarise(n = dplyr::n(), communes = dplyr::n_distinct(commune_id),
                           villages = dplyr::n_distinct(village_w),
                           events_median = stats::median(events), .by = wave),
    by = "wave")

repeat_communes <- ad |> dplyr::distinct(commune_id, n_waves) |> dplyr::count(n_waves)
w_share <- stats::var(ad$v_within) / stats::var(ad$v_commune)
```

The analysis file has `r format(nrow(ad), big.mark = ",")` respondents (`r format(nrow(survey) - nrow(ad), big.mark = ",")` dropped for missing data) in `r dplyr::n_distinct(ad$commune_id)` communes. Communes surveyed in one, two, and three waves: `r toString(repeat_communes$n)`. Each wave is an independent two-stage sample within regions: communes first, then villages within communes.

```{r}
#| label: tbl-waves
#| tbl-cap: "Sample by wave, with the design-weighted percent favorable (design-based SE) and the median number of events per respondent's commune."
wave_table |>
  dplyr::transmute(Wave = wave, Respondents = n, Communes = communes, Villages = villages,
                   `% favorable` = 100 * fav, SE = 100 * se, `Median events` = events_median) |>
  kable(format = "pipe", digits = 1)
```

# Models

## Violence measure

For commune $c$ in wave $w$, with $E_{cw}$ events,

$$
V_{cw} = \log(1 + E_{cw}).
$$

Zero events gives $V = 0$. Adding 1 to $V$ multiplies $(1 + E)$ by $e \approx 2.72$; adding $\log 2 \approx 0.69$ doubles it. Because these steps are hard to picture, results are also shown as predicted favorability at 0, 1, 5, and 20 events. Only `r sprintf("%.0f%%", 100 * w_share)` of the variation in $V_{cw}$ is within communes across waves; the rest is between communes.

## Equations

Respondent $i$ lives in commune $c$ and was interviewed in wave $w$; $y_{icw} = 1$ if they view the leader favorably. All models are logistic multilevel models with communes as the grouping level; M3b adds region-wave cells and M4 villages.

**M1. Empty model.**

$$
\operatorname{logit} \Pr(y_{icw} = 1) = \gamma_{00} + u_{0c}, \qquad u_{0c} \sim N(0, \tau_{00})
$$

$\gamma_{00}$ is the log-odds of a favorable view in a typical commune and $\tau_{00}$ the variance of communes around it. A logistic model has no residual variance term, so the intraclass correlation uses the variance of the standard logistic distribution, $\pi^2/3 \approx 3.29$:

$$
\text{ICC} = \frac{\tau_{00}}{\tau_{00} + \pi^2/3}
$$

**M2a. Violence, random intercept.**

$$
\operatorname{logit} \Pr(y_{icw} = 1) = \gamma_{00} + \gamma_{10} V_{cw} + u_{0c}
$$

**M2b. Violence, random intercept and random slope.**

$$
\operatorname{logit} \Pr(y_{icw} = 1) = \gamma_{00} + \gamma_{10} V_{cw} + u_{0c} + u_{1c} V_{cw}
$$

$$
\begin{pmatrix} u_{0c} \\ u_{1c} \end{pmatrix} \sim N\left(\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \begin{pmatrix} \tau_{00} & \tau_{01} \\ \tau_{01} & \tau_{11} \end{pmatrix}\right)
$$

$\gamma_{10}$ is the violence effect in a typical commune; $\tau_{11}$ is how much that effect varies across communes; $\tau_{01}$ is the covariance between a commune's baseline favorability and its violence effect.

**Random slope test.** $H_0: \tau_{11} = 0$ against $H_A: \tau_{11} > 0$. The statistic is $\text{LRT} = -2(\ell_{\text{M2a}} - \ell_{\text{M2b}})$. Because a variance cannot be negative, the null value is on the boundary of the parameter space, and the p-value comes from a 50:50 mixture of $\chi^2_1$ and $\chi^2_2$ distributions. The decision rule was fixed in advance: keep the random slope only if $p < .05$ and M2b is not singular.

**M3a. Region, wave, and covariates.** The M2 winner plus region dummies (Bamako is the reference), wave dummies (wave 1 is the reference), and respondent covariates $\mathbf{x}_i$ (female, age per decade centered, ethnicity, education, urban):

$$
\operatorname{logit} \Pr(y_{icw} = 1) = \gamma_{00} + \gamma_{10} V_{cw} + \sum_{r \neq \text{Bamako}} \alpha_r \text{Region}_r + \gamma_2 \text{Wave}_2 + \gamma_3 \text{Wave}_3 + \mathbf{x}_i' \boldsymbol\beta + u_{0c} \left[ + u_{1c} V_{cw} \right]
$$

The term in brackets is included in M3a, M3b, and M4 only if the random slope test keeps it.

**M3b. Region-by-wave shocks.** M3a plus a random intercept for each region-wave cell, $v_{rw} \sim N(0, \tau_{rw})$, independent of $u_{0c}$:

$$
\operatorname{logit} \Pr(y_{icw} = 1) = \text{M3a} + v_{rw}
$$

The region and wave dummies stay. $v_{rw}$ captures shifts that hit one region in one wave (a local event, a security operation) and that separate region and wave dummies cannot absorb; with 24 cells, the cells with few respondents are pulled toward the region and wave main effects (partial pooling). Communes are nested in regions and crossed with waves, which `lme4` handles directly.

**Region-by-wave test.** $H_0: \tau_{rw} = 0$. M3b adds one variance, so the boundary mixture is $0.5\chi^2_0 + 0.5\chi^2_1$ (half the usual $\chi^2_1$ p-value). The rule was fixed in advance: M3b proceeds only if $p < .05$, its AIC is lower than M3a's, and it is not singular. The winner is called M3 below.

**M4. Final model: M3 plus the survey design.** The design enters the multilevel model directly (Pfeffermann et al., 1998; Gelman, 2007): strata are already the region dummies, first-stage units (communes) are already a random intercept, and M4 adds the second stage, villages within communes, $w_{vc} \sim N(0, \tau_{v})$, and the centered log of the survey weight as a covariate:

$$
\operatorname{logit} \Pr(y_{ivcw} = 1) = \text{M3} + \delta \log \tilde w_i + w_{vc}
$$

$\tau_v$ shows how much villages differ within communes. $\delta$ checks for informative sampling: if respondents with larger weights differ in favorability after the predictors, $\delta$ is clearly nonzero and estimates that ignore the weights may not represent the population. Sampling weights themselves cannot be passed to `glmer`: its `weights` argument means something else for binomial models.

**Design-based check.** For comparison, the same fixed effects are fit with `survey::svyglm`: weights rescaled within wave ($w^*_i = w_i \, n_w / \sum_{i \in w} w_i$), strata = regions within wave, first stage = communes within wave, second stage = villages, and Taylor-linearized standard errors at the first stage (Binder, 1983):

$$
\widehat{\operatorname{Var}}(\hat{\boldsymbol\beta}) = \hat{\mathbf J}^{-1}\left[\sum_h \frac{n_h}{n_h - 1}\sum_{j=1}^{n_h} (\mathbf z_{hj} - \bar{\mathbf z}_h)(\mathbf z_{hj} - \bar{\mathbf z}_h)'\right]\hat{\mathbf J}^{-1}, \qquad \mathbf z_{hj} = \sum_{i \in (h, j)} w^*_i \mathbf x_i (y_i - \hat\mu_i)
$$

Its coefficient is population-averaged, smaller than the multilevel (typical-commune) coefficient by roughly $\sqrt{1 + 0.346\,\tau_{00}}$ (Zeger, Liang & Albert, 1988), so the two are compared by $z$ statistics, not raw coefficients.

**Sensitivity: baseline level and change (M3-MW).** Violence is split into a commune's usual level, $\bar V_c$ (the mean of $V_{cw}$ over the waves in which it was surveyed), and its deviation in a given wave (Mundlak, 1978; Bell & Jones, 2015), with the same random part as M3:

$$
\operatorname{logit} \Pr(y_{icw} = 1) = \gamma_{00} + \gamma_W (V_{cw} - \bar V_c) + \gamma_B \bar V_c + \ldots + u_{0c}
$$

$\gamma_B$ compares communes with persistently more violence to those with less; $\gamma_W$ asks whether a commune is less favorable in a wave when its violence is above its usual level. If the two are similar, the single coefficient is a fair summary.

**Estimation.** All multilevel models are fit by maximum likelihood with the Laplace approximation (`lme4::glmer`, `bobyqa` optimizer), without weights, on the same respondents, so their likelihoods, AIC, and BIC can be compared. The design-based check has a pseudo-likelihood and is never compared by AIC or likelihood ratio tests.

# Results

## M1: how much favorability differs between communes

```{r}
#| comment: ""
# confint = FALSE: jtools 2.3 fails on confidence intervals for an intercept-only glmer
jtools::summ(m1, digits = 3, r.squared = FALSE, confint = FALSE, re.variance = "var")
```

The commune variance is $\hat\tau_{00} =$ `r f3(tau00_m1)`, so the ICC is `r f3(icc_m1)`: `r sprintf("%.0f%%", 100 * icc_m1)` of the variation in the latent propensity to view the leader favorably lies between communes. Respondents in the same commune are more alike than respondents in different communes, which is why every model that follows keeps a commune random intercept. In a typical commune, the predicted percent favorable is `r pp(plogis(lme4::fixef(m1)[[1]]))`%.

## M2: violence and the random slope test

```{r}
#| comment: ""
md_summs(m2a, m2b, model.names = c("M2a: random intercept", "M2b: random slope"),
         coefs = coef_labels[1])
```

```{r}
#| label: tbl-lrt
#| tbl-cap: "Likelihood ratio test for the random violence slope (50:50 mixture of chi-square with 1 and 2 df)."
tibble::tibble(
  Model = c("M2a: random intercept", "M2b: + random slope"),
  `Log-likelihood` = f2(c(stats::logLik(m2a), stats::logLik(m2b))),
    LRT = c("", f2(lrt_slope$lrt)),
  `p (mixture)` = c("", sub("= ", "", pv(lrt_slope$p_value))),
  Singular = c(lme4::isSingular(m2a), m2b_singular)
) |>
  kable(format = "pipe")
```

The likelihood ratio statistic is `r f2(lrt_slope$lrt)` with mixture $p$ `r pv(lrt_slope$p_value)``r if (m2b_singular) ", and M2b is singular (at least one variance or correlation is estimated at its boundary)" else ""`. `r if (keep_slope) "The random slope is kept: the violence effect differs across communes, so it is carried into M3a, M3b, and M4." else "The random slope is not kept, so M3a, M3b, and M4 have a commune random intercept only."` `r if (!keep_slope) paste0("This should not be read as proof that violence affects every commune in the same way. A random slope for a commune-level exposure is identified only by changes in violence within the same commune across waves, and only ", sprintf("%.0f%%", 100 * w_share), " of the variation in violence is of that kind. The data have little information to estimate the slope variance.") else ""`

## M3: region, wave, and region-by-wave shocks

```{r}
#| comment: ""
md_summs(m3a, m3b, model.names = c("M3a", "M3b: + region-wave"), coefs = coef_labels[c(1, 4:8)])
```

```{r}
#| label: tbl-lrt-rw
#| tbl-cap: "Likelihood ratio test for the region-by-wave random intercept (50:50 mixture of chi-square with 0 and 1 df)."
tibble::tibble(
  Model = c("M3a", "M3b: + region-wave"),
  `Log-likelihood` = f2(c(stats::logLik(m3a), stats::logLik(m3b))),
  AIC = f2(c(stats::AIC(m3a), stats::AIC(m3b))),
  BIC = f2(c(stats::BIC(m3a), stats::BIC(m3b))),
  LRT = c("", f2(lrt_region_wave$lrt)),
  `p (mixture)` = c("", sub("= ", "", pv(lrt_region_wave$p_value))),
  Singular = c(lme4::isSingular(m3a), m3b_singular)
) |>
  kable(format = "pipe")
```

```{r}
#| include: false
tau_rw <- as.numeric(lme4::VarCorr(m3b)$region_wave)
```

The region-by-wave variance is $\hat\tau_{rw} =$ `r f3(tau_rw)`, with likelihood ratio statistic `r f2(lrt_region_wave$lrt)` and mixture $p$ `r pv(lrt_region_wave$p_value)`; AIC is `r if (stats::AIC(m3b) < stats::AIC(m3a)) "lower" else "higher"` for M3b. `r if (keep_region_wave) "M3b proceeds: favorability shifts by region and wave in ways the separate dummies do not capture, and those shifts are now kept out of the violence coefficient." else "M3a proceeds: there is no clear evidence of region-wave shocks beyond the region and wave dummies."` `r if (keep_region_wave) paste0("The violence coefficient is ", f3(lme4::fixef(m3b)[['v_commune']]), " in M3b against ", f3(lme4::fixef(m3a)[['v_commune']]), " in M3a.") else ""`

## M4: final model with the survey design

```{r}
#| comment: ""
jtools::summ(m4, digits = 3, r.squared = FALSE, confint = TRUE, re.variance = "var")
```

```{r}
#| include: false
b4      <- lme4::fixef(m4)[["v_commune"]]
se4     <- violence_rows$std.error[violence_rows$model == "M4 (final)"]
ci4     <- b4 + c(-1, 1) * 1.96 * se4
or4     <- exp(b4)
or4_dbl <- exp(b4 * log(2))
```

**Violence.** Holding region, wave, respondent characteristics, and the design terms constant, a one-unit increase in $\log(1 + E)$ changes the log-odds of a favorable view by `r f3(b4)` (95% CI `r f3(ci4[1])` to `r f3(ci4[2])`) in a typical commune, village`r if (keep_region_wave) ", and region-wave" else ""`. As an odds ratio, the odds of a favorable view are multiplied by `r f2(or4)` each time $(1 + E)$ is multiplied by 2.72, and by `r f2(or4_dbl)` each time $(1 + E)$ doubles. On the probability scale, the average marginal effect is `r pp(ame$estimate)` percentage points per unit (95% CI `r pp(ame$conf.low)` to `r pp(ame$conf.high)`).

**Region and wave.** Region coefficients compare each region with Bamako, and wave coefficients compare waves 2 and 3 with wave 1, holding the other predictors constant. Positive values mean higher favorability than the reference.

**Variance components.** After the predictors, communes vary with $\hat\tau_{00} =$ `r f3(tau00_m3)` in M3 (residual ICC `r f3(icc_m3)`, down from `r f3(icc_m1)` in M1), and villages within communes with $\hat\tau_v =$ `r f3(tau_village)`.

**Informative sampling.** The log-weight coefficient is `r f3(b_weight[1])` (SE `r f3(b_weight[2])`, $p$ `r pv(b_weight[4])`). `r if (b_weight[4] < 0.05) "Respondents with larger weights differ in favorability beyond the predictors, so the unweighted estimates may not represent the population; compare with the design-based check below." else "There is no clear sign that the weights carry information about favorability beyond the predictors, which supports the unweighted multilevel estimates."`

## Cost of the design terms and the design-based check

```{r}
#| label: tbl-design
#| tbl-cap: "Violence coefficient (log-odds) across models. M2a to M4 are conditional (typical commune); the design-based check is population-averaged, so compare z, not the coefficient."
violence_rows |>
  dplyr::transmute(Model = model, Estimate = f3(estimate), SE = f3(std.error), z = f2(z)) |>
  kable(format = "pipe")
```

Adding the design terms changes the standard error of the violence coefficient by a factor of `r f2(design_cost$se_ratio)` and the coefficient by a factor of `r f2(design_cost$coef_ratio)` (M4 against M3). The design-based check, which uses the weights directly, gives $z =$ `r f2(violence_rows$z[4])` against `r f2(violence_rows$z[3])` for M4. `r if (abs(violence_rows$z[4]) < 0.7 * abs(violence_rows$z[3])) "The design-based evidence is noticeably weaker, so the weights and clustering matter here and the result should be reported with that caveat." else "The two lead to the same conclusion."`

## Predicted favorability

```{r}
#| label: tbl-predicted
#| tbl-cap: "M4: predicted percent with a favorable view if every respondent's commune had the given number of events, other characteristics as observed, typical commune and village (95% CI)."
predicted |>
  dplyr::transmute(`Events in commune` = events,
                   `Predicted % favorable` = pp(estimate),
                   `95% CI` = sprintf("%s to %s", pp(conf.low), pp(conf.high))) |>
  kable(format = "pipe", align = "rrr")
```

```{r}
#| label: fig-predicted
#| fig-cap: "M4: predicted percent favorable across event counts (95% CI). The x-axis is the event count; the model is linear in log(1 + events)."
#| fig-height: 3
ggplot(predicted_curve, aes(events, estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = pal[1]) +
  geom_line(linewidth = 0.8, color = pal[1]) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Events of violence against civilians in the commune", y = "Predicted % favorable")
```

## Sensitivity: usual level of violence versus change

```{r}
#| comment: ""
md_summs(m3, m3_mw, model.names = c("M3", "M3-MW"), coefs = coef_labels[1:3])
```

The between-commune coefficient is `r f3(lme4::fixef(m3_mw)[["v_between"]])` and the within-commune coefficient `r f3(lme4::fixef(m3_mw)[["v_within"]])`. The between part compares communes with persistently different violence; the within part uses only communes surveyed in more than one wave whose violence changed, so it is less precise.

# Diagnostics

```{r}
#| label: tbl-compare
#| tbl-cap: "Fit of the multilevel models (same respondents, maximum likelihood). ICC here is from performance::icc() and combines all random effects in the model (commune, and region-wave and village where included), so it differs from the commune-only ICC in the text. NA: not computed for a singular fit (M2b)."
performance::compare_performance(m1, m2a, m2b, m3a, m3b, m4, metrics = c("AIC", "BIC", "ICC")) |>
  tibble::as_tibble() |>
  dplyr::transmute(Model = c("M1", "M2a", "M2b", "M3a", "M3b", "M4"),
                   AIC = f2(AIC), BIC = f2(BIC), ICC = f3(ICC)) |>
  kable(format = "pipe")
```

```{r}
#| label: tbl-vif
#| tbl-cap: "Collinearity of the M4 predictors (generalized variance inflation factors)."
performance::check_collinearity(m4) |>
  tibble::as_tibble() |>
  dplyr::transmute(Term = Term, VIF = f2(VIF)) |>
  kable(format = "pipe")
```

```{r}
#| include: false
# lme4's own convergence messages (NULL = no warnings). performance::check_convergence()
# is not used: lme4 2.0 does not store the derivatives it needs and it returns FALSE.
m4_messages <- m4@optinfo$conv$lme4$messages
```

M4 `r if (performance::check_singularity(m4)) "is singular" else "is not singular"` and `r if (is.null(m4_messages)) "converged without warnings" else paste("returned:", toString(m4_messages), "(refitting with another optimizer is the usual next step)")`.

```{r}
#| label: fig-checks
#| fig-cap: "(A) M4 binned residuals: each point is the average residual in a bin of fitted values, with its 95% interval (gray); most intervals should cross zero (dashed line), with no pattern across fitted values. (B) M4 commune random effects against a normal distribution."
#| fig-height: 3
binned <- performance::binned_residuals(m4) |> tibble::as_tibble()

p_a <- ggplot(binned, aes(xbar, ybar)) +
  geom_hline(yintercept = 0, color = "gray40", linetype = "dashed") +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), color = "gray70", width = 0) +
  geom_point(color = pal[1], size = 0.9) +
  labs(title = "A. Binned residuals", x = "Fitted probability", y = "Average residual")

u_hat <- lme4::ranef(m4)$commune_id |> tibble::as_tibble() |> dplyr::rename(u = `(Intercept)`)
p_b <- ggplot(u_hat, aes(sample = u)) +
  stat_qq(color = pal[2], size = 0.7) + stat_qq_line(color = pal[3]) +
  labs(title = "B. Commune effects", x = "Normal quantiles", y = expression(hat(u)[0*c]))

p_a | p_b
```

# Maps

Gray communes were not surveyed in that wave: this is missing data, not zero.

```{r}
#| label: fig-map-violence
#| fig-cap: "Events of violence against civilians by commune and wave (surveyed communes only)."
#| fig-height: 3.9
map_violence
```

```{r}
#| label: fig-map-predicted
#| fig-cap: "Predicted percent favorable by commune and wave from M4, including each commune's and village's own random effects, averaged over its respondents with survey weights."
#| fig-height: 4.1
map_predicted
```

# How to explain the result

> Among people surveyed in the same region and survey round, those living in communes with more violence against civilians were `r if (b4 < 0) "less" else "more"` likely to view the leader favorably. If every commune had experienced no such events, about `r pp(predicted$estimate[predicted$events == 0])`% would be expected to hold a favorable view; with 5 events, about `r pp(predicted$estimate[predicted$events == 5])`%. This mostly compares different communes, so it shows an association, not proof that violence caused the change.

# Limitations

- **Mostly between communes.** Violence varies mainly between communes, so the estimate mainly compares different communes rather than the same commune over time. Stable differences the covariates miss (remoteness, local history) could drive part of it; the M3-MW sensitivity model shows how much the within-commune change agrees.
- **Random slope.** With little within-commune variation, the random slope test has low power. Not rejecting it does not show that the effect is the same everywhere.
- **Unweighted multilevel models.** The multilevel models do not use the survey weights directly. M4 checks for informative sampling with the log weight, and the design-based check shows whether weighting changes the conclusion.
- **Region-wide shocks.** Region and wave dummies enter separately; M3b tests for region-by-wave shocks with a random intercept, which pools thin cells toward the main effects.
- **Violence measure.** Only surveyed communes have event counts. ACLED undercounts events in remote areas, which most likely weakens the association.
- **Coverage.** Insecure areas are harder to survey, so estimates describe the population that could be reached.
- **Map placement.** Communes are joined to the boundaries by name key; names shared by two communes in one region cannot be placed and are counted in each caption.

# References

Bell, A., & Jones, K. (2015). Explaining fixed effects: Random effects modeling of time-series cross-sectional and panel data. *Political Science Research and Methods, 3*(1), 133-153.

Binder, D. A. (1983). On the variances of asymptotically normal estimators from complex surveys. *International Statistical Review, 51*(3), 279-292.

Gelman, A. (2007). Struggles with survey weighting and regression modeling. *Statistical Science, 22*(2), 153-164.

Lumley, T. (2010). *Complex surveys: A guide to analysis using R*. Wiley.

Mundlak, Y. (1978). On the pooling of time series and cross section data. *Econometrica, 46*(1), 69-85.

OCHA (United Nations Office for the Coordination of Humanitarian Affairs). Mali administrative boundaries, levels 0 to 3 (COD-AB). Source: Direction Nationale des Collectivités Territoriales (DNCT), release of 10 November 2021. Humanitarian Data Exchange.

Pfeffermann, D., Skinner, C. J., Holmes, D. J., Goldstein, H., & Rasbash, J. (1998). Weighting for unequal selection probabilities in multilevel models. *Journal of the Royal Statistical Society: Series B, 60*(1), 23-40.

Raleigh, C., Linke, A., Hegre, H., & Karlsen, J. (2010). Introducing ACLED: An armed conflict location and event dataset. *Journal of Peace Research, 47*(5), 651-660.

Zeger, S. L., Liang, K.-Y., & Albert, P. S. (1988). Models for longitudinal data: A generalized estimating equation approach. *Biometrics, 44*(4), 1049-1060.
