---
title: "Violence Against Civilians and Leader Favorability in Mali"
subtitle: "Multilevel models of cercle-level ACLED events and a three-wave household survey"
author: "Kevin Linares"
date: today
format:
  typst:
    papersize: us-letter
    margin:
      x: 1in
      y: 1in
    fontsize: 11pt
    section-numbering: "1.1"
    fig-format: png
    fig-dpi: 300
    fig-width: 6.5
    fig-height: 3.5
execute:
  echo: false
  warning: false
  message: false
---

```{r config}
# ---- Settings: change these, not the scripts ------------------------------------
# The data file is prepared upstream (see R/prep_data.R for the columns it needs).
cfg <- list(
  data_file = "data/survey_sim.rds",
  violence = "violence_civilian",          # cercle-wave event count used in every model
  violence_label = "Violence against civilians (ACLED)",
  covariates = c("female", "age", "ethnicity", "education"),      # M2a
  covariates_m2b = c("female", "age", "education"),               # M2b: an alternative set to compare
  final = "M2a",                           # covariate set carried into M3 (read the M2 comparison)
  cercle_wave = TRUE,                      # keep the cercle-by-wave intercept (read the M1 test)
  event_levels = c(0, 1, 5, 20),           # events per surveyed commune shown in predictions
  model_dir = "models",                    # fitted models are saved here
  refit = TRUE                             # set FALSE after the first run; TRUE after changes
)
```

```{r setup}
#| include: false
pacman::p_load(lme4, survey, marginaleffects, performance, broom.mixed, modelsummary, tinytable,
               sf, viridis, patchwork, knitr, tidyverse)

source("R/models.R")   # runs R/prep_data.R first
source("R/maps.R")

theme_set(theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank()))
pal <- viridis::viridis(3, end = 0.8)

# Formatting helpers
pct <- function(x) sprintf("%.1f", 100 * x)
f2 <- function(x) sprintf("%.2f", x)
f3 <- function(x) sprintf("%.3f", x)
pv <- function(p) if (p < 0.001) "< .001" else paste("=", sprintf("%.3f", p))

ci3 <- b3$estimate + c(-1.96, 1.96) * b3$std.error
se_wt <- sqrt(diag(as.matrix(stats::vcov(m3))))[["log_wt_c"]]
b_wt <- lme4::fixef(m3)[["log_wt_c"]]
n_cw <- dplyr::n_distinct(ad$cercle_wave)
```

# Question and data

Is violence against civilians in a respondent's cercle associated with how favorably they view the national leader, comparing people in the same region, urban or rural area, and survey wave? Violence is the number of ACLED events in the cercle during each wave's survey window. About eight in ten respondents view the leader favorably, so any association is expected to be modest. The model sequence and decision rules were fixed before the violence coefficient was examined.

The analysis file has `r format(nrow(ad), big.mark = ",")` respondents in `r dplyr::n_distinct(ad$commune_id)` communes and `r dplyr::n_distinct(ad$cercle_id)` cercles. Each wave is a separate sample: communes are drawn within strata (region by urban or rural), with eight respondents per commune. Violence takes one value per cercle and wave, so the violence effect rests on `r n_cw` cercle-waves, not on the number of respondents.

```{r}
#| label: tbl-waves
#| tbl-cap: "Sample by wave: design-weighted percent favorable (design-based SE), cercles, and median events per surveyed commune."
survey::svyby(~fav, ~wave, des, survey::svymean) |>
  tibble::as_tibble() |>
  dplyr::left_join(ad |> dplyr::summarise(n = dplyr::n(), cercles = dplyr::n_distinct(cercle_id),
                                          events = stats::median(events_per_commune), .by = wave),
                   by = "wave") |>
  dplyr::transmute(Wave = wave, Respondents = n, Cercles = cercles,
                   `% favorable` = pct(fav), SE = pct(se), `Median events per commune` = f2(events)) |>
  kable(format = "pipe")
```

# Models

## Violence measure

The ACLED count for cercle $k$ in wave $w$ is a sum over the communes surveyed there, so a cercle with more surveyed communes would look more violent. It is divided by the number of surveyed communes, $n_{kw}$:

$$
V_{kw} = \log\left(1 + \frac{E_{kw}}{n_{kw}}\right)
$$

Zero events gives 0; adding $\log 2 \approx 0.69$ doubles $(1 + E/n)$. Results are also given as predicted favorability at `r toString(cfg$event_levels)` events per surveyed commune. Respondent covariates $\mathbf{x}_i$: M2a uses `r toString(cfg$covariates)`; M2b uses `r toString(cfg$covariates_m2b)`. M3 carries the `r cfg$final` set.

## Model sequence

| Model | Adds | Compared by |
|---|---|---|
| M1a | cercle random intercept | |
| M1b | adds a cercle-by-wave random intercept | M1b vs M1a: likelihood ratio test with boundary correction, AIC, BIC |
| M2a | violence, wave, region, urban, covariate set A | |
| M2b | the same with covariate set B | M2b vs M2a: AIC, BIC; likelihood ratio test if one set contains the other |
| M3 | centered log survey weight | final model; design-based check with `svyglm` |
| Sensitivity | violence split into the cercle's usual level and its wave-to-wave change | |

All models are logistic multilevel models (`lme4::glmer`, maximum likelihood with the Laplace approximation), unweighted, on the same respondents. The cercle-by-wave intercept is kept if its test gives $p < .05$, AIC is lower, and the fit is not singular. Communes are not a random effect: violence does not vary within a cercle, and with eight respondents per commune the commune variance made models fail to converge. Communes remain the sampling units in the design-based check.

## Equations

Respondent $i$ in cercle $k$, wave $w$; $y_{ikw} = 1$ for a favorable view.

**M1a.** $\operatorname{logit} \Pr(y = 1) = \gamma_{0} + u_{k}$, with $u_k \sim N(0, \tau_k)$.

**M1b.** $\operatorname{logit} \Pr(y = 1) = \gamma_{0} + u_{k} + v_{kw}$, with $v_{kw} \sim N(0, \tau_{kw})$.

$u_k$ is the cercle's lasting level; $v_{kw}$ is how that cercle departs from it in one wave. Because everyone in a cercle-wave shares one violence value, $v_{kw}$ keeps the standard error of the violence effect honest. On the latent scale the individual variance is $\pi^2/3 \approx 3.29$, so

$$
\rho_{\text{cercle}} = \frac{\tau_k}{\tau_k + \tau_{kw} + \pi^2/3}, \qquad
\rho_{\text{cercle-wave}} = \frac{\tau_k + \tau_{kw}}{\tau_k + \tau_{kw} + \pi^2/3}
$$

The test is $H_0: \tau_{kw} = 0$; because a variance cannot be negative, the p-value is half of the $\chi^2_1$ p-value from `anova()`.

**M2a and M2b.**
$$
\operatorname{logit} \Pr(y = 1) = \gamma_0 + \beta V_{kw} + \sum_{w=2}^{3}\delta_w \text{Wave}_w + \sum_{r \ne \text{ref}} \alpha_r \text{Region}_r + \lambda\,\text{Urban} + \mathbf{x}_i'\boldsymbol\theta + u_k\,[+\,v_{kw}]
$$

The reference region is `r levels(ad$region)[1]` and the reference wave is wave 1. Region and urban are the survey strata. Wave enters as fixed effects (the national shift between waves); $v_{kw}$ captures each cercle's departure from that shift. M2a and M2b differ only in $\mathbf{x}_i$.

**M3 (final).** The chosen M2 plus $\phi \log \tilde w_i$, the centered log of the survey weight. `glmer` cannot use sampling weights, so the weight enters as a covariate; a clearly nonzero $\phi$ would mean people with larger weights differ in favorability beyond the predictors (informative sampling; Pfeffermann et al., 1998; Gelman, 2007).

**Design-based check.** The same fixed effects fit with `survey::svyglm`: weights rescaled within wave, strata within wave, communes as primary sampling units, Taylor-linearized standard errors (Binder, 1983). Its coefficients are population-averaged and smaller than the multilevel ones by roughly $\sqrt{1 + 0.346\,\tau}$ (Zeger, Liang & Albert, 1988), so it is compared by $z$.

**Sensitivity (usual level and change).** $V_{kw}$ is replaced by $\bar V_{k}$, the cercle's mean over the waves it was surveyed, and $V_{kw} - \bar V_{k}$ (Mundlak, 1978). The change coefficient asks whether a cercle is less favorable in a wave when its violence is above its usual level, so stable cercle differences cannot explain it.

# Results

## How much favorability clusters

```{r}
#| label: tbl-icc
#| tbl-cap: "Latent-scale intraclass correlations from the empty models. The commune-only model is shown for reference."
icc |>
  dplyr::transmute(Model = model, Level = level, ICC = f3(icc)) |>
  kable(format = "pipe")
```

## Model comparison

```{r}
#| label: tbl-compare
#| tbl-cap: "Fit of the model sequence (same respondents, maximum likelihood). Tests compare M1b with M1a and, when the covariate sets are nested, M2b with M2a."
m2_test <- if (is.null(test_m2)) "not nested: compare AIC, BIC" else
  sprintf("chi2(%d) = %.2f, p %s", abs(test_m2$Df[2]), test_m2$Chisq[2], pv(p_m2))
comparison |>
  dplyr::mutate(test = dplyr::case_when(
    model == "M1b" ~ sprintf("chi2(1) = %.2f, p %s (boundary)", test_m1$Chisq[2], pv(p_m1)),
    model == "M2b" ~ m2_test,
    .default = "")) |>
  dplyr::transmute(Model = model, `Log-lik.` = f2(log_lik), Parameters = n_par,
                   AIC = f2(aic), BIC = f2(bic), Test = test, Singular = singular) |>
  kable(format = "pipe")
```

M3 uses the `r cfg$final` covariates. M2a, M2b and M3 use `r if (cfg$cercle_wave) "cercle and cercle-by-wave random intercepts" else "a cercle random intercept only"`, as set in the configuration after reading this test.

## The violence effect

```{r}
#| label: tbl-violence
#| tbl-cap: "Violence coefficient (log-odds per unit of log(1 + events per commune)) across models. The design-based check is population-averaged: compare z, not the estimate."
violence_rows |>
  dplyr::transmute(Model = model, Estimate = f3(estimate), SE = f3(std.error), z = f2(z)) |>
  kable(format = "pipe")
```

In M3, a one-unit increase in violence changes the log-odds of a favorable view by `r f3(b3$estimate)` (95% CI `r f3(ci3[1])` to `r f3(ci3[2])`, $p$ `r pv(p_v)`), holding wave, region, urban, and respondent characteristics constant.

```{r}
#| label: tbl-predicted
#| tbl-cap: "M3: predicted percent favorable if every cercle had the given number of events per surveyed commune, everything else as observed, typical cercle (95% CI)."
predicted |>
  dplyr::transmute(`Events per commune` = events, `Predicted % favorable` = pct(estimate),
                   `95% CI` = paste(pct(conf.low), "to", pct(conf.high))) |>
  kable(format = "pipe", align = "rrr")
```

**What the data can and cannot rule out.** Going from 0 to 5 events per surveyed commune changes predicted favorability by `r pct(diff_0_5$estimate)` percentage points (95% CI `r pct(diff_0_5$conf.low)` to `r pct(diff_0_5$conf.high)`). Effects outside this interval are unlikely; if it includes zero, the data cannot distinguish the association from no association, which is not the same as showing no effect. The smallest 0-to-5 change this design would detect with 80% power at the 5% level is about `r pct(abs(mde_0_5))` percentage points.

**Design.** The log-weight coefficient is `r f3(b_wt)` (SE `r f3(se_wt)`); a value more than about two standard errors from zero would suggest informative sampling. The design-based check gives $z =$ `r f2(z_svy)` against `r f2(b3$z)` for M3.

```{r}
#| label: fig-by-variable
#| fig-cap: "M3: predicted percent favorable for each level of each factor in the model, setting everyone to that level and keeping everything else as observed (95% CI)."
#| fig-height: 7
predicted_by_var |>
  dplyr::mutate(level = forcats::fct_inorder(level)) |>
  ggplot(aes(estimate, level)) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = pal[1], size = 0.25) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Predicted % favorable", y = NULL)
```

## Sensitivity: usual level and change

```{r}
#| label: tbl-sensitivity
#| tbl-cap: "Violence split into the cercle's usual level and its wave-to-wave change (log-odds, SE)."
summary(m_split)$coefficients[c("v_cercle", "v_change"), ] |>
  tibble::as_tibble(rownames = "term") |>
  dplyr::transmute(Term = c("Cercle's usual level", "Change from usual level"), Estimate = f3(Estimate),
                   SE = f3(`Std. Error`), z = f2(`z value`)) |>
  kable(format = "pipe")
```

The change coefficient is the localized, over-time comparison; the usual-level coefficient compares cercles with persistently different violence.

# Diagnostics

```{r}
#| label: tbl-vif
#| tbl-cap: "Collinearity of the M3 predictors (generalized variance inflation factors)."
performance::check_collinearity(m3) |>
  tibble::as_tibble() |>
  dplyr::transmute(Term = Term, VIF = f2(VIF)) |>
  kable(format = "pipe")
```

M3 is `r if (lme4::isSingular(m3)) "singular" else "not singular"`. Convergence messages from lme4: `r if (is.null(m3@optinfo$conv$lme4$messages)) "none" else toString(m3@optinfo$conv$lme4$messages)`.

```{r}
#| label: fig-checks
#| fig-cap: "(A) M3 calibration: observed percent favorable in 20 bins of expected probability, against the range the model itself produces (gray, 95% of 200 simulated data sets with new cercle effects). (B) M3 cercle random effects against a normal distribution."
#| fig-height: 3
p_a <- ggplot(calibration, aes(expected, observed)) +
  geom_abline(color = "gray40", linetype = "dashed") +
  geom_errorbar(aes(ymin = sim_low, ymax = sim_high), color = "gray70", width = 0) +
  geom_point(color = pal[1], size = 0.9) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "A. Calibration", x = "Expected % favorable", y = "Observed % favorable")
u_hat <- lme4::ranef(m3)$cercle_id |> tibble::as_tibble() |> dplyr::rename(u = `(Intercept)`)
p_b <- ggplot(u_hat, aes(sample = u)) +
  stat_qq(color = pal[2], size = 0.8) + stat_qq_line(color = pal[3]) +
  labs(title = "B. Cercle effects", x = "Normal quantiles", y = "Estimated cercle effect")
p_a | p_b
```

# Maps

Gray cercles had no surveyed commune, so there is neither survey data nor an ACLED count for them.

```{r}
#| label: fig-map-violence
#| fig-cap: "Violence by cercle: events per surveyed commune, averaged over the waves the cercle was surveyed. Bins follow the data (zero, then quartiles of the positive values)."
#| fig-height: 5
map_violence
```

```{r}
#| label: fig-map-predicted
#| fig-cap: "M3 predicted percent favorable by cercle and wave, including each cercle's own effects, averaged over the respondents of its surveyed communes. It describes the surveyed communes, not the whole cercle."
#| fig-height: 4.1
map_predicted
```

# How to explain the result

> Among people surveyed in the same region, area, and survey round, predicted favorability toward the leader is about `r pct(predicted$estimate[1])`% where a cercle had no violence against civilians and about `r pct(predicted$estimate[predicted$events == 5])`% where it had 5 events per surveyed commune, a difference of `r pct(diff_0_5$estimate)` percentage points (95% CI `r pct(diff_0_5$conf.low)` to `r pct(diff_0_5$conf.high)`). This is an association, not proof that violence caused the change.

# Limitations

- **Few violence values.** Violence varies across `r n_cw` cercle-waves, which limits precision however many people were interviewed.
- **Surveyed communes only.** ACLED counts exist only where a commune was surveyed, and are scaled by the number surveyed; they describe the surveyed parts of each cercle, not the whole cercle.
- **Commune clustering not modeled.** Communes are not a random effect; the cercle and cercle-wave intercepts absorb most shared variation, and the design-based check keeps communes as sampling units.
- **Unweighted multilevel models.** The log-weight covariate and the design-based check guard against informative sampling but do not replace weighting.
- **Violence measure.** ACLED undercounts remote areas, which most likely weakens any association. Perpetrator is not available.
- **Sensitive outcome.** Favorability toward the leader may be over-reported where people fear consequences.
- **Coverage.** Insecure areas are harder to survey; estimates describe the population that could be reached.

# References

Binder, D. A. (1983). On the variances of asymptotically normal estimators from complex surveys. *International Statistical Review, 51*(3), 279-292.

Gelman, A. (2007). Struggles with survey weighting and regression modeling. *Statistical Science, 22*(2), 153-164.

Mundlak, Y. (1978). On the pooling of time series and cross section data. *Econometrica, 46*(1), 69-85.

OCHA (United Nations Office for the Coordination of Humanitarian Affairs). Mali administrative boundaries, levels 0 to 3 (COD-AB). Source: Direction Nationale des Collectivités Territoriales (DNCT), release of 10 November 2021. Humanitarian Data Exchange.

Pfeffermann, D., Skinner, C. J., Holmes, D. J., Goldstein, H., & Rasbash, J. (1998). Weighting for unequal selection probabilities in multilevel models. *Journal of the Royal Statistical Society: Series B, 60*(1), 23-40.

Raleigh, C., Linke, A., Hegre, H., & Karlsen, J. (2010). Introducing ACLED: An armed conflict location and event dataset. *Journal of Peace Research, 47*(5), 651-660.

Zeger, S. L., Liang, K.-Y., & Albert, P. S. (1988). Models for longitudinal data: A generalized estimating equation approach. *Biometrics, 44*(4), 1049-1060.

# Appendix: final model

```{r}
#| label: tbl-full
#| tbl-cap: "Final model M3: all fixed effects (log-odds [95% CI]) and random-effect standard deviations."
modelsummary::modelsummary(
  list(M3 = m3), output = "tinytable",
  estimate = "{estimate} [{conf.low}, {conf.high}]", statistic = NULL,
  gof_map = c("nobs", "aic", "bic"), fmt = 2
)
```
