---
title: "Local Violence and Leader Favorability in Mali"
subtitle: "Four models compared: region versus commune exposure, and three ways to handle clustering"
author: "Kevin Linares"
date: today
format:
  typst:
    papersize: us-letter
    margin:
      x: 1in
      y: 1in
    fontsize: 11pt
    section-numbering: "1.1"
    fig-format: png
    fig-dpi: 300
    fig-width: 6.5
    fig-height: 3.5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
# Everything is computed in plain R scripts; this report only presents it.
#   R/prep_data.R   edit this for real data
#   diagnostics.R   name checks and K, A, P, W
#   models.R        the four models and their violence effects
#   maps.R          map data
source("models.R")
source("maps.R")

pacman::p_load(patchwork, viridis, knitr)
theme_set(theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank()))
pal <- viridis::viridis(4, end = 0.85)
pp  <- function(x) sprintf("%.1f", 100 * x)   # proportion -> percentage points
```

# Question and approach

Is violence in a respondent's commune associated with how favorably they view the national leader? Violence is the number of ACLED events per commune and survey wave. Favorability comes from three survey waves, each with its own stratified, clustered sample and weights.

The report compares four models. Each step changes **one thing**, so any difference between neighbouring models has a single explanation:

| Model | Violence | Clustering | Weights | Shows |
|---|---|---|---|---|
| M1 | Region-wave | PSU (design) | Yes | Original specification |
| M2 | Commune-wave | None | No | Gain from finer exposure |
| M3 | Commune-wave | Random intercept | No | Multilevel alternative |
| M4 | Commune-wave | Commune (design) | Yes | **Main model** |

All four models share region and wave dummies and the same respondent covariates.

# Data checks

The analysis file has `r format(nrow(ad), big.mark = ",")` respondents (`r format(nrow(dat) - nrow(ad), big.mark = ",")` dropped for missing data) in `r dplyr::n_distinct(ad$commune_key)` communes across `r dplyr::n_distinct(ad$region)` regions and `r nlevels(ad$wave)` waves.

```{r}
#| label: tbl-names
#| tbl-cap: "Commune name matching. Communes are matched by name within region. A sampled commune missing from the violence file is assigned 0 events, so spelling mismatches would silently become zeros."
kable(name_checks, format = "pipe")
```

```{r}
#| label: tbl-diag
#| tbl-cap: "Diagnostics computed before any model is fit. K: exposure units. A: share of violence variation absorbed by region and wave dummies; whatever is absorbed cannot inform the violence estimate. P: PSUs per commune-wave. W: within-commune share of violence variation."
kable(diag_table, format = "pipe")
```

**Why Model 1 cannot work.** Region and wave dummies explain `r sprintf("%.0f", 100 * A_region)`% of the variation in region-wave violence. Model 1 must estimate the violence effect from the remaining `r sprintf("%.0f", 100 * (1 - A_region))`%, spread over only 24 region-wave values. Measured at the commune level, the same dummies explain only `r sprintf("%.0f", 100 * A_commune)`% of the variation, leaving most of it to inform the estimate. @fig-spread shows why: communes within the same region and wave experience very different levels of violence, and Model 1 gives all of them one number.

```{r}
#| label: fig-spread
#| fig-cap: "Commune violence within each region and wave. Each dot is a sampled commune; the bar is the region-wave average. Model 1 replaces every dot with one value per region-wave."
#| fig-height: 3.6
ggplot(cells, aes(region, v_commune, color = wave)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25, dodge.width = 0.75),
             alpha = 0.5, size = 0.9) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.6, linewidth = 0.3,
               position = position_dodge(width = 0.75)) +
  scale_color_viridis_d(end = 0.8) +
  labs(x = NULL, y = "log(1 + events) in commune", color = "Wave") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# Models

## Exposure

For commune $c$ in wave $w$, $E_{cw}$ is the ACLED event count and $V_{cw} = \log(1 + E_{cw})$. The log compresses a skewed count in which most commune-waves have few or no events. For Model 1, events are summed over all communes in the region: $V_{rw} = \log\left(1 + \sum_{c \in r} E_{cw}\right)$.

Models 2 to 4 split commune violence into two parts (Mundlak, 1978; Bell & Jones, 2015):

$$
\bar V_c = \frac{1}{T_c} \sum_{w \in S_c} V_{cw}, \qquad V^{\text{within}}_{cw} = V_{cw} - \bar V_c
$$

where $S_c$ is the set of waves in which commune $c$ was surveyed and $T_c$ is how many there are. $\bar V_c$ is the commune's usual level of violence; $V^{\text{within}}_{cw}$ is how much more or less violent that wave was than usual. A commune surveyed only once has $V^{\text{within}}_{cw} = 0$ and contributes nothing to the within estimate.

## Equations

All models share the same base, with region and wave dummies and respondent covariates $\mathbf{x}_i$ (female, age per decade, ethnicity, education, urban):

$$
\eta_i = \beta_0 + \sum_{r \neq \text{Bamako}} \alpha_r \text{Region}_r + \sum_{w=2}^{3} \gamma_w \text{Wave}_w + \mathbf{x}_i' \boldsymbol\beta
$$

**Model 1** (region exposure):
$$
\operatorname{logit} \Pr(y_i = 1) = \eta_i + \beta_R V_{rw}
$$

**Models 2 and 4** (commune exposure, within and between):
$$
\operatorname{logit} \Pr(y_i = 1) = \eta_i + \beta_W V^{\text{within}}_{cw} + \beta_B \bar V_c
$$

**Model 3** (same, plus a commune random intercept):
$$
\operatorname{logit} \Pr(y_i = 1) = \eta_i + \beta_W V^{\text{within}}_{cw} + \beta_B \bar V_c + u_c, \qquad u_c \sim N(0, \tau^2)
$$

- $\beta_W$ (**within**): when a commune is more violent than its own usual level, is favorability lower? Stable features of the commune (ethnic make-up, remoteness, history) cannot drive this estimate. This is the main quantity.
- $\beta_B$ (**between**): are people in usually more violent communes less favorable? This compares different communes, so it is more exposed to confounding.

## Estimation

- **M1 and M4** are survey-weighted logistic regressions (`survey::svyglm`). Weights are rescaled within each wave, $w^*_i = w_i \, n_w / \sum_{i \in w} w_i$, so no wave dominates. Strata are wave-specific. Standard errors use Taylor linearization, clustering on wave-specific **PSUs in M1** and wave-specific **communes in M4**. M4 clusters at the commune because that is where violence is assigned: everyone in a commune-wave shares one violence value.
- **M2** is an ordinary logistic regression (`glm`) that ignores weights and clustering, included only to show how much the clustering correction matters.
- **M3** is a multilevel logistic regression (`lme4::glmer`, Laplace approximation) with a commune random intercept and no weights.

**Reported effect.** All models report the **average marginal effect** (AME): the average change in the probability of a favorable view, in percentage points, per one-unit increase in log events. Because $\log(1 + E)$ rises by about 0.69 when events double, the table also gives the effect of a **doubling of events**. M3's AME is evaluated at $u_c = 0$ (a typical commune), so it is close to, but not exactly comparable with, the population-averaged AMEs of the other models.

# Results

```{r}
#| label: tbl-ame
#| tbl-cap: "Violence AMEs in percentage points, with 95% CIs. 'Per doubling' multiplies the AME by log(2), the approximate change in log(1 + events) when events double. The binary rows compare any event with none, so their doubling column is blank."
results |>
  dplyr::transmute(
    Model = model,
    Part  = part,
    `AME [95% CI]` = sprintf("%s [%s, %s]", pp(estimate), pp(conf.low), pp(conf.high)),
    SE = pp(std.error),
    `Per doubling` = dplyr::if_else(stringr::str_starts(term, "any"), "",
                                    pp(estimate * log(2)))) |>
  kable(format = "pipe")
```

**Main result (M4).** When a commune experiences more violence than its usual level, favorability changes by **`r pp(main_within$estimate)` points** per unit of log events (95% CI `r pp(main_within$conf.low)` to `r pp(main_within$conf.high)`), or about **`r pp(main_within$estimate * log(2))` points per doubling** of events. Comparing communes with each other, the between-commune AME is `r pp(main_between$estimate)` points (95% CI `r pp(main_between$conf.low)` to `r pp(main_between$conf.high)`).

**Model 1 for comparison.** The region-wave model gives `r pp(m1_effect$estimate)` points (95% CI `r pp(m1_effect$conf.low)` to `r pp(m1_effect$conf.high)`). Its interval is wide because, as @tbl-diag shows, almost all of its exposure variation is absorbed by the dummies.

**Effect of handling clustering.** M2 ignores clustering. The M4 standard error for the within effect is `r sprintf("%.1f", se_ratio)` times larger than in M2. The wider M4 interval is the honest one.

**Attitudes.** Adding local government approval, support for democracy, perceived violence, and stress (M4 + attitudes) shows whether the association runs partly through these attitudes. Because violence may itself change them, this row is descriptive, not the main estimate.

```{r}
#| label: fig-ame
#| fig-cap: "Violence AMEs across models (percentage points, 95% CI). M1 has a single region-wave effect; the commune models split it into within and between."
#| fig-height: 3.8
ggplot(results, aes(100 * estimate, forcats::fct_rev(model), color = part)) +
  geom_vline(xintercept = 0, color = "gray60") +
  geom_pointrange(aes(xmin = 100 * conf.low, xmax = 100 * conf.high),
                  position = position_dodge(width = 0.5), size = 0.3) +
  scale_color_viridis_d(end = 0.8) +
  labs(x = "AME (percentage points per unit of log events)", y = NULL, color = NULL) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-curve
#| fig-cap: "Main model (M4): weighted average predicted probability of a favorable view as a commune's violence moves above or below its own usual level. +0.69 on the x-axis is roughly a doubling of events."
#| fig-height: 3
curve_m4 <- marginaleffects::avg_predictions(
  m4,
  variables = list(v_within = seq(-1.5, 1.5, by = 0.25)),
  newdata = ad, wts = "wt_scaled")

ggplot(curve_m4, aes(v_within, estimate)) +
  geom_vline(xintercept = 0, color = "gray80") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = pal[2], alpha = 0.2) +
  geom_line(color = pal[2], linewidth = 1.1) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Violence relative to the commune's usual level (log events)", y = "Pr(favorable)")
```

# Maps

```{r}
#| label: fig-map-violence
#| fig-cap: "Exposure used by Models 2 to 4: ACLED events by commune and wave. Darker red means more events."
#| fig-height: 2.8
draw_map(map_violence) +
  facet_wrap(~ paste("Wave", wave)) +
  scale_fill_gradient(low = "#fff5f0", high = "#67000d",
                      breaks = log1p(c(0, 1, 3, 10, 30)), labels = c(0, 1, 3, 10, 30),
                      name = "Events")
```

```{r}
#| label: fig-map-m4
#| fig-cap: "Model 4 view: design-weighted percent favorable by commune and wave (sampled communes only; gray = not sampled that wave). Each value rests on few respondents, so read the map for broad patterns."
#| fig-height: 2.8
draw_map(map_fav) +
  facet_wrap(~ paste("Wave", wave)) +
  scale_fill_viridis_c(labels = scales::percent, limits = c(0, 1), name = "Favorable")
```

Median respondents per commune-wave on the map: `r stats::median(fav_commune_wave$n)`.

```{r}
#| label: fig-map-m3
#| fig-cap: "Model 3 view: commune random intercepts. Green communes are more favorable than the model predicts from region, wave, violence, and respondent characteristics; purple are less. Same-colour clusters spanning many communes would suggest a geographic factor the model is missing."
#| fig-height: 3.4
draw_map(map_u) +
  scale_fill_gradient2(low = "#440154", mid = "white", high = "#21918c", midpoint = 0,
                       name = "Commune\nintercept")
```

# Model diagnostics

Model 3 `r if (m3_singular) "is singular (commune variance near zero)" else if (is.null(m3_messages)) "converged without warnings" else paste("returned:", toString(m3_messages))`. Its commune variance is $\hat\tau^2 =$ `r sprintf("%.3f", tau2)` (ICC `r sprintf("%.3f", icc)` on the latent scale).

```{r}
#| label: fig-diagnostics
#| fig-cap: "(A) M4 within AME with each region dropped in turn; dashed line = full sample. (B) M4 weighted binned residuals; points should fall inside the dashed bounds with no pattern. (C) M4 calibration by decile. (D) M3 commune intercepts against a normal distribution."
#| fig-height: 5.6

# Fitted values from the main model, used by panels B and C
fit_m4 <- tibble::tibble(p = fitted(m4), y = ad$fav, w = ad$wt_scaled)

# A. Leave one region out
p_a <- ggplot(loo, aes(100 * estimate, dropped)) +
  geom_vline(xintercept = 100 * main_within$estimate, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = 0, color = "gray80") +
  geom_pointrange(aes(xmin = 100 * conf.low, xmax = 100 * conf.high), color = pal[1], size = 0.3) +
  labs(title = "A. Leave one region out", x = "Within AME (pp)", y = NULL)

# B. Binned residuals: 30 bins of fitted probability
binned <- fit_m4 |>
  dplyr::mutate(bin = dplyr::ntile(p, 30)) |>
  dplyr::summarise(fitted   = stats::weighted.mean(p, w),
                   residual = stats::weighted.mean(y - p, w),
                   n        = dplyr::n(),
                   .by = bin) |>
  dplyr::mutate(bound = 2 * sqrt(fitted * (1 - fitted) / n))

p_b <- ggplot(binned, aes(fitted, residual)) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_line(aes(y =  bound), linetype = "dashed", color = "gray50") +
  geom_line(aes(y = -bound), linetype = "dashed", color = "gray50") +
  geom_point(color = pal[2]) +
  labs(title = "B. Binned residuals (M4)", x = "Fitted", y = "Residual")

# C. Calibration by decile
calib <- fit_m4 |>
  dplyr::mutate(decile = dplyr::ntile(p, 10)) |>
  dplyr::summarise(predicted = stats::weighted.mean(p, w),
                   observed  = stats::weighted.mean(y, w),
                   .by = decile)

p_c <- ggplot(calib, aes(predicted, observed)) +
  geom_abline(linetype = "dashed", color = "gray50") +
  geom_point(color = pal[3], size = 2) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "C. Calibration (M4)", x = "Predicted", y = "Observed")

# D. Normal Q-Q plot of commune intercepts
p_d <- ggplot(u_hat, aes(sample = u)) +
  stat_qq(color = pal[1], size = 0.7) +
  stat_qq_line(color = pal[3]) +
  labs(title = "D. Commune intercepts (M3)", x = "Normal quantiles", y = "u_c")

(p_a | p_b) / (p_c | p_d)
```

# Explaining the result

A plain-language version for non-technical audiences:

> We compared the same communes across survey rounds. When a commune saw more violent events than it usually does, residents were about `r pp(abs(main_within$estimate * log(2)))` percentage points `r if (main_within$estimate < 0) "less" else "more"` likely to view the leader favorably for each doubling of events, after accounting for region, survey round, and who was interviewed. The range of plausible values is `r pp(main_within$conf.low * log(2))` to `r pp(main_within$conf.high * log(2))` points.

# Limitations

- **Name matching.** Communes are linked by name. A misspelled name silently becomes "0 events". @tbl-names reports names missing from the official commune list; fix those before interpreting results.
- **Communes surveyed once** add nothing to the within estimate (W in @tbl-diag).
- **Serial correlation.** Strata and clusters are wave-specific, so correlation within the same commune across waves is not reflected in M4's standard errors. A conservative check clusters on commune across waves with region as the stratum (see MODELING_NOTES.md).
- **Exposure measurement.** ACLED undercounts events in remote areas, which likely weakens the association. The event window must end before each wave's fieldwork.
- **Confounding.** The within estimate removes stable commune differences but not changes over time that coincide with violence (displacement, military operations, local economic shocks).
- **Coverage.** Insecure areas may be harder to survey, so estimates describe the reachable population.

# References

Bell, A., & Jones, K. (2015). Explaining fixed effects: Random effects modeling of time-series cross-sectional and panel data. *Political Science Research and Methods, 3*(1), 133-153.

Lumley, T. (2010). *Complex surveys: A guide to analysis using R*. Wiley.

Mundlak, Y. (1978). On the pooling of time series and cross section data. *Econometrica, 46*(1), 69-85.

Raleigh, C., Linke, A., Hegre, H., & Karlsen, J. (2010). Introducing ACLED: An armed conflict location and event dataset. *Journal of Peace Research, 47*(5), 651-660.

Runfola, D., et al. (2020). geoBoundaries: A global database of political administrative boundaries. *PLoS ONE, 15*(4), e0231866. https://doi.org/10.1371/journal.pone.0231866
