---
# =============================================================================
#  EDIT BLOCK 1 of 2: everything you change to swap LAPOP for the work project
#  lives here and in the LaTeX block below. Nothing further down needs edits.
# =============================================================================
title: "Segmenting Public Opinion Without Breaking the Survey Design"
subtitle: "Design-weighted latent class analysis with replicate variance estimation"
author: "Kevin Linares"
institute: "Public Opinion Research"
date: today
date-format: "D MMMM YYYY"

params:
  out_dir: "../output/mexico"          # folder holding the pipeline artifacts
  study: "AmericasBarometer 2023, Mexico"
  battery: "trust in national institutions"
  n_total: 1622                        # respondents in the released file
  n_analysis: 1430                     # complete cases entering estimation
  n_assigned: 1618                     # respondents receiving a segment
  n_strata: 4
  n_psu: 130
  design_df: 126                       # PSUs minus strata
  focus_demo: "prez_rating"            # demographic featured on the composition slide
  focus_demo_label: "Presidential job approval"
  focus_demo_levels: ["Very bad", "Bad", "Neither good nor bad", "Good", "Very good"]

format:
  beamer:
    aspectratio: 169
    theme: default
    slide-level: 2
    keep-tex: true
    # pdflatex uses only the fonts bundled with TeX itself. No system fonts,
    # no fontspec, nothing to install. Do not switch to xelatex or lualatex:
    # those resolve fonts from the operating system, which is the failure
    # you have been hitting.
    pdf-engine: pdflatex
    header-includes: |
      \usepackage{tikz}
      \usetikzlibrary{arrows.meta,positioning}
      \usepackage{booktabs}

      % ======================================================================
      %  EDIT BLOCK 2 of 2: the classification banner.
      %  Change the marking text and the bar color on these two lines only.
      %  Common choices: UNCLASSIFIED (006400 green), CUI (522398 purple),
      %  UNCLASSIFIED//FOUO (006400). Confirm against your marking guide.
      % ======================================================================
      \newcommand{\classification}{UNCLASSIFIED}
      \definecolor{classbg}{HTML}{006400}
      % ======================================================================

      \definecolor{vpurple}{HTML}{440154}
      \definecolor{vteal}{HTML}{21918C}
      \definecolor{vgreen}{HTML}{5EC962}

      \setbeamercolor{classbanner}{bg=classbg,fg=white}
      \setbeamercolor{frametitle}{fg=vpurple,bg=white}
      \setbeamercolor{title}{fg=vpurple}
      \setbeamercolor{structure}{fg=vteal}
      \setbeamertemplate{navigation symbols}{}

      % classification banner on the very top of every slide
      \setbeamertemplate{headline}{%
        \begin{beamercolorbox}[wd=\paperwidth,ht=2.0ex,dp=0.7ex,center]{classbanner}%
          \bfseries\small\classification%
        \end{beamercolorbox}%
      }
      % matching bottom banner with slide number. If your marking guide wants
      % a top banner only, delete this \setbeamertemplate{footline} block.
      \setbeamertemplate{footline}{%
        \begin{beamercolorbox}[wd=\paperwidth,ht=2.0ex,dp=0.7ex]{classbanner}%
          \hspace*{2ex}\bfseries\small\insertframenumber\hfill\classification\hfill\hspace*{2ex}%
        \end{beamercolorbox}%
      }

      \newcommand{\kicker}[1]{\par\noindent{\footnotesize\color{vteal}\textbf{\MakeUppercase{#1}}}\par\vspace{0.4em}}

      % Uncomment to project speaker notes on a second screen:
      % \setbeameroption{show notes on second screen=right}

# =============================================================================
#  CUT GUIDE. Every slide is tagged [CORE], [OPTIONAL] or [BACKUP] in a comment.
#  Delete the [OPTIONAL] slides for a 15 minute talk; keep all for ~27 minutes.
#  Cut order: prior typologies, local independence, replicate detail,
#  LLM guardrails. Keep the BCH slide if anyone may ask how it works.
# =============================================================================
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(arrow)
library(knitr)
library(viridisLite)
library(tidyverse)   # loaded last so dplyr verbs win

out_dir        <- params$out_dir
enum           <- arrow::read_parquet(file.path(out_dir, "enumeration.parquet"))
ci_tbl         <- arrow::read_parquet(file.path(out_dir, "ci_tbl.parquet"))
bvr_tbl        <- arrow::read_parquet(file.path(out_dir, "bvr_tbl.parquet"))
validation     <- arrow::read_parquet(file.path(out_dir, "validation.parquet"))
fit_w          <- readRDS(file.path(out_dir, "fit_weighted.rds"))
pl_aligned     <- readRDS(file.path(out_dir, "polca_aligned.rds"))
segment_labels <- readr::read_csv(file.path(out_dir, "segment_labels.csv"),
                                  show_col_types = FALSE)
dom            <- readr::read_csv(file.path(out_dir, "segment_demographics.csv"),
                                  show_col_types = FALSE)

K_star <- length(fit_w$pi)

lab_lev <- stringr::str_wrap(paste0("S", segment_labels$K, ": ",
                                    segment_labels$Label), 20)
lab_of  <- function(k) factor(lab_lev[k], levels = lab_lev)

theme_slide <- function() {
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          legend.position  = "bottom",
          legend.text      = element_text(size = 7),
          legend.key.size  = unit(0.4, "cm"),
          plot.title       = element_blank(),
          strip.text       = element_text(face = "bold"))
}
```
## Roadmap {.shrink}

<!-- [CORE] 2 -->
\kicker{the plan}

\vspace{0.4em}

\begin{columns}[T]
\begin{column}{0.52\textwidth}
\begin{enumerate}\itemsep2pt
  \item What JPSM trained me to look for
  \item Where this work sits in Total Survey Error
  \item The problem: trust is a pattern, not a score
  \item Why this method, and why not k-means
  \item The method, in four moves
  \item What we found
  \item Naming segments: an anchor and adjust approach
  \item What I would bring to our work
\end{enumerate}
\end{column}
\begin{column}{0.44\textwidth}
\small
\textbf{The claim in one line}

\vspace{0.4em}

Segmentation of survey data is usually done as if the sample were a simple random sample. It is not, and fixing that changes both the estimates and what we are entitled to say about them.
\end{column}
\end{columns}

\note{Set expectations: a methods talk with a substantive payoff, not a software demo. The one-liner on the right is the thesis; everything else supports it.}

## What I bring back from JPSM {.shrink}

<!-- [CORE] 3 -->
\kicker{where i trained}

\small

**Joint Program in Survey Methodology** (Maryland, Michigan SRC, Westat). MS in Survey and Data Science, MA in Psychology.

\vspace{0.5em}

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\textbf{Coursework that shows up here}
\begin{itemize}\itemsep2pt
  \item Design-based sampling and estimation
  \item Latent variable measurement models
  \item Replication variance methods
  \item Responsible LLM integration
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{The two portable habits}
\begin{itemize}\itemsep2pt
  \item Name where error can enter, then decide where to spend effort
  \item Everything in code, rendered from source, so a result can be audited a year later
\end{itemize}
\end{column}
\end{columns}

\vspace{0.5em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.93\textwidth}{\centering
The degree is organised around one framework, \textbf{Total Survey Error}. It is the spine of the program and the lens I apply before choosing any method.}}
\end{center}

\note{Forty seconds. Credibility, not biography. The second habit is what makes the results slide defensible later.}

## Total Survey Error: the framework the program is built on {.shrink}

<!-- [CORE] 4 -->
\kicker{groves and colleagues}

\begin{center}
\resizebox{!}{5.6cm}{%
\begin{tikzpicture}[
  b/.style={draw=gray!55, rounded corners=1.5pt, minimum width=27mm, minimum height=6mm, align=center, font=\scriptsize},
  h/.style={b, draw=vteal, fill=vteal!25, font=\scriptsize\bfseries},
  hd/.style={font=\small\bfseries, text=vpurple},
  e/.style={font=\scriptsize\itshape, text=vteal},
  a/.style={-{Latex[length=1.6mm]}, gray!70}]
\node[hd] at (0,0.95)   {MEASUREMENT (the what)};
\node[hd] at (6.4,0.95) {REPRESENTATION (the who)};
\node[b] (c1) at (0,0)     {Construct};
\node[b] (c2) at (0,-1.35) {Measurement};
\node[b] (c3) at (0,-2.7)  {Response};
\node[b] (c4) at (0,-4.05) {Edited Response};
\draw[a] (c1)--(c2) node[e,midway,right,xshift=2mm]{validity};
\draw[a] (c2)--(c3) node[e,midway,right,xshift=2mm]{measurement error};
\draw[a] (c3)--(c4) node[e,midway,right,xshift=2mm]{processing error};
\node[b] (r1) at (6.4,0)     {Target Population};
\node[b] (r2) at (6.4,-1.35) {Sampling Frame};
\node[b] (r3) at (6.4,-2.7)  {Sample};
\node[b] (r4) at (6.4,-4.05) {Respondents};
\node[b] (r5) at (6.4,-5.4)  {Post-survey Adjustments};
\draw[a] (r1)--(r2) node[e,midway,right,xshift=2mm]{coverage error};
\draw[a] (r2)--(r3) node[e,midway,right,xshift=2mm]{sampling error};
\draw[a] (r3)--(r4) node[e,midway,right,xshift=2mm]{nonresponse error};
\draw[a] (r4)--(r5) node[e,midway,right,xshift=2mm]{adjustment error};
\node[h] (s) at (3.2,-6.7) {Survey Statistic};
\draw[a] (c4) -- (s); \draw[a] (r5) -- (s);
\end{tikzpicture}}
\end{center}

\note{Walk left to right down each arm. The labels beside the arrows are the named error sources. Both arms converge on the survey statistic: accuracy is what survives to the final number. As analysts we usually receive the data at the far bottom, after every one of these steps has already happened.}

## Accuracy is the inverse of mean squared error {.shrink}

<!-- [CORE] 5 -->
\kicker{the core of the framework}

\small
$$\text{MSE} \;=\; \text{Bias}^2 \;+\; \text{Variance}$$

\footnotesize
\vspace{0.2em}
Expanded (Biemer and Lyberg), with each component drawn from one arm or the other:

\vspace{0.4em}
$\text{Bias}^2 = \big(\underbrace{\text{specification}}_{\text{meas}} + \underbrace{\text{coverage}}_{\text{rep}} + \underbrace{\text{nonresponse}}_{\text{rep}} + \underbrace{\text{measurement}}_{\text{meas}} + \underbrace{\text{processing}}_{\text{meas}}\big)^2$

\vspace{0.5em}
$\text{Variance} = \underbrace{\text{sampling}}_{\text{rep}} + \underbrace{\text{measurement}}_{\text{meas}} + \underbrace{\text{processing}}_{\text{meas}}$

\vspace{0.6em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.9\textwidth}{\centering\footnotesize
Bias and variance do not map onto the two arms. Sampling is representation-side variance, interviewer effects are measurement-side variance. Both arms feed both terms.}}
\end{center}

\vspace{0.3em}
\footnotesize
MSE is never zero. Even a census carries measurement and processing error. The practical question is never how to eliminate error, it is which component is largest and most tractable under a budget.

\note{The common mistake this corrects: people assume measurement error is bias and sampling error is variance. Both arms contribute to both. That is why chasing a single number like the response rate is a poor strategy.}

## Most segmentation work uses only half the map {.shrink}

<!-- [CORE] 6 -->
\kicker{where this project sits}

\footnotesize

\begin{columns}[T]
\begin{column}{0.49\textwidth}
\textbf{\color{vpurple}Measurement arm}

\vspace{0.3em}
\begin{itemize}\itemsep3pt
  \item \textbf{Validity.} Is institutional trust one thing? A summed scale assumes yes. The model tests it.
  \item \textbf{Measurement error.} Each item is a fallible indicator with its own response probabilities.
  \item \textbf{Processing error.} Assigning a respondent to a segment is itself an error source. We correct it rather than ignore it.
\end{itemize}
\end{column}
\begin{column}{0.49\textwidth}
\textbf{\color{vpurple}Representation arm}

\vspace{0.3em}
\begin{itemize}\itemsep3pt
  \item \textbf{Sampling error.} Variance comes from the strata and clusters actually used, not a textbook formula.
  \item \textbf{Adjustment error.} Base weights enter the estimation itself, not a post hoc reweighting of results.
  \item \textbf{Nonresponse.} Item-partial respondents are scored from what they did answer, above an evidence floor.
\end{itemize}
\end{column}
\end{columns}

\vspace{0.6em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.93\textwidth}{\centering
Typical practice segments on the measurement arm and reports sample shares. \\
Connecting both arms is what turns a segment share into a \textbf{population estimate with defensible uncertainty}.}}
\end{center}

\note{The slide the audience should remember. Ask them: when you last clustered survey data, did the cluster sizes come with intervals that used the design? The answer is almost always no.}

## This is an established method in our field {.shrink}

<!-- [CORE] 7 -->
\kicker{lca in public opinion research}

\footnotesize

\begin{itemize}\itemsep4pt
  \item \textbf{Survey measurement error.} Breen (2000, \textit{BJPS}) used LCA to show why surveys underestimate support for extreme parties.
  \item \textbf{Cross-national attitudes.} Blaydes and Linzer (2008, \textit{World Politics}) characterized respondents across eighteen countries on gender norms, political Islam, and religiosity.
  \item \textbf{Latin America.} Alvarez, Levin and N\'u\~nez (2017, \textit{Journal of Politics}) identified four faces of political participation in Argentina.
  \item \textbf{Participation repertoires.} Oser, Hooghe and Marien (2013, \textit{PRQ}) tested whether online participation is distinct from offline.
  \item \textbf{National identity.} Bonikowski and DiMaggio (2016, \textit{ASR}) recovered varieties of American popular nationalism.
  \item \textbf{Complex survey designs.} Patterson, Dayton and Graubard (2002, \textit{JASA}) fit LCA to NHANES dietary data under the sample design.
\end{itemize}

\vspace{0.5em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.92\textwidth}{\centering
The method is not exotic. What is still uncommon is carrying the \textbf{sample design} all the way through it.}}
\end{center}

\note{This slide preempts the "is this a real method" question. Name Breen first: it is the one that frames LCA as a survey error tool rather than a clustering tool.}

## And specifically for institutional trust {.shrink}

<!-- [OPTIONAL] 8 -->
\kicker{prior typologies}

\small

Trust batteries have been segmented before, in several countries and data sources:

\vspace{0.5em}

\begin{itemize}\itemsep4pt
  \item An Australian national survey recovered four trust classes, from the largely distrusting through the unsure to the broadly trusting.
  \item Work on Italian regional data used a latent class measure of trust in others and trust in institutions to map a divide more complex than the usual north and south split.
  \item German civic institution data yielded profiles that separate confidence in courts, police, and media from confidence in political institutions.
\end{itemize}

\vspace{0.6em}

Two findings recur, and both appear in our results: **trust is not one dimension**, and a substantial group trusts some institutions while rejecting others.

\vspace{0.4em}

\begin{center}
\small What these studies generally do not report is a design-based interval on any of those group sizes.
\end{center}

\note{Optional depth. The value is showing that our five-segment structure is consistent with prior work rather than an artifact. The closing line is the contribution, stated without overclaiming.}

## The problem: trust is a pattern, not a score {.shrink}

<!-- [CORE] 9 -->
\kicker{motivation}

\small

The standard move with a trust battery is to sum or average it into a scale.

\vspace{0.4em}

That assumes everyone who scores a 4 got there the same way. In `r params$study` they did not:

\vspace{0.5em}

\begin{columns}[T]
\begin{column}{0.47\textwidth}
\colorbox{vteal!12}{\parbox{0.95\textwidth}{\small
\textbf{Two respondents, same average}

\vspace{0.3em}
One trusts the courts and the auditor, distrusts the president.

\vspace{0.3em}
One trusts the president and the military, distrusts parties and police.}}
\end{column}
\begin{column}{0.47\textwidth}
\small
A scale calls them identical.

\vspace{0.5em}
They are not the same people, they do not respond to the same messages, and they will not move the same way.

\vspace{0.5em}
\textbf{Latent class analysis recovers the configuration instead of averaging it away.}
\end{column}
\end{columns}

\note{The hook for this audience. Everyone has summed a battery. Do not attack the practice, show what it hides.}

## Why not k-means? {.shrink}

<!-- [CORE] 10 -->
\kicker{algorithm versus model}

\footnotesize

| | k-means | Latent class analysis |
|:--|:--|:--|
| What it is | An algorithm that partitions this sample | A statistical model of the population |
| Requires | A distance you choose, usually Euclidean | A likelihood the data imply |
| Number of groups | Elbow or silhouette, both heuristics | AIC, BIC, entropy: comparable fit |
| Assignment | Hard. Everyone gets exactly one label | A posterior probability per person |
| Survey weights | No principled route | Enter the likelihood directly |
| What you can report | Shares in this file | Population shares with intervals |

\vspace{0.5em}

\begin{columns}[T]
\begin{column}{0.62\textwidth}
\scriptsize
\textbf{Three specific failures on these data}
\begin{itemize}\itemsep1pt
  \item Euclidean distance assumes the gap from 1 to 2 equals the gap from 6 to 7. Nothing in a trust item justifies that.
  \item Hard assignment cannot measure its own error, so the classification correction on slide 17 is impossible in principle.
  \item A cluster share describes the respondents in the file, not the country.
\end{itemize}
\end{column}
\begin{column}{0.35\textwidth}
\vspace{0.6em}
\colorbox{vteal!15}{\parbox{0.95\textwidth}{\scriptsize
k-means is a fine exploratory tool. The objection is to reporting a cluster share as a population estimate.}}
\end{column}
\end{columns}

\note{Do not read the table. Land two things: distance on ordinal answers is arbitrary, and k-means cannot quantify its own assignment error so the resulting bias is invisible. Close with the fairness line so the people who run k-means weekly stay with you.}

## What the model estimates {.shrink}

<!-- [CORE] 11 -->
\kicker{method, part one}

\small
Each respondent belongs to one unobserved segment. Within a segment, items are answered independently:

$$
P(\mathbf{y}_i) \;=\; \sum_{k=1}^{K} \underbrace{\pi_k}_{\text{segment share}} \prod_{j=1}^{J} \underbrace{\rho_{j\,y_{ij}\,k}}_{\substack{\text{P(answer} \mid \text{segment)}}}
$$

\vspace{0.4em}

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\textbf{What we get out}
\begin{itemize}\itemsep2pt
  \item How large each segment is in the population
  \item The full response profile that defines it
  \item Each respondent's probability of belonging
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{What it costs}
\begin{itemize}\itemsep2pt
  \item The number of segments is a judgment, not an output
  \item Local independence is an assumption we have to check
\end{itemize}
\end{column}
\end{columns}

\note{Do not read the equation. Point at the two braces: a share and a response profile. Everything else is about estimating those two honestly under a complex design.}

## The assumption, and how we check it {.shrink}

<!-- [OPTIONAL] 12 -->
\kicker{local independence}

\small

The model says: once you know a respondent's segment, their answers to any two items are unrelated. All the correlation between items is explained by segment membership.

\vspace{0.5em}

If two items stay correlated inside a segment, the model is missing something. We rank every pair by that residual association:

```{r}
#| label: tbl-bvr
bvr_tbl |>
  dplyr::arrange(desc(bvr)) |>
  dplyr::slice_head(n = 4) |>
  dplyr::transmute(`Item A` = item_a, `Item B` = item_b,
                   `Residual association` = sprintf("%.1f", bvr)) |>
  knitr::kable(align = "llr")
```

\footnotesize
The top pairs are institutionally adjacent, two audit and prosecution bodies, and two national bodies. We report this as a ranked caution, not a test: under a weighted likelihood there is no valid reference distribution, so an absolute cutoff borrowed from simple random sampling would be misleading.

\note{Optional but strong with a methods-literate audience. The honest note is that we decline to compute a p-value we would not be able to defend.}

## Why this needed building {.shrink}

<!-- [CORE] 13 -->
\kicker{the tooling gap}

\footnotesize

| Tool | Weighted likelihood | Design-based variance | Open |
|:--|:--:|:--:|:--:|
| `poLCA` (R) | no | no | yes |
| `survey` (R) | n/a, no mixture models | yes | yes |
| Mplus, Latent GOLD | yes | linearization | no |
| `baysc` (R) | yes, Bayesian | sandwich adjustment | yes |
| **This pipeline** | **yes** | **replicate (JKn)** | **yes** |

\vspace{0.7em}

\small
No open frequentist implementation carried a weighted likelihood, design-consistent variance, classification-error correction, and a labelled analyst data product end to end. So the estimation core, the variance machinery, and the profiling correction are custom, and every one of them is validated against a reference implementation.

\note{Be fair to the alternatives. Mplus and Latent GOLD do this well and cost money; baysc is an excellent Bayesian counterpart. The gap is specific, not a claim that nobody solved segmentation.}

## The method, in four moves {.shrink}

<!-- [CORE] 14 -->
\kicker{method, part two}

\footnotesize

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\textbf{1. Weight the likelihood}

Every respondent contributes in proportion to the population they represent.
$$\ell_w = \sum_i w_i \log P(\mathbf{y}_i)$$

\vspace{0.5em}
\textbf{2. Calibrate the criteria}

Base weights do not sum to the sample size, so fit statistics are rescaled before being compared to a sample-size penalty.
\end{column}
\begin{column}{0.48\textwidth}
\textbf{3. Replicate the variance}

A weighted likelihood gives no usable information matrix. So the whole model is refit on `r params$n_psu` jackknife replicates built from the real strata and clusters.

\vspace{0.5em}
\textbf{4. Correct the classification}

Assigning people to segments is imperfect. The BCH three-step correction removes the resulting bias before any demographic comparison.
\end{column}
\end{columns}

\vspace{0.5em}
\begin{center}
\small Design: `r params$n_strata` strata, `r params$n_psu` PSUs, `r params$design_df` degrees of freedom for every interval in this deck.
\end{center}

\note{Move three is the expensive one: refitting 130 times. Move four is the one most analysts have never heard of, and it is where the biggest silent bias lives.}

## Move 3: where the uncertainty comes from {.shrink}

<!-- [OPTIONAL] 15 -->
\kicker{replicate variance}

\small

For an ordinary model, standard errors come from the curvature of the likelihood. A **weighted** likelihood is not a real likelihood, so that shortcut is not available and using it anyway understates uncertainty.

\vspace{0.5em}

Instead we rebuild the estimate many times:

\begin{enumerate}\itemsep2pt
  \item Drop one primary sampling unit, reweight its stratum-mates to compensate
  \item Refit the entire model on those weights
  \item Repeat for all `r params$n_psu` sampling units
  \item The spread across those refits is the design-based variance
\end{enumerate}

\vspace{0.5em}

\footnotesize
Two practical details. Each replicate refit is warm-started from the full-sample solution, so the variance measures design uncertainty rather than optimizer wandering. And because segment 1 in one replicate need not be segment 1 in another, replicates are matched back to the full-sample solution before anything is compared.

\note{The label-switching point is worth making: a naive implementation silently differences segment 1 against segment 3 and reports enormous standard errors. This is the kind of bug that does not announce itself.}

## Move 4: correcting the classification {.shrink}

<!-- [OPTIONAL] 16 -->
\kicker{the bch three step}

\footnotesize

Respondents are assigned to their most likely segment, but that assignment is imperfect. First we measure exactly how imperfect, using weighted posteriors. Rows are true segments, columns are assigned segments:

$$
\mathbf{D} = \begin{bmatrix} 0.94 & 0.04 & 0.02 \\ 0.05 & 0.91 & 0.04 \\ 0.03 & 0.06 & 0.91 \end{bmatrix}
\qquad D_{ks} = P(\text{assigned } s \mid \text{truly } k)
$$

\vspace{0.4em}

Row one reads: of everyone truly in segment 1, ninety four percent get labelled segment 1 and six percent leak elsewhere. A perfect classifier would be the identity matrix.

\vspace{0.4em}

Then we invert it. Each respondent's hard label is replaced by the corresponding row of $\mathbf{D}^{-1}$, which un-mixes the leakage. Cross-tabs built on those corrected values are unbiased for true segment composition.

\vspace{0.4em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.9\textwidth}{\centering
Without this step, every demographic difference is compressed toward the average, and the compression is invisible.}}
\end{center}

\note{If asked about the negative values: inverting a near-identity matrix produces small negative entries, which is the correction removing contamination. Each respondent's corrected row still sums to one, so shares still add to 100 percent.}

## Choosing the number of segments {.shrink}

<!-- [CORE] 17 -->
\kicker{results, part one}

```{r}
#| label: fig-enum
#| fig-width: 8
#| fig-height: 2.9
#| out-width: 98%
enum |>
  dplyr::select(K, BIC, entropy) |>
  tidyr::pivot_longer(-K, names_to = "criterion", values_to = "value") |>
  dplyr::mutate(criterion = dplyr::recode(criterion,
                                          BIC = "Weighted BIC (lower is better)",
                                          entropy = "Classification sharpness")) |>
  ggplot(aes(K, value)) +
  geom_line(linewidth = 0.7, color = viridisLite::viridis(1, begin = 0.4)) +
  geom_point(size = 2.2, color = viridisLite::viridis(1, begin = 0.4)) +
  facet_wrap(~ criterion, scales = "free_y") +
  scale_x_continuous(breaks = enum$K) +
  labs(x = "Number of segments (K)", y = NULL) +
  theme_slide()
```

\footnotesize
The criterion is nearly flat between five and six segments. We took five, on parsimony, sharper classification, and interpretability. **The software never picks this.** The choice is recorded in configuration and the full evidence is reported so a reader can weigh the alternative.

\note{A credibility slide. Say plainly that an automated choice here is the first thing a reviewer attacks, and that the flat region is reported rather than hidden.}

## Five segments, and two of them are configurational {.shrink}

<!-- [CORE] 18 -->
\kicker{results, part two}

```{r}
#| label: fig-profiles
#| fig-width: 8
#| fig-height: 3.4
#| out-width: 98%
prof <- ci_tbl |>
  dplyr::filter(kind == "rho") |>
  dplyr::group_by(item, class) |>
  dplyr::summarise(mean_resp = sum(category * estimate), .groups = "drop")

item_order <- prof |>
  dplyr::group_by(item) |>
  dplyr::summarise(m = mean(mean_resp), .groups = "drop") |>
  dplyr::arrange(m) |>
  dplyr::pull(item)

prof |>
  dplyr::mutate(item = factor(item, levels = item_order),
                Segment = lab_of(class)) |>
  ggplot(aes(mean_resp, item, color = Segment, group = Segment)) +
  geom_line(orientation = "y", linewidth = 0.5, alpha = 0.65) +
  geom_point(size = 2) +
  scale_color_viridis_d(begin = 0.05, end = 0.85) +
  scale_x_continuous(limits = c(1, 7), breaks = 1:7) +
  labs(x = "Expected response (1 = no trust, 7 = high trust)", y = NULL,
       color = NULL) +
  theme_slide() +
  guides(color = guide_legend(nrow = 1))
```

\footnotesize
Levels separate three of the segments. The other two cross: same overall trust, opposite shape.

\note{Trace one crossover with your finger: the two middle segments swap order between political parties and the president. That crossing is the entire argument against a scale.}

## Population size of each segment {.shrink}

<!-- [CORE] 19 -->
\kicker{results, part three}

```{r}
#| label: fig-prev
#| fig-width: 8
#| fig-height: 2.9
#| out-width: 98%
ci_tbl |>
  dplyr::filter(kind == "pi") |>
  dplyr::mutate(Segment = lab_of(class),
                unweighted = pl_aligned$pi[class]) |>
  ggplot(aes(estimate, Segment)) +
  geom_pointrange(aes(xmin = lo, xmax = hi),
                  color = viridisLite::viridis(1, begin = 0.35), linewidth = 0.7) +
  geom_point(aes(x = unweighted), shape = 5, size = 2.4, color = "grey35") +
  scale_x_continuous(limits = c(0, NA)) +
  labs(x = "Share of the population (95% jackknife interval)", y = NULL) +
  theme_slide()
```

\footnotesize
Filled points are design-weighted estimates with jackknife intervals. Open diamonds are what an unweighted analysis would have reported. In this file the weights are nearly constant, so the two agree closely, and that is worth saying out loud. **The intervals are the deliverable**: without them, a three point difference between two segments looks like a finding.

\note{Honesty pays. The weighting correction is small in this dataset; do not oversell it. The variance correction always matters and is the part people skip.}

## Who belongs to which segment {.shrink}

<!-- [CORE] 20 -->
\kicker{results, part four}

```{r}
#| label: fig-demo
#| fig-width: 8
#| fig-height: 3.2
#| out-width: 98%
dom |>
  dplyr::filter(demographic == params$focus_demo) |>
  dplyr::mutate(level = factor(level, levels = params$focus_demo_levels),
                Segment = lab_of(segment)) |>
  ggplot(aes(prevalence, level, color = Segment)) +
  geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.32, linewidth = 0.45,
                  position = position_dodge(width = 0.65)) +
  scale_color_viridis_d(begin = 0.05, end = 0.85) +
  labs(x = "Share of that group belonging to each segment", y = NULL, color = NULL) +
  theme_slide() +
  guides(color = guide_legend(nrow = 1))
```

\footnotesize
`r params$focus_demo_label`, corrected for classification error. Composition is strongly ordered, and the intervals tell you which contrasts are real.

\note{Name one clean contrast and one that is not significant. Analysts trust a speaker who names the null results. These are BCH corrected: uncorrected cross-tabs would compress every difference toward the middle.}

## Why you can believe the numbers {.shrink}

<!-- [CORE] 21 -->
\kicker{trust checks}

\small

```{r}
#| label: val
#| results: asis
cat(sprintf(paste0("**Validation.** Run at equal weights, this estimator and an ",
                   "independent published implementation must maximize the same ",
                   "function. They agree to %.5f on every segment share and ",
                   "%.5f on every response probability.\n\n"),
            validation$max_pi_gap, validation$max_rho_gap))
```

\vspace{0.2em}

\begin{itemize}\itemsep3pt
  \item \textbf{One verdict, one place.} The validation result is written to disk by the estimation script and read by every downstream document, so no two outputs can disagree.
  \item \textbf{Frozen decisions.} The number of segments and the segment names live in files under version control, not in someone's memory.
  \item \textbf{Rendered from source.} Every figure in this deck was generated from the stored model objects at compile time. Nothing was pasted.
\end{itemize}

\vspace{0.4em}
\begin{center}
\colorbox{vteal!15}{\parbox{0.9\textwidth}{\centering
`r params$n_assigned` of `r params$n_total` respondents received a segment, including `r params$n_assigned - params$n_analysis` who skipped items but answered enough to be scored.}}
\end{center}

\note{This separates a research prototype from something the shop can depend on. If someone asks about reproducibility, this is the answer.}

## Naming the segments is the real bottleneck {.shrink}

<!-- [CORE] 22 -->
\kicker{from measurement to language}

\small

The model returns `r K_star` segments as `r nrow(dplyr::filter(ci_tbl, kind == "rho"))` conditional response probabilities. Someone still has to read that grid and say what each segment *is*.

\vspace{0.5em}

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\textbf{Done by hand, this is}
\begin{itemize}\itemsep2pt
  \item Slow, and it is the step that delays every deliverable
  \item Inconsistent across analysts and across waves
  \item Undocumented: the reasoning lives in someone's head
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{But it cannot simply be automated}
\begin{itemize}\itemsep2pt
  \item A name is an interpretive claim about a population
  \item A fluent wrong label is worse than no label, because it propagates into every downstream table
\end{itemize}
\end{column}
\end{columns}

\vspace{0.7em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.93\textwidth}{\centering
So the model drafts and the analyst decides. The design question is how to constrain the draft so that verifying it is faster than writing it.}}
\end{center}

\note{Frame this as a measurement problem, not a productivity hack. The payoff sentence is the last one: the goal is not to save the analyst from thinking, it is to make the draft cheap to check.}

## Anchor and adjust: the same pattern as my capstone {.shrink}

<!-- [CORE] 23 -->
\kicker{architecture}

\small

In my JPSM capstone (Latent-to-Language) the pattern was: fit a measurement model, translate its **conclusions** into natural language, then ask the model to adjust from that anchor rather than reason from scratch. The same architecture applies here.

\vspace{0.6em}

\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  every node/.style={font=\scriptsize},
  g/.style={draw=gray!50, fill=gray!8, rounded corners=2pt, minimum width=23mm, minimum height=10mm, align=center},
  b/.style={draw=vteal, fill=vteal!15, rounded corners=2pt, minimum width=23mm, minimum height=10mm, align=center},
  a/.style={-{Latex[length=1.8mm]}, gray!65, thick}
]
\node[g] (m) at (0,0)    {Fitted LCA\\measurement model};
\node[b] (t) at (3.1,0)  {Translate\\to language};
\node[b] (l) at (6.2,0)  {LLM: anchor\\and adjust};
\node[g] (d) at (9.3,0)  {Draft label\\and description};
\node[g] (f) at (12.4,0) {Analyst verifies,\\then freezes};
\draw[a] (m)--(t); \draw[a] (t)--(l); \draw[a] (l)--(d); \draw[a] (d)--(f);
\end{tikzpicture}}
\end{center}

\vspace{0.5em}

\footnotesize
\textbf{Translation, not raw parameters.} The model never sees a probability matrix. It sees each item's actual question wording and each probability against the answer text the respondent saw:

\vspace{0.2em}
\texttt{\scriptsize police "Confianza en la polic\'ia nacional"} \\
\texttt{\scriptsize P(Nada)=0.08, P(2)=0.17, P(3)=0.29, P(4)=0.26, P(5)=0.14, P(6)=0.04, P(Mucho)=0.02}

\note{This is the slide that connects the work to the capstone and shows a method, not a prompt hack. The principle worth stating aloud: feeding raw parameters does not help a language model, but the model's conclusions written as language do.}

## The instrument, verbatim {.shrink}

<!-- [CORE] 24 -->
\kicker{persona and rules}

\scriptsize

\colorbox{vpurple!8}{\parbox{0.96\textwidth}{
\textbf{SYSTEM PROMPT (the persona)}

\vspace{0.3em}
You are a senior survey methodologist who reads latent class analysis (LCA) measurement models. In this work each latent class is called a SEGMENT; that is a word-choice preference and the statistical object is unchanged. Each segment is described only by its item-response probabilities: for every survey item, the probability that a member of that segment gives each answer. A segment leans toward the answers with high probability. You interpret a segment strictly from these probabilities and the item wording, never from outside assumptions.}}

\vspace{0.5em}

\colorbox{vteal!12}{\parbox{0.96\textwidth}{
\textbf{RULES} (sent with every call)

\vspace{0.3em}
1. Use only the response probabilities and item wording shown. The survey context only clarifies what the items refer to; attribute nothing to the segment that the probabilities do not show.

2. Anchor every statement to the high-probability answers of this segment.

3. If the profile is diffuse (no clear high-probability answers), say so.

4. Return only valid JSON: no prose before or after, no markdown fences.}}

\vspace{0.5em}

\footnotesize
Rule 2 is the anchor. Rule 3 is calibrated hedging: the model is required to admit when a segment has no clear shape rather than invent one. Rule 1 fences the survey context to resolving what an institution *is*, so no outside knowledge can enter the interpretation.

\note{Read rule 1 aloud. It is the one that prevents the model from importing what it thinks it knows about Mexican politics into a segment description. Rule 3 is the one reviewers like: the instrument has a defined way to say I do not know.}

## What keeps it honest {.shrink}

<!-- [OPTIONAL] 25 -->
\kicker{guardrails}

\footnotesize

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\textbf{Reproducible by construction}
\begin{itemize}\itemsep2pt
  \item Decoding is deterministic: temperature zero, fixed seed
  \item One isolated call per segment. Joint prompts empirically confused near-neighbour segments
  \item Output is JSON, validated on arrival, never free prose
  \item Labels are frozen to a CSV under version control. That file, not a seed, is the reproducibility mechanism
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Audited, not trusted}
\begin{itemize}\itemsep2pt
  \item A second bounded call harmonizes labels only when two collide, so most runs make no extra call at all
  \item The original draft is retained alongside any edit, so every change is visible in the frozen file
  \item The whole instrument is certified against planted test cases: near-neighbour segments, a deliberately diffuse segment, and a fictional survey context to test for context leakage
  \item Any edit to the prompt obliges a re-certification run
\end{itemize}
\end{column}
\end{columns}

\vspace{0.6em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.92\textwidth}{\centering
The prompt is treated as a measurement instrument: versioned, tested, and re-validated when changed.}}
\end{center}

\note{The certification harness is the part that distinguishes this from prompt engineering. We plant a segment whose meaning we already know and check the instrument recovers it, including checking that it hedges on the diffuse one and ignores the fictional context.}

## Where the line sits {.shrink}

<!-- [CORE] 26 -->
\kicker{a deliberate boundary}

\small

\begin{columns}[T]
\begin{column}{0.47\textwidth}
\colorbox{vgreen!18}{\parbox{0.95\textwidth}{\small
\textbf{The model drafts}
\begin{itemize}\itemsep1pt
  \item A label and description per segment
  \item From the estimated probabilities only
  \item Into a file the analyst edits and owns
\end{itemize}}}
\end{column}
\begin{column}{0.47\textwidth}
\colorbox{vpurple!12}{\parbox{0.95\textwidth}{\small
\textbf{The model never decides}
\begin{itemize}\itemsep1pt
  \item How many segments there are
  \item Which items to keep or drop
  \item What the results mean
  \item Anything that appears as a finding
\end{itemize}}}
\end{column}
\end{columns}

\vspace{0.8em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.93\textwidth}{\centering
The rule I would apply to any AI use in our shop: \\
\textbf{a model may draft configuration, never a decision and never a claim about results.}}}
\end{center}

\vspace{0.4em}

\footnotesize
\begin{center}
Every substantive statement in this briefing rests on the estimated probabilities and their intervals, not on anything a language model wrote.
\end{center}

\note{This is the slide to slow down on. It generalizes past this project into a usable policy, which is what the room can take away and apply to their own work.}

## What I would bring to our work {.shrink}

<!-- [CORE] 27 -->
\kicker{next steps}

\small

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\textbf{Ready now}
\begin{itemize}\itemsep3pt
  \item Segmentation of any categorical battery with population estimates and design-based intervals
  \item Design-based inference review on analyses we already publish
  \item Reproducible pipelines that render from source
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Where I would like to go next}
\begin{itemize}\itemsep3pt
  \item Package the workflow so a new study is a configuration file, not a copied folder
  \item Extend the profiling correction to trends across waves
  \item Evaluate language model assistance where it is auditable
\end{itemize}
\end{column}
\end{columns}

\vspace{0.8em}

\begin{center}
\colorbox{vteal!15}{\parbox{0.92\textwidth}{\centering
\textbf{The offer:} bring me a battery of categorical items and its design variables, \\ and I will tell you how many groups are really in there, how big they are in the population, and how sure we are.}}
\end{center}

\note{Close on the offer, not the method. Leave time for questions; this is where the room decides whether to send you work.}

## Questions {.shrink}

<!-- [CORE] 28 -->
\vspace{1.2em}

\begin{center}
\Large\color{vpurple}\textbf{Thank you}
\end{center}

\vspace{0.8em}

\small
\begin{center}
Methods write-up, code, and the full set of diagnostics available on request.

\vspace{0.5em}
Backup slides follow: full response profiles, item screening, references.
\end{center}

\note{Timing target: reach this slide at minute twelve of fifteen, or minute twenty five of thirty for the long version.}

## Backup: full response profiles {.shrink}

<!-- [BACKUP] 29 -->
```{r}
#| label: fig-heat
#| fig-width: 8.5
#| fig-height: 4.2
#| out-width: 100%
ci_tbl |>
  dplyr::filter(kind == "rho") |>
  dplyr::mutate(Segment = lab_of(class)) |>
  ggplot(aes(factor(category), item, fill = estimate)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", estimate)), size = 1.4) +
  scale_fill_viridis_c(name = "P", limits = c(0, 1)) +
  facet_wrap(~ Segment, ncol = min(3, K_star)) +
  labs(x = "Response category", y = NULL) +
  theme_slide() +
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5))
```

\note{Only if asked. This is the complete measurement model behind the summary plot.}

## Backup: which items carry the segmentation {.shrink}

<!-- [BACKUP] 30 -->
```{r}
#| label: fig-disc
#| fig-width: 8
#| fig-height: 3.2
#| out-width: 96%
# Computed from ci_tbl, which carries item names as a column: fit_w$rho is
# positional and unnamed, so name-based lookup there returns NULL.
pairs_k <- utils::combn(K_star, 2, simplify = FALSE)
ci_tbl |>
  dplyr::filter(kind == "rho") |>
  dplyr::select(item, category, class, estimate) |>
  dplyr::group_by(item) |>
  dplyr::group_modify(~ {
    m <- .x |>
      tidyr::pivot_wider(names_from = class, values_from = estimate) |>
      dplyr::arrange(category) |>
      dplyr::select(-category) |>
      as.matrix()
    tibble::tibble(discrimination = mean(purrr::map_dbl(
      pairs_k, function(p) 0.5 * sum(abs(m[, p[1]] - m[, p[2]])))))
  }) |>
  dplyr::ungroup() |>
  dplyr::arrange(discrimination) |>
  dplyr::mutate(item = factor(item, levels = item)) |>
  ggplot(aes(discrimination, item)) +
  geom_col(fill = viridisLite::viridis(1, begin = 0.45), width = 0.7) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "How strongly the item separates segments", y = NULL) +
  theme_slide()
```

\footnotesize
Every item earns its place. The two lowest, the president and the armed forces, are the institutions on which otherwise different segments agree, which is itself the finding.

\note{If item names do not print, the rho list is unnamed in this build; the modeling document has the labelled figure.}

## Backup: references {.shrink}

<!-- [BACKUP] 31 -->
\kicker{sources cited}

\scriptsize

**Method**

Goodman, L. (1974). Exploratory latent structure analysis. *Biometrika* 61(2).

Patterson, B., Dayton, C. M., and Graubard, B. (2002). Latent class analysis of complex sample survey data. *JASA* 97(459), 721 to 728.

Bolck, A., Croon, M., and Hagenaars, J. (2004). Estimating latent structure models with categorical variables. *Political Analysis* 12(1), 3 to 27.

Vermunt, J. (2010). Latent class modeling with covariates: two improved three-step approaches. *Political Analysis* 18(4), 450 to 469.

Bakk, Z., Oberski, D., and Vermunt, J. (2014). Relating latent class assignments to external variables. *Political Analysis* 22(4), 520 to 540.

Lumley, T., and Scott, A. (2015). AIC and BIC for modeling with complex survey data. *JSSAM* 3(1), 1 to 18.

Linzer, D., and Lewis, J. (2011). poLCA: an R package for polytomous variable latent class analysis. *JSS* 42(10).

**Applications in public opinion**

Breen, R. (2000). Why is support for extreme parties underestimated by surveys? *BJPS* 30(2), 375 to 382.

Blaydes, L., and Linzer, D. (2008). The political economy of women's support for fundamentalist Islam. *World Politics* 60(4), 576 to 609.

Oser, J., Hooghe, M., and Marien, S. (2013). Is online participation distinct from offline participation? *PRQ* 66(1), 91 to 101.

Bonikowski, B., and DiMaggio, P. (2016). Varieties of American popular nationalism. *ASR* 81(5).

Alvarez, R. M., Levin, I., and N\'u\~nez, L. (2017). The four faces of political participation in Argentina. *Journal of Politics* 79(4), 1386 to 1402.

\note{Leave up during questions if the room is academic.}
